### 2021-03-30 old1.tex
init

%for a more compact document, add the option openany to avoid
%starting all chapters on odd numbered pages
\documentclass[12pt]{cmuthesis}

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        comar    = Cyrus Omar
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

% My main contribution is putting everything into a single class files and small
% template since I prefer this to some complicated sprawling directory tree with
% makefiles.

% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subcaption}

\usepackage[capitalize,noabbrev,nameinlink]{cleveref}
\usepackage{float}
\usepackage[scaled]{inconsolata}

\captionsetup{labelfont=bf}
% \captionsetup{font=small}
\captionsetup{labelfont=bf}
\captionsetup{textfont={}}
%\captionsetup[subfloat]{font=small}
%\captionsetup[subfloat]{farskip=5pt}
%\captionsetup[subfloat]{captionskip=1pt}

\DeclareUnicodeCharacter{0097}{~}

% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}

% Provides a draft mark at the top of the document. 
\draftstamp{\today}{DRAFT}

\begin {document} 
\frontmatter

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{empty}

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Filter Representation in Vectorized Query Execution}}
\author{Amadou Latyr Ngom}
\date{June}
\Year{2020}
\trnumber{}

\committee{
Andy Pavlo, Chair \\
Todd C. Mowry \\
}

\support{}
\disclaimer{}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{Vectorization, Compilation, Filter, SIMD}

\maketitle

\begin{dedication}
\end{dedication}

\pagestyle{plain} % for toc, was empty

%% Obviously, it's probably a good idea to break the various sections of your thesis
%% into different files and input them into this file...

\begin{abstract}
In-memory Database Management Systems (DBMSs) use vectorization to reduce interpretation overhead, improve cache locality, and perform loop optimizations. For each \textit{vector} (i.e. batch of tuples), the DBMS needs to track the set of valid tuples after a predicate application. Systems employ one of two representations: Selection Vectors (SelVecs) or Bitmaps. In this work, we analyze each approach's strengths and weaknesses to provide recommendations on how to implement vectorized operations. We determine, through a wide range of micro-benchmarks, that the optimal implementation strategy is a function of many factors: the cost of the iterating through tuples, the cost of the operation itself and how amenable it is to SIMD vectorization. Our analysis shows that Bitmaps perform better for operations that can be vectorized using SIMD instructions but that SelVecs otherwise perform better due to a cheaper iteration logic.
\end{abstract}

\begin{acknowledgments}
My advisor is cool.
\end{acknowledgments}

\tableofcontents
\listoffigures
\listoftables

\mainmatter

%% Double space document for easy review:
%\renewcommand{\baselinestretch}{1.66}\normalsize

% The other requirements Catherine has:
%
%  - avoid large margins.  She wants the thesis to use fewer pages, 
%    especially if it requires colour printing.
%
%  - The thesis should be formatted for double-sided printing.  This
%    means that all chapters, acknowledgements, table of contents, etc.
%    should start on odd numbered (right facing) pages.
%
%  - You need to use the department standard tech report title page.  I
%    have tried to ensure that the title page here conforms to this
%    standard.
%
%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
%
% Other than that, just make it look good...

\chapter{Introduction and Background}
DBMSs typically handle two broad classes of applications: Online Transaction Processing (OLTP) and Online Analytical Processing (OLAP) applications \cite{abadi_phd}. OLTP applications tend to be short-running, write-heavy queries that only touch a few entries in a database. OLAP applications, which are the focus of this thesis, are long-running, read-mostly queries that analyze large amounts of data. Historically, this data was disk-resident; disk accesses were the bottleneck of OLAP queries. As DRAM sizes increase, and the cost per GB decreases, large databases begin to fit in main memory. Rather than optimize for disk access, modern DBMSs must optimize for CPU efficiency by reducing instruction count and cycles per instruction when executing queries.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/SampleQuery.png}
    \caption{\textbf{Sample SQL Query and Plan.}}
    \label{fig:intro_query}
\end{figure}

The traditional execution technique used in most DBMSs \cite{mysql, posrgres}, the Iterator Model \cite{volcano}, became ill-suited for modern OLAP applications. It executes a query by interpreting the query plan. Each database operator implements a virtual \texttt{next()} function that returns the next tuple it can generate; each expression (i.e. addition, multiplication) also provides an virtual \texttt{evaluate()} function that takes in a tuple and returns a value. \cref{fig:volcano_graph} showcases the Iterator model with the sample query in \cref{fig:intro_query}. The operators \texttt{Scan A} and \texttt{Scan B} iterate through the table and emit the tuples to the parent operator. The filter only returns tuples that satisfy the \texttt{WHERE} clause. The join first builds a hash table, then probes it to find matches.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Volcano.png}
    \caption{\textbf{The Iterator Model on the Sample SQL Query.}}
    \label{fig:volcano_graph}
\end{figure}

We can see why the Volcano model is ill-suited for in-memory DBMSs. For every tuple -- of which there can be billions in an OLAP setting -- the virtual function calls (marked in red in the figure) constitute a significant overhead.

To minimize this overhead for in-memory DBMSs, researchers have developed the Vectorization Model.

\iffalse
\section{Compilation}
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/Pipeline.png}
 \caption{Pipelines within the sample query.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/PipelineCode.png}
 \caption{Code generated for the sample query.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{The compilation model with the sample query. Each pipeline and its corresponding code have the same color.}
\label{fig:update_map_intro}
\end{figure}

Query compilation \cite{system_r, hique, hyper_llvm} aims to eliminate the cost of interpretation by compiling the query plan down to machine code.
{
\color{red}
Describe what compilation means.
}

The data-centric compilation technique pioneered by Hyper \cite{hyper_llvm} splits the plan into several pipelines and compiles them individually. A pipeline is a sequence of relational operators that do not materialize the input result set followed by a pipeline-breaker -- an operator that does materialize the input set (e.g., to build a hash table, or to sort tuples). \cref{fig:pipeline_graph} showcases the pipelines of the sample query. The first pipeline-breaker occurs in the join operator because it needs all input tuples to build a hash table. \cref{fig:pipeline_code} lists the code generated for the query. Conceptually, it performs the same task as \cref{fig:volcano_graph} without any of the virtual function calls. In addition to eliminating interpretation, this approach maximizes register/cache efficiency because a pipeline fully processes a single tuple before fetching the next one. Compilation thus reduces both the number of instructions executed and the number cycles per instructions.
\fi

\section{Vectorization}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Vectorization.png}
    \caption{\textbf{The Vectorization Model with the Sample Query} -- The code is nearly identical to that in \cref{fig:volcano_graph}}
    \label{fig:vectorization_graph}
\end{figure}

The vectorization model, pioneered by Vectorwise \cite{vectorwise}, is almost identical to the Volcano model, except that it deals with batches of tuples or \textit{vectors} (e.g., 2048 tuples at a time). The \texttt{next()} function returns a vector of tuples, and the \texttt{evaluate()} function returns a vector of values. \cref{fig:vectorization_graph} showcases vectorization model on the sample query in \cref{fig:intro_query}; it is near identical to \cref{fig:volcano_graph}.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/VectorMul.png}
    \caption{\textbf{A Vectorized Multiplication Function} -- The code in green indicates compiler or programmer optimizations to exploit Intel's AVX512 features.}
    \label{fig:vector_mul}
\end{figure}
Vectorization provides two main benefits. First, it amortizes the cost of interpretation by the size of vectors. Second, most functions that operate on vectors are highly amenable to loop optimizations (e.g., loop unrolling, SIMD vectorization) by the compiler or by the programmer. Consider, for a example, the function shown in \cref{fig:vector_mul}. The iterations of the loop are independent, and the floating-point multiplication possesses a SIMD variant, which means the loop can be unrolled and vectorized by the compiler.

\iffalse
\begin{figure}[H]
    \centering
    \includegraphics[scale=0.5]{images/VectorCompilation.png}
    \caption{A hybrid vectorization and compilation model with the sample query..}
    \label{fig:mix_compilation_vectorization}
\end{figure}
Our system \cite{rof, noise_page} combines both data-centric code generation and vectorization : it generates code that operates on vectors. This approach eliminates interpretation overhead and still enables loop optimizations. \cref{fig:mix_compilation_vectorization} showcases the mixed model. It differs from \cref{fig:compilation_graph} only in that it operates on vectors.
\fi

Vectorization introduces a new challenge. Individual tuples within a vector may be \textit{filtered out}, meaning that should not be processed. Only \textit{selected} tuples -- those not filtered out -- should be processed. For example, in \cref{fig:vectorization_graph}, only tuples that satisfy the \texttt{WHERE} clause should be inserted in the hash table. Next, we discuss the methods used to represent filters, i.e., to identify selected tuples within a vector.

\section{Filter Representation}
\cref{fig:repr_intro} displays the two common ways to represent filters: Selection Vectors (SelVecs) and Bitmaps. SelVecs index into the selected tuples; tuples without an index are implicitly filtered out. For Bitmaps, on the other hand, the set bits indicate the selected tuples; the unset bits indicated filtered out ones.
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.7]{images/FilterRepresentationIntro.png}
    \caption{\textbf{Filter Representation} -- Selection Vectors (left) index into the tuple vector. Bitmaps (right) are only set on selected rows.}
    \label{fig:repr_intro}
\end{figure}
Previous work has assumed one of these two representations, without providing a reason for that choice. Vectorwise \cite{vectorwise}, and work derived from it \cite{miro_adapt, everything_vectorized, sompolski_vec},  rely on SelVecs, whereas the more recent VIP \cite{orestis_bitmap} relies on Bitmaps. This paper aims to analyze the trade-offs between these two representations and derive insights for developers implementing vectorized query processing.

\chapter{Computing on Filtered Vectors}
\section{Operations on Filtered Vectors}
\begin{figure}[t!]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/UpdateIntro.png}
 \caption{Update operation.}
  \label{fig:update_intro}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MapIntro.png}
 \caption{Map operation.}
  \label{fig:map_intro}
\end{subfigure}
\caption{\textbf{Operations on Filtered Vectors} -- Updates change the set of selected tuple. Maps apply a function to the set of selected tuples without updating the set.}
\label{fig:update_map_intro}
\end{figure}
\cref{fig:update_map_intro} shows the two kinds of operations on vectors of tuples: Updates and Maps. Updates apply a new predicate function to the vector and update the selected tuples accordingly. A scan filter (i.e., WHERE clause) is an example of an update operation. Maps, on the other hand, keep the selected set but compute a new vector using a mapping function, or reduce the vector to a single value (e.g., for aggregation). A projection (i.e., SELECT clause) is an example of a map operation. All relational operations considered in this paper are decomposable into a set of maps and updates. \cref{fig:probe_example} shows the example of join probes. First, we compute the hashes of the input tuples; the result is a vector of hash values (Map). Then, we perform a hash table lookup using the hashes to obtain a vector of hash table entries (Map). Due to hashing collisions, the entries may not match the input, so we need to compare the join keys to filter out inexact matches (Update). We then advance the hash table chains (Map) and repeat the comparison until entries are exhausted.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/JoinProbeExample.png}
    \caption{\textbf{Multi-step Join Probe} -- Bold squares indicate the hash table entries currently pointed to.}
    \label{fig:probe_example}
\end{figure}

Other relational operators are similarly decomposable. The next section briefly discusses transitioning between steps before we move on to the core of the thesis: optimizing each step.

\section{Transitioning Between Steps}
In this section, we briefly talk about the cost of transitioning between steps. Most of the time, this cost is non-existent because the filter: the transition is \textit{implicit}. The hashing phase and the lookup phase in \cref{fig:probe_example} have the same selection vector. On the other hand, the Chain Following phase operates on entries that do not satisfy the comparison: the transition involves a set difference. Transitions in decomposed disjunctive filters \cite{pcq} (shown in \cref{fig:disjunctive_filter_howto}) involve both a set difference for short-circuiting and a union to get the final result.
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DisjunctionHowto.png}
    \caption{\textbf{Transitioning Between Steps} -- We apply the first clause on all input tuples. In the second step, however, we only apply the second clause on those tuples that don't pass the first clause. In the third step, we take the union between the two outputs.}
    \label{fig:disjunctive_filter_howto}
\end{figure}
To the best of our knowledge, all transitions take the form of set differences (for elements that do not pass the previous predicate) and unions (to obtain the final result). These operations are cheap to implement with Bitmaps because we can easily rely on SIMD instructions. Efficiently implementing set operations on SelVecs is more complicated.  Whenever necessary, Updates can write to one additional SelVec for tuples that do not pass the filter; this second SelVec implicitly corresponds to the result of a set difference. Note that it does not overlap with the first SelVec, so a union of any of their subsets is just an array append, a fast operation. \cref{fig:disjunctive_filter_perf} shows that the overhead of set operations is negligible for both bitmaps and selection vectors. The rest of the paper, therefore, focuses on optimizing the individual steps.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{eval/transition.pdf}
    \caption{\textbf{Transition Overheads} -- The overheads are $3.55\%$ and $4.06\%$ for Bitmaps and SelVecs respectively.}
    \label{fig:disjunctive_filter_perf}
\end{figure}

\section{Compute Strategies}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/ComputeIntro.png}
    \caption{\textbf{Compute Strategies} -- Selective Compute (left) only considers selected tuples. An \texttt{x} indicates an uninitialized value. Full Compute (right) considers all tuples.}
    \label{fig:compute_intro}
\end{figure}
 A compute strategy determines how to perform the Update or Map operations. We consider three compute strategies: Selective, Full, and Mixed Compute. As their names indicate, Selective Compute only applies the given functions to selected tuples, whereas Full Compute applies them to all tuples. \cref{fig:compute_intro} shows each approach's effect. While Full Compute usually performs more work, it greatly benefits from SIMD vectorization, simple loop structure (for loop unrolling and interleaving), and easy branch prediction. To exploit this trade-off, the third strategy, namely Mixed Strategy, switches from Full Compute to Selective Compute when the selectivity (i.e., the ratio of selected tuples) goes below a certain threshold. The next sections will show how to obtain this threshold.

Note that for Update operations, Full Compute is not nicely compatible with SelVecs because a slow intersection of sorted sets \cite{sorted_set} would be required to compute the final selection. This limitation will become important when choosing a strategy to implement. Furthermore, operations with side-effects (e.g., hash table build, aggregation) cannot use Full Compute for correctness reason. This thesis focuses on side-effect free operations.

Furthermore, for SelVecs, there is another dimension we did not consider: Branching Store versus Non-Branching Store. We do not focus on this dimension, for it has been studied extensively before \cite{miro_adapt}. All experiments below use the optimal branching strategy.

\section{The Filter Decision Tree}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DecisionTree.png}
    \caption{\textbf{The Filter Decision Tree.}}
    \label{fig:filter_decision_tree}
\end{figure}
\cref{fig:filter_decision_tree} shows the filter decision tree. For every kind of operation on filtered tuples, the tree indicates what the best representation and compute strategy pair is. It assumes that the developer implements all representations and compute strategies, which is a fairly costly endeavor engineering-wise. When this is not affordable, systems designers can still use this tree as a starting point before making compromises. The remaining part of this section will give a high-level rationale behind the tree. Chapter 4 will provide practical case studies that justify the shape of this tree. Chapter 5 will use the insights provided by this tree to execute an end-to-end multi-step query and show the performance gains.

\section{High Level Rationale}
\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleFull.png}
 \caption{Full Compute.}
  \label{fig:rationale_full}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%  
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleBitmap.png}
 \caption{Selective Compute with a Bitmap. The iteration logic is adapted from \cite{bitmap_iteration}. The SIMD version is equivalent to Full Compute.}
  \label{fig:rationale_bitmap}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%  
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleSelVec.png}
 \caption{Selective Compute with a SelVec. The last \texttt{scatter} is only necessary to maintain consistent indices between the input and output vector during multi-step Map operations. It should otherwise be avoided due to its slowness.}
  \label{fig:rationale_selvec}
\end{subfigure}
\hspace*{\fill}%
\vspace*{8pt}%

\caption{\textbf{Implementing Vector Primitives} -- The code in green indicates the core operation (multiplication). The code in red indicates iteration logic.}
\label{fig:rationale_code}
\end{figure}

Every operation on filtered tuples is a loop, the body of which performs the actual computation. Three factors influence this loop's performance: the iteration logic, number of processed tuples, and time per processed tuple. \cref{fig:rationale_code} shows the three ways to multiply a filtered vector by a constant (a Map operation). Full Compute (e.g., \cref{fig:rationale_full}), regardless of representation, has the cheapest iteration logic because it does not need to differentiate selected indices. For Selective Compute, iterating over a Bitmap (e.g., \cref{fig:rationale_bitmap}) is more expansive than iterating over a SelVec (e.g., \cref{fig:rationale_selvec}) because of the bit operations required. On the other hand, Full Compute processes more tuples than Selective Compute, especially when the selectivity is low. Finally, SIMD instructions usually reduce the number of iterations and the number of cycles per processed tuples. SIMD code is cheapest with Full Compute; Selective Compute requires \texttt{gather} and possibly \texttt{scatter} to differentiate selected tuples.

We can use these these three factors to explain the shape of the decision tree.

\textbf{The root node}: The root of the tree addresses the case of operations on variable-length data, as is often the case for string operations (e.g., string comparison, sub-string). The number of cycles dedicated to each tuple usually far dominates the total cost. In this scenario, Full Compute is highly inefficient because it processes more tuples than Selective Compute. Also, the cost iteration logic is negligible, making representation irrelevant. We, therefore, choose to use Selective Compute, with either SelVecs or Bitmaps.

\textbf{The branching node}: We then move on to the Branching node, which addresses operations with branches in them (e.g., logical operations, \texttt{CASE} statements, \texttt{IN} statements). Branching increases the cost of SIMD relative to SISD: SISD skips unnecessary instructions, SIMD executes all of them using bit-masks. Furthermore, branches within a loop complicate loop optimizations (e.g., unrolling, interleaving). As a result, Selective Compute beats Full Compute. SelVecs only win over Bitmaps because of the simplicity of their iteration logic.

\textbf{The non-simdable node}: We now consider the non-simdable node. The most (only?) relevant operations that qualify are integer division and modulo. Complex pointer manipulations, while technically simdable using expensive scatters and gathers, also fall under this category. No compiler will auto-vectorize them, and few programmers will write and maintain such SIMD code. Like with variable-length instructions, Full Compute would introduce too much extra computation, and the representation does not matter.

\textbf{The SNBS nodes}: The most interesting nodes of this tree are those dedicated to scalar non-branching simdable (SNBS) operations; all three factors simultaneously play an essential role. Besides, for Updates, Full Compute is incompatible with SelVecs; it may be necessary to switch both compute strategy and filter representation depending on the task. The next section will justify the claims in these nodes and provide an efficient way to switch compute strategy and representation.

\chapter{Experimental Evaluation}
In this section, we experimentally justify the shape of the tree. In doing so, we will also provide a general technique that systems can use optimally switch between strategies. \cref{tab:strategies} summarizes all the strategies we implemented. We run all experiments on an Intel Xeon Platinum 8124M CPU @ 3.00GHz with AVX512 support.

\begin{table}[t!]
\centering
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Strategy Name} & \textbf{Representation} & \textbf{Compute} & \textbf{SIMD}      & \textbf{Compatibility} \\ \hline
SelVecPartial         & Selection Vectors       & Selective        & None               & Update and Map         \\ \hline
SelVecManual          & Selection Vectors       & Selective        & Manually Written   & Update and Map         \\ \hline
BitmapPartial         & Bitmaps                 & Selective        & None               & Update and Map         \\ \hline
BitmapFull            & Bitmaps                 & Full             & Auto Vectorization & Update and Map         \\ \hline
BitmapFullManual      & Bitmaps                 & Full             & Manually Written   & Update and Map         \\ \hline
Full                  & Either                  & Full             & Auto Vectorization & Map                    \\ \hline
FullManual            & Either                  & Full             & Manully Written    & Map                    \\ \hline
\end{tabular}
\caption{\textbf{Implemented Strategies} -- For the SIMD columns, `Auto Vectorization' indicates that we rely on compiler auto vectorization, whereas `Manually written' indicates that we write SIMD code ourselves using intrinsics. On the compatibility column, Update operations cannot use both Full Compute and Selection Vectors, which explains the last two entries.}
\label{tab:strategies}
\end{table}

\section{Variable Length Data}
\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_string_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE country < `Japan'}}
  \label{fig:varlen_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_string_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT POSITION(`a' in country)}.}
  \label{fig:varlen_map}
\end{subfigure}
\caption{\textbf{Variable Length Data} -- Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
\label{fig:varlen_map_update}
\end{figure}

In this section, we analyze the performance profile of operations on variable-length data. We choose cheap Update and Map operations -- string comparison, and character location on short strings (country names) -- to show that time spent per tuple typically dominates the cost of string operations. Here, the primitives process one tuple at a time; it is impossible to use SIMD registers on a vector with variable-length data.

The profiles are shown in \cref{fig:varlen_map_update}. The results confirm the high-level intuition described in the previous section. The string operations themselves are so expensive that representation matters little, and full compute is not viable. As a result, we decide to use that Selective Compute with either representation.

\section{Branching Operations}
\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_and_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \&\& col2 < val2}}.
  \label{fig:logical_and_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_and_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \&\& col2 < val2}.}
  \label{fig:logical_and_map}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_or_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 || col2 < val2}}.
  \label{fig:logical_or_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_or_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 || col2 < val2}.}
  \label{fig:logical_or_map}
\end{subfigure}

\caption{\textbf{Branching Operations} -- Each operation contains one branch because of boolean short-circuiting.}
\label{fig:logical_map_update}
\end{figure}

\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_and_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \& col2 < val2}}.
  \label{fig:bitwise_and_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_and_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \& col2 < val2}.}
  \label{fig:bitwise_and_map}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_or_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 | col2 < val2}}.
  \label{fig:bitwise_or_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_or_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 | col2 < val2}.}
  \label{fig:bitwise_or_map}
\end{subfigure}

\caption{\textbf{Bitwise Operations} -- The increase in performance is due to the lack of branching.}
\label{fig:bitwise_map_update}
\end{figure}

We now consider operations involving branching execution paths. We choose the logical or/and operations because boolean short-circuiting involves a single branch to avoid evaluating unnecessary conditions. To isolate the effect of branching, we also evaluate branch-less bitwise or/and operations. We have confirmed that the compiler manages to auto-vectorize the operations.

The results for logical operations are shown in \cref{fig:logical_map_update}. We see that PartialSelVec performs better thanks to a cheaper iteration logic and that Full Compute is not viable because SIMD does not interact well with branching. \cref{fig:bitwise_map_update} confirms that branching is truly responsible for the difference in performance. The only difference between bitwise and logical operations is the short-circuiting that takes place. While Full Compute can sometimes be competitive in these scenarios (see \cref{fig:logical_or_map} at high selectivities), we decide, for the sake of simplicity, to use Selective Compute with SelVecs.

\section{Non Simdable Operations: Integer Division}
\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_division_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 \% col2 < val}}.
  \label{fig:division_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_division_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 / col2}.}
  \label{fig:division_map}
\end{subfigure}
\caption{Simple operations on a variable-length column. Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
\label{fig:division_map_update}
\end{figure}

For this section, we consider operations involving integer division. There is no SIMD instruction for integer division; the primitives process one tuple at a time.

The results are shown in \cref{fig:division_map_update}. As predicted in the high-level rationale, the cost of integer division is so high that we recommend Selective Compute with either representation.
\section{SNBS Operations}
\label{snbs_section}
\begin{figure}[t!]
    \centering
    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_update.pdf}}
    \caption{\textbf{SNBS Update Operations} -- For Updates, SelVecManual relies on the \texttt{compress\_store} SIMD instruction rather than on \texttt{scatter}, which reduces its overhead.}
    \label{fig:snbs_update}
\end{figure}

\begin{figure}[t!]
    \centering
    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_map.pdf}}
    \caption{\textbf{SNBS Map Operations} -- For Maps, SelVecManual relies on the \texttt{scatter} SIMD instruction, which explains why it initially does not perform better than SelVecPartial.}
    \label{fig:snbs_map}
\end{figure}

As mentioned in the rationale, the performance of SNBS operations will depend on all three performance factors: iteration logic, number of tuples processed, time per tuple. To isolate each component's effect, we experimentally evaluate the running time of Maps and Updates as a function of the selectivity and the number of single-cycle instructions (e.g., addition, subtraction, bitwise and, or, xor) in each operation.

The experimental profile of operations with multi-cycle instructions (e.g., multiplication, load, store) can be approximated by treating such operations as multiple single-cycle. The multiplying factor can be found in various instruction tables \cite{intel2019}. For example, in the machine we use for testing, a Map operation with a multiplication should have a profile similar to that of a Map operation with three additions. This method is likely not optimal, though; finding a better one is left as future work.

\cref{fig:snbs_update} and \cref{fig:snbs_map} show the performance profile of our different implementations as the number of single-cycle operations varies. The profiles correspond to what we expect based on the three performance factors, but a detailed explanation of the main features in order.
\begin{itemize}
    \item SelVecPartial generally performs better than BitmapPartial. We attribute this fact to the cost of the iteration logic. This difference is particularly relevant for cheap operations, but become irrelevant as complexity increases.
    \item Full Compute is always worst when the selectivity is low; it processes far more tuples than Selective Compute. 
    \item Full Compute is always best when the selectivity is high thanks to cheaper iteration logic and cheaper SIMD code. For example, SelVecManual always needs a gather instruction, whereas BitmapFullManual does not.
    \item SIMD beats SISD because of less time spent per tuple, as exemplified by the difference between SelVecPartial and SelVecManual.
    \item Manual Vectorization (with FullManual and BitmapFullManual) sometimes performs much better than Auto-Vectorization (with Full and BitmapFull). With further investigation of the generated code, we discovered that the compiler is overly conservative when it comes to using AVX512. AVX512 registers result in decreased CPU frequency \cite{avx512_registers}, so the compiler is careful not always to use them. It often uses AVX2 registers instead. We, on the other hand, always use AVX512.
\end{itemize}

\section{Implementing Mixed Compute}
\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/snbs_thres_update.pdf}
 \caption{Update thresholds between BitmapFullManual and Selective Compute strategies.}.
  \label{fig:snbs_thres_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/snbs_thres_map.pdf}
 \caption{Map thresholds between FullManual and Selective Compute strategies.}
  \label{fig:snbs_thres_map}
\end{subfigure}
\caption{\textbf{Mixed Compute Thresholds} -- The Full Compute strategies are BitmapFullManual and FullManual because compiler auto-vectorization does not consistently use AVX512 registers (as explained in the previous section).}
\label{fig:snbs_thres}
\end{figure}

Both \cref{fig:snbs_update} and \cref{fig:snbs_map} explain the necessity of Mixed Compute. Depending on the operation, there is a selectivity threshold at which we should switch from Full Compute to Selective Compute. As \cref{fig:snbs_thres} shows, this threshold a function of the cost of the operation. The curves for BitmapPartial and SelVecPartial are relatively flat because they lack the SIMD capabilities. For example, Full Compute dedicates fewer cycles per tuple even as it processes around five times more tuples at a selectivity of $0.2$. As for SelVecManual, the \texttt{gather} instruction (show in \cref{fig:rationale_selvec}) dominates its cost when the number of operations is low. As the number of operations increases, this \texttt{gather} plays a less important role in the overall performance; SelVecManual becomes more and more competitive with Full Compute because they both utilize SIMD instructions. The threshold, therefore, continually increases.

For a new query, the optimal thresholds can either be obtained by running micro-benchmarks, or by using the pre-computed thresholds in \cref{fig:snbs_thres}. The optimal strategy is as follows: 
\begin{itemize}
\item For Maps: use Full when the selectivity is above the threshold, and use SelVecManual when the selectivity is below the threshold.
\item For Update: use BitmapFull or BitmapFullManual when the selectivity is above the threshold, and use SelVecManual when the selectivity is below the threshold. There is a small cost associated with the conversion from Bitmap to SelVec, but, in practice, other operations always follow an update (e.g., there is a projection after a scan filter). The subsequent operations amortize the conversion cost.
\end{itemize}

\section{End-To-End Evaluation: TPCH Q6}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Q6.png}
    \caption{\textbf{The TPCH Q6 query.}}
    \label{fig:tpch_q6}
\end{figure}

\begin{figure}[t!]
    \centering
    \includegraphics[width=.9\linewidth]{eval/q6_bench.pdf}
    \caption{\textbf{End-To-End Evaluation: TPCH Q6}. The scale factor is 1 (around $600000$ tuples). The horizontal line indicates the performance of the optimal strategy (BitmapFull + SelVecManual).}
    \label{fig:tpch_q6_perf}
\end{figure}

We now study the benefits of using the decision tree on a full query: TPCH Q6 \cite{tpch} (shown in \cref{fig:tpch_q6}). The query consists of five \texttt{WHERE} clause filters (Updates), and one aggregation (Map). All operations are SNBS, and all transitions are implicit. We implemented three mixed strategies: BitmapFull + BitmapPartial, BitmapFull + SelVecPartial, BitmapFull + SelVecManual. In all these strategies, the second filter triggers the threshold-based switch, which amortizes the conversion cost by a factor of four.

We compare these mixed strategies to the fixed ones of previous systems. Vectorwise and its related systems \cite{vectorwise, miro_adapt, everything_vectorized, sompolski_vec} exclusively rely on SelVecs: they cannot use on Full Compute for the five Update filters. \cite{miro_adapt} does use Full Compute on Map tasks with high selectivity, but the selectivity of Q6's final vector is in the order of $0.01$; Full Compute is too wasteful in this scenario. SelVecPartial is, therefore, representative of these systems. The recent \cite{orestis_bitmap} is more similar to BitmapFull.

The results are shown in \cref{fig:tpch_q6_perf}. As predicted in the previous chapter, BitmapFull + SelVecManual is indeed the optimal strategy.  We can also see that the fixed strategies are all worse than the mixed ones. For example, SelVecPartial is $2\times$ worse that the optimal strategy; BitmapFull is $1.37\times$ worse.

For software engineering reasons, it may be more practical to use the BitmapFull + BitmapPartial strategy. It involves only one representation (i.e., no duplicate functionality), does not require manual vectorization, and is the second-best strategy (worse by $1.2\times$). Software engineers should balance performance and maintainability.

\iffalse
\chapter{End-To-End Evaluation: TPCH Q6}

\chapter{End-To-End Evaluation: The Join Probe}
We now study the benefits of using the decision tree above on a multi-step operator involving both Maps and Updates: the join probe.

\section{Hashing}
\cref{fig:hashing} shows the cost of hashing for both strings and integers. XXHash operates on variable length data. As the root of decision tree indicates, using selective compute is the best strategy. Murmur Hash, on the other hand, is an SNBS operations. As such, Mixed Compute is the best strategy.

\section{Hash Table Lookup}
SIMD Hash table lookup can be implemented as vector gather using the hash table array as the sbase address, and the lower bits of the hash values as the indices. \cref{fig:lookup} shows the performance profile of hash table lookup. The compiler is somehow unable to perform auto-vectorization, so we only consider Manually written SIMD code. Note that the cost of a vector gather is at least 16 times that of the single-cycle instructions in the previous chapter which explains why the threshold is fairly high. Again Mixed Compute is the best strategy.

\section{Key Equality Check}
To perform SIMD equality check, we need to first gather the values that we will compare because they are part of an arbitrary hash table payload (which may contain a pointer to the next entry, other join keys and values). In addition, we need to avoid operating on null entries by using a extra SIMD mask. \cref{fig:key_check} shows that the BitmapFullManual + SelVecManual is best. We can ignore the conversion overhead because the equality check is always followed by more operations.

\section{Chain Following}
The final step we consider is chain following. The SIMD implementation, again involves a gather of the \texttt{next} pointer of hash table entries. The profile is shown in \cref{fig:chain_follow}. Unsurprisingly the Mixed Strategy described above wins again.
\fi

\chapter{Related Works}
There is a large corpus of previous work dedicated to vectorized query execution. They mostly differ from this thesis in that they do not consider the impact of filter representation in the performance of vectorized primitives. We now discuss this work and how they relate to our proposed method.

\cite{sompolski_vec} compares the cost of vectorization to data-centric query compilation. It combines the two approaches to generate new vectorized primitive at runtime, which is the method we use in \labelcref{snbs_section}. Using only the SelVec representation, the authors compare the performance of vectorized primitives under various compute strategies (e.g., Full v Selective Compute, Branching v Non-Branching) but do not provide a way to switch between these strategies. \cite{everything_vectorized} performs a similar analysis for whole queries rather than individual primitives.

\cite{miro_adapt} recognizes the importance of switching compute strategy at run-time. For each operation, the authors implement several \textit{flavors}, i.e., different ways to perform the same operations. They then use reinforcement learning (RL) to switch the best \textit{flavors} dynamically at run-time. This approach is more general than our static micro-benchmarks because it takes into account changes in system load. SelVec is the only representation considered, limiting the range of adaptivity on Update tasks. Besides, our decision tree provides more explainability that RL techniques. As future work, we could employ a similarly dynamic strategy to determine our thresholds.

\cite{orestis_bitmap} builds a query engine from the ground up using manually written SIMDed primitives. Its implementation of SNBS operations is most similar to our BitmapFullManual strategy. It considers a more extensive range of tasks than our side-effect-free Update/Map tasks (e.g., partitioning, building hash tables, sorting). As future work, we should perform our analysis on side-effect-full operations.

\chapter{Conclusion}
This work analyzed the impact of filter representation (i.e., Bitmap v SelVec) and compute strategy (i.e., Full v Selevective Compute) on the performance of vectorized primitives. We identified the factors that influence performance: iteration logic, number of tuples processed, and time spent per tuple. We explained how each combination of representation and compute strategy balances between these three strategies. Full Compute has the cheapest iteration logic, processes all tuples, but spends less time on each tuple when SIMD vectorization is possible. Full Compute is, however, only available with Bitmaps on Update tasks. Selective Compute with SelVecs has a cheaper iteration logic than Selective Compute with Bitmaps, and is more amenable to SIMD vectorization. Using these observations, we created a decision tree to help systems developers find the optimal representation for a given primitive and performed an empirical evaluation to confirm the shape of this decision tree. Finally, we showcased the benefits of our analysis using an end-to-end query involving multiple primitives.

%\appendix
%\include{appendix}

\backmatter

%\renewcommand{\baselinestretch}{1.0}\normalsize

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}}
%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{plainnat}
\bibliography{citations} %your bib file

\end{document}

### 2021-03-30 paper.tex
init

%for a more compact document, add the option openany to avoid
%starting all chapters on odd numbered pages
\documentclass[12pt]{cmuthesis}

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        comar    = Cyrus Omar
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

% My main contribution is putting everything into a single class files and small
% template since I prefer this to some complicated sprawling directory tree with
% makefiles.

% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subcaption}

\usepackage[capitalize,noabbrev,nameinlink]{cleveref}
\usepackage{float}
\usepackage[scaled]{inconsolata}

\captionsetup{labelfont=bf}
% \captionsetup{font=small}
\captionsetup{labelfont=bf}
\captionsetup{textfont={}}
%\captionsetup[subfloat]{font=small}
%\captionsetup[subfloat]{farskip=5pt}
%\captionsetup[subfloat]{captionskip=1pt}

\DeclareUnicodeCharacter{0097}{~}

% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}

% Provides a draft mark at the top of the document. 
%\draftstamp{\today}{DRAFT}

\begin {document} 
\frontmatter

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{empty}

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Filter Representation in Vectorized Query Execution}}
\author{Amadou Latyr Ngom}
\date{July}
\Year{2020}
\trnumber{}

\committee{
Andy Pavlo, Chair \\
Todd C. Mowry \\
}

\support{}
\disclaimer{}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{Vectorization, Compilation, Filter, SIMD}

\maketitle

\begin{dedication}
\end{dedication}

\pagestyle{plain} % for toc, was empty

%% Obviously, it's probably a good idea to break the various sections of your thesis
%% into different files and input them into this file...

\begin{abstract}
Advances in memory capacity have allowed Database Management Systems (DBMSs) to store large amounts of data in memory, thereby shifting the performance bottleneck of query execution from disk accesses to CPU efficiency (i.e., instruction count and cycles per instruction). One technique used to achieve such efficiency in analytical applications is batch-oriented processing or \textit{vectorization}: it reduces interpretation overhead, improves cache locality, and allows for efficient loop optimizations (e.g., loop unrolling, SIMD vectorization). For each \textit{vector} (i.e., a batch of tuples), the DBMS needs to track the set of valid tuples after a predicate application. To that end, systems employ one of two data structures, or \textit{filter representations}: Selection Vectors (SelVecs) or Bitmaps. In this work, we analyze each approach's strengths and weaknesses to provide recommendations on how to implement vectorized operations. Through a wide range of micro-benchmarks, we determine that the optimal implementation strategy is a function of many factors: the cost of iterating through tuples, the cost of the operation itself, and how amenable it is to SIMD vectorization. Our analysis shows that Bitmaps perform better for operations that can be efficiently vectorized using SIMD instructions but that SelVecs perform better on all other operations due to a cheaper iteration logic.
\end{abstract}

\begin{acknowledgments}
I would like to thank my advisor Andy Pavlo for his guidance and help during my last two years at CMU. He has developed my interest in databases and computer science research in general and made my experience at CMU very educational and fun. I would also like to thank my amazing peers of the CMUDB group that helped me in this and other projects: Prashanth Menon, Matt Buttrovich, Wan Shen Lim, Lin Ma, Tanuj Nayak, Rohan Aggarwal, Gustavo Angulo, Tianyu Li, John Rollinson. Special shoutout to Prashanth Menon, whose help was invaluable in this project. Finally, I would like to thank my parents and siblings who guided and supported their son/little bro all throughout his life.
\end{acknowledgments}

\tableofcontents
\listoffigures
\listoftables

\mainmatter

%% Double space document for easy review:
%\renewcommand{\baselinestretch}{1.66}\normalsize

% The other requirements Catherine has:
%
%  - avoid large margins.  She wants the thesis to use fewer pages, 
%    especially if it requires colour printing.
%
%  - The thesis should be formatted for double-sided printing.  This
%    means that all chapters, acknowledgements, table of contents, etc.
%    should start on odd numbered (right facing) pages.
%
%  - You need to use the department standard tech report title page.  I
%    have tried to ensure that the title page here conforms to this
%    standard.
%
%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
%
% Other than that, just make it look good...

\chapter{Introduction and Background}
\label{chapter1}
DBMSs typically handle two broad classes of applications: Online Transaction Processing (OLTP) and Online Analytical Processing (OLAP) applications. OLTP applications tend to be short-running, write-heavy queries that only touch a few entries in a database. OLAP applications are long-running, read-mostly queries that analyze large amounts of data. Due to the unique characteristics of each class, storage engines (i.e., the component of the DBMS stores the data), and query engines (i.e., the component that executes queries) must specialize storage formats and database algorithms to efficiently handle OLAP or OLTP applications \cite{one_size_fits_all}.

OLTP applications aim for fast access and update on individual entries. Therefore, to improve cache performance, OLTP-specialized storage engines \cite{mysql,postgres,aurora,voltdb}, use a row-store (i.e., a storage format that stores all the attributes of an entry contiguously in memory and in cache). OLAP applications, on the other hand, optimize for fast reads and computing on attributes across multiple entries (e.g., to find the average value of an attribute across all tuples). OLAP-specialized storage engines \cite{redshift,vertica,saphana,duckdb}, thus, use a column-store (i.e., a storage format that stores similar attributes of different entries contiguously). Though most traditional DBMSs \cite{postgres,mysql} focused on OLTP applications and thus used row stores, column stores have resulted in significant performance gains for OLAP applications \cite{abadi_phd,one_size_fits_all}.

Besides the storage engine, disk-accesses were, historically,  the performance bottleneck of OLAP applications. For example, it takes around $1ns$ to perform random access in the  L1 cache, $100ns$ for DRAM, and $17000ns$ for SSDs \cite{numbersbyyear}. Disk accesses, thus, dominate the execution time of any query whose data is primarily disk-resident. Recent advances in DRAM manufacturing have exponentially decreased the price and increased the capacity of main-memory. The cost of DRAM per GB was around \$1000 in 2000, \$10 in 2010, and less than \$1 in 2020 \cite{dramprices}. This dramatic decrease in cost, and the corresponding increase in capacity, have both obeyed Moore's Law \cite{moorelaw}. It is now possible to store large databases entirely in DRAM. Rather than optimize for disk access, the query engines of in-memory DBMSs (i.e., DBMSs whose data fits in memory) must optimize for CPU efficiency by reducing instruction count and cycles per instruction when executing queries.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/SampleQuery.png}
    \caption{\textbf{Sample SQL Query and Plan.}}
    \label{fig:intro_query}
\end{figure}

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Volcano.png}
    \caption{\textbf{The Iterator Model on the sample SQL query} -- The code in red indicates virtual function calls.}
    \label{fig:volcano_graph}
\end{figure}

The traditional query execution technique, the Iterator Model \cite{volcano}, is ill-suited for in-memory DBMSs due to CPU efficiency reasons. It executes a query by interpreting the query plan. In this model, each database operator implements a virtual \texttt{next()} function that returns the next tuple it can generate; each expression (i.e. addition, multiplication) also provides an virtual \texttt{evaluate()} function that takes in a tuple and returns a value. \cref{fig:volcano_graph} showcases the Iterator model with the sample query in \cref{fig:intro_query}. The operators \texttt{Scan A} and \texttt{Scan B} iterate through their respective tables and output the tuples to their parent operators. The filter only outputs tuples that satisfy the \texttt{WHERE} clause. The join first builds a hash table, then probes it to find matches. For every tuple -- of which there can be billions in an OLAP setting -- the virtual function calls (marked in red in the figure) constitute a significant overhead, which decreases CPU efficiency. Since the 2000s, there has been much work dedicated to optimizing the CPU efficiency of in-memory column stores for high-performance OLAP applications.

\citeauthor{monetdb2} proposed an alternative to the Iterator Model: the Materialization Model \cite{monetdb2}. In this model, rather than process and output one tuple at a time, each operator processes all its input tuples, then outputs all its results to the next operator. The authors also introduce specialized algorithms for each relational operator to exploit the columnar format of their system. Although the Materialization Model removes the Iterator Model's interpretation overhead, it can become memory-bound due to frequent materializations (e.g., copies) of large columns. To solve this issue, \citeauthor{boncz_vector} proposed a middle ground between the Materialization and Iterator Model: the Vectorization Model \cite{boncz_vector}. The thesis explores strategies for speeding up vectorized queries (i.e., queries executed using the Vectorization Model). As such, we will dedicate the rest of this chapter to an overview of the Vectorization Model. We discuss alternative query engine designs and optimizations for in-memory analytical DBMSs in \cref{chapter5}.

\section{The Vectorization Model}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Vectorization.png}
    \caption{\textbf{The Vectorization Model with the Sample Query} -- The code is nearly identical to that in \cref{fig:volcano_graph}, except that it processes vectors.}
    \label{fig:vectorization_graph}
\end{figure}

The Vectorization Model is almost identical to the Iterator model except that it operates on batches of $T$ tuples or \textit{vectors} (e.g., 2048 tuples at a time). The \texttt{next()} function returns vectors and the \texttt{evaluate()} function operates on vectors too. \cref{fig:vectorization_graph} showcases Vectorization Model on the sample query in \cref{fig:intro_query}; it is near identical to \cref{fig:volcano_graph}. The difference is that processing occurs one vector at at time rather than one tuple at a time.

\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MulByConstSISD.png}
 \caption{Unoptimized version.}
  \label{fig:vector_mul_sisd}
\end{subfigure}%
\begin{subfigure}[t]{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MulByConstSIMD.png}
 \caption{Optimization version. There are two optimizations here: SIMD vectorization and loop unrolling with a factor of 2. The code in green indicates Intel's AVX512 instructions.}
  \label{fig:vector_mul_simd}
\end{subfigure}
\caption{\textbf{A Vectorized Multiplication Function} -- The programmer usually writes only the code on the left; the compiler can perform the optimizations on the right.}
\label{fig:vector_mul}
\end{figure}

This batch-oriented processing provides several benefits. First, virtual function calls occur for every $T$ tuples, which amortizes their overhead by a factor of $T$. Second, materializing $T$ tuples at a time instead of all tuples eliminates the memory pressure of the Materialization Model. Third, functions that operate on vectors, also known as \textit{vectorized primitives}, are usually amenable to loop optimizations (e.g., loop unrolling, SIMD vectorization) applied by either the compiler or the programmer. Consider, for a example, the primitive shown in \cref{fig:vector_mul_sisd}. It multiplies the elements in a vector by a constant number, and stores the result in a new vector. The loop of this primitive has independent iterations. Furthermore, the body of the loop only contains a load, a multiplication, and a store, which can all operate on SIMD vectors. As a result, the compiler or programmer can unroll and SIMD vectorize the loop. \cref{fig:vector_mul_simd} shows one way to optimize the loop. The body now contains AVX512 instructions \cite{intel2019} that perform that load, the multiplication, and the store, each operating on eight elements at a time. Besides, the loop unrolling (by a factor of two) optimization reduces the cost of loop maintenance.

Despite these benefits, the Vectorization Model introduces a new challenge. Individual tuples within a vector may be \textit{filtered out}, meaning that subsequent operators should not process them. For example, in \cref{fig:vectorization_graph}, the \texttt{Join} operator should not insert tuples that do no satisfy the \texttt{WHERE} clause in the \texttt{JoinHashTable}. It should only insert \textit{selected} tuples -- those not filtered out. Next, we discuss the methods used to identify selected tuples within a vector.

\section{Filter Representation}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.7]{images/FilterRepresentationIntro.png}
    \caption{\textbf{Filter Representation} -- Selection Vectors (left) index into selected tuples. Bitmaps (right) are only set on selected rows.}
    \label{fig:repr_intro}
\end{figure}

A \textit{filter representation} is the data structure used to identify selected tuples with a vector. \cref{fig:repr_intro} displays the two common representations: Selection Vectors (SelVecs) and Bitmaps. A SelVec stores a list of indices into the selected tuples. Tuples at an index stored within the SelVec are selected; tuples without an index are implicitly filtered out. A Bitmap, on the other hand, stores a list of bits. Its set bits indicate the selected tuples; its unset bits indicate filtered out ones. In the rest of this thesis, we will use the noun \textit{filter} to denote a particular instance of either data structure.

\subsection{Contribution}
Previous works on vectorized execution have chosen a filter representation strategy without providing an experimental comparison of the two approaches. Vectorwise \cite{vectorwise}, and works derived from it \cite{miro_adapt,everything_vectorized,sompolski_vec},  rely on SelVecs. IBM DB2's BLU \cite{ibm_blu} and the more recent VIP \cite{orestis_bitmap} rely on Bitmaps for the intermediary results of a table scan's \texttt{WHERE} clause filters, and SelVecs for other relational operators. Nonetheless, we find that supporting both representations, and dynamically choosing between them results in better performance than static implementations. Depending on the specific primitive and on the selectivity (i.e., the ratio of selected tuples) of its input vector, SelVecs can outperform Bitmaps and vice-versa. This work's main contribution is a methodology on how to optimize the performance of arbitrary vectorized primitives by taking into account filter representation, selectivity, and loop optimizations. We also provide recommendations for developers working on vectorized query engines on how to best implement primitives.

The remainder of this thesis is structured as follows. In \cref{chapter2}, we analyze the factors that influence the performance of vectorized primitive. We then derive insights for developers implementing vectorized query processing in \cref{chapter3}. Next, we use these insights on the OLAP queries of the TPCH \cite{tpch} benchmark and show the performance gains we obtain in \cref{chapter4}. Finally, we discuss related work in \cref{chapter5} and conclude in \cref{chapter6}.

\chapter{Computing on Filtered Vectors}
\label{chapter2}
The purpose of this chapter is to discuss the factors that influence the performance of vectorized primitives. We first categorize the kinds of operations that primitives perform and the possible ways to implement them. Afterward, we will present and explain the equation that quantifies the performance of each implementation. Finally, we raise the main questions whose answers, provided in \cref{chapter3}, will determine how to implement primitives optimally.

\section{Operations on Filtered Vectors}
\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/UpdateIntro.png}
 \caption{Update operation.}
  \label{fig:update_intro}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MapIntro.png}
 \caption{Map operation.}
  \label{fig:map_intro}
\end{subfigure}
\caption{\textbf{Operations on Filtered Vectors} -- Updates change the set of selected tuple. Maps apply a function to the set of selected tuples without updating the set. Selected input tuples are in green. Selected output tuples are in blue.}
\label{fig:update_map_intro}
\end{figure}
\cref{fig:update_map_intro} shows the two kinds of primitives we seek to optimize: Updates and Maps. Updates apply a predicate function to the vector and modify the set of selected tuples accordingly. A scan filter (i.e., \texttt{WHERE} clause) is an example of an update operation. On the other hand, Maps keep the selected set but compute a new vector using a mapping function. A projection (i.e., \texttt{SELECT} clause) is an example of a map operation. There are other primitives, but these often have side-effects (e.g., insertion in a hash table for joins or an array for sorting) and are, therefore, not amenable to our optimizations for correctness reasons (e.g., Full Compute introduced in \cref{computestrategies}). As such, we focus on side-effect-free primitives.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.45]{images/NewJoinProbeExample.png}
    \caption{\textbf{Multi-Step Join Probe} -- The color green indicates selected items in the input vector. Orange indicates hash table lookup hits. Blue indicates exact matches after key comparison; these represent the output of the probe. Greyed out squares are filtered out tuples, or empty hash table chains.}
    \label{fig:probe_example}
\end{figure}
All relational operators are decomposable into a set of consecutive primitives. As mentioned above, simple relational operators like scan filters or projections consist of one individual primitive. Complex operators, like joins and aggregations, consist of multiple primitives. As a proof of concept, we showcase, in detail, the example of join probes in \cref{fig:probe_example}. The first primitive computes the hashes of the input tuples; the result is a vector of hash values. It is, therefore, a Map primitive. The second primitive performs a hash table lookup using the hashes to obtain a vector of hash table entries. It is also a Map primitive. Due to hashing collisions, the entries may not match the input, so the algorithm compares the join keys to filter out inexact matches. For example, key 90's hit in the hash table is an exact match, whereas key 73's is not. The third primitive performs this comparison; it is an Update primitive. The fourth primitive advances the hash table chains to obtain the next entries (e.g., 85 follows 37); it is a Map primitive. The probe then loops back to the comparison primitive to find exact matches (e.g., 85 now has an exact match). The set of exact matches constitutes the output of the join probe.

Other relational operators (e.g., aggregations) are similarly decomposable.

\section{Transitions}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DisjunctionHowto.png}
    \caption{\textbf{Decomposed Disjunctions} -- We apply the first clause (in orange) on all input tuples. In the second step, however, we only apply the second clause (in red) on those tuples that do not pass the first clause. In the third step, we take the union between the two outputs. The output tuples are in blue.}
    \label{fig:disjunctive_filter_howto}
\end{figure}

In between each primitive, operators may need to perform a \textit{transition} (i.e., modifications to the filter passed into the next primitive). We now discuss transitions and their associated cost. Most of the time, this cost is non-existent because the same filter is passed on to the next primitive: the transition is \textit{implicit}. For example, the hashing primitive and the lookup primitive in \cref{fig:probe_example} have the same filter; the transition performs no modifications. On the other hand, the Chain Following primitive operates on entries that do not satisfy the comparison: the transition involves a set difference. Set unions are another type of transition. Decomposed disjunctions \cite{pcq} (shown in \cref{fig:disjunctive_filter_howto}), for example, involve both set differences and unions. They decompose disjunctive \texttt{WHERE} clauses into a sequence of primitives -- one each for each conditional term. Each primitive operates on only tuples that have not satisfied the previous conditional terms, hence the set differences. The final filter consists of tuples that satisfy at least one primitive, hence the unions.

To the best of our knowledge, transitions are either implicit or take the form of set differences (for elements that do not pass the previous predicate) and unions (to obtain the final result). These operations are efficient to implement with Bitmaps by leveraging single-cycle bitwise AVX512 instructions \cite{intel2019} that operate on 512 bits at a time. Efficiently implementing set operations on SelVecs is more complicated.  Whenever necessary, Updates can write to one additional SelVec for tuples that do not pass the filter; this second SelVec implicitly corresponds to the result of a set difference. Note that it does not overlap with the first SelVec, so a union of any of their subsets is just an array append, a fast operation that can leverage SIMD instructions.

To show that the overhead of these set operations is similar and negligible for both Bitmaps and SelVecs, we run an experiment that performs the decomposed disjunction shown in \cref{fig:disjunctive_filter_howto}. We synthetically generate the data so that each tuple has a $50\%$ chance of satisfying either conditional terms, the average case scenario. The data contains around $2000000$ tuples split into $1000$ vectors. The results are shown in \cref{fig:disjunctive_filter_perf}. The overheads are negligible; they are also nearly equal ($3.55\%$ and $4.06\%$ for Bitmaps and SelVecs, respectively). The representation, therefore, matters little when it comes to transitions. Given this, we will focus on optimizing primitives and will ignore transition costs.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{eval/transition.pdf}
    \caption{\textbf{Transition Overheads} -- The overheads are $3.55\%$ and $4.06\%$ for Bitmaps and SelVecs respectively.}
    \label{fig:disjunctive_filter_perf}
\end{figure}

\section{Compute Strategies}
\label{computestrategies}
\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/SelectiveCompute.png}
 \caption{Selective Compute. An \texttt{x} indicates an uninitialized value.}
  \label{fig:compute_intro_selective}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/FullComputeMap.png}
 \caption{Full Compute. There are no uninitialized values}
  \label{fig:compute_intro_full}
\end{subfigure}
\caption{\textbf{Compute Strategies}}
\label{fig:compute_intro}
\end{figure}

A compute strategy determines the implementation of Map and Update primitives. We consider three compute strategies: (1) Selective, (2) Full, and (3) Mixed Compute. Selective Compute only applies the given functions on selected tuples, whereas Full Compute applies them on all tuples. \cref{fig:compute_intro} shows each approach's effect. The Selective Compute strategy (\cref{fig:compute_intro_selective}) leaves undefined values because it does not operate on filtered out entries, whereas the Full Compute strategy (\cref{fig:compute_intro_full}) operates of every tuple and thus produces a result for each. As we show in \cref{chapter3}, although Full Compute usually performs more work, it benefits from SIMD vectorization, simple loop structure (for loop unrolling and interleaving), and easy branch prediction. To exploit this trade-off, the third strategy, Mixed Compute, switches from Full Compute to Selective Compute when the selectivity (i.e., the ratio of selected tuples) goes below a threshold. The next chapter will show how to derive this threshold experimentally.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/FullComputeUpdate.png}
    \caption{\textbf{Update with Full Compute} -- The leftmost vector is the input vector; its selected elements are shown in green. The middle vector is the intermediary output of the full compute; its selected elements are shown in orange. The last vector has the intersection of the input filter and the intermediary filter; its selected elements are shown in blue.}
    \label{fig:full_compute_update}
\end{figure}

On Update primitives, Full Compute first finds the set of all tuples in the vector that pass the predicate, regardless of the input filter. The final filter is the intersection of this intermediary filter and the input filter. This process is shown in \cref{fig:full_compute_update}. Notice how the intermediary filter contains elements filtered out of the input vector (e.g., 100 and 200). The intersection is, thus, necessary to remove these elements from the final output. Bitmap intersection is efficient thanks to the AVX512 \texttt{and} instruction that intersects 512 bits in one cycle \cite{intel2019}, but SelVecs require a slow intersection of sorted sets \cite{sorted_set}. Thus, Full Compute is not efficiently compatible with SelVecs on Update primitives. This incompatibility will become important when choosing a strategy to implement.

\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{images/ConditionalStoreConditional.png}
 \caption{Conditional Store. We only update the SelVec when the condition is satisfied (code in green).}
  \label{fig:conditional_store}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{images/ConditionalStoreUnconditional.png}
 \caption{Unconditional Store. We always update the SelVec regardless of the condition (code in green). The condition determines whether to increment the write index or not.}
  \label{fig:unconditional_store}
\end{subfigure}
\caption{\textbf{Conditional versus Unconditional Store} -- Both listings select the elements within the input vector (first parameter) that are less than the second parameter. Initially, the SelVec (third parameter) contains selected indices of the input vector. In the end, it only contains those that pass the predicate.}
\label{fig:conditional_selvec}
\end{figure}

Furthermore, for Updates with SelVecs, there is another implementation decision: Conditional versus Unconditional Store. \cref{fig:conditional_selvec} shows these options. Both primitives attempt to select the values less than a given input constant. They differ only in how they update the SelVec (code in green). The Conditional Store (\cref{fig:conditional_store}) updates the SelVec when the condition is true. The Unconditional Store (\cref{fig:unconditional_store}) always the updates the SelVecs, but only increments the write index when the condition is true. We do not focus on this decision, for it has been studied extensively before \cite{miro_adapt, sompolski_vec}. Previous work identified branch mispredictions costs as the main factor influencing this decision. The Conditional Store implementation is competitive when the same branch is taken around $90\%$ of the time or more, meaning that less than $10\%$, or more than $90\%$ of tuples satisfy the condition. All the experiments in this thesis use the optimal strategy on Update primitives that use SelVecs.

Given a compute strategy and a filter representation, the next section provides a framework for analyzing the performance of any vectorized primitive.

\section{Primitive Performance}
\label{prim_perf_section}

\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleMapFull.png}
 \caption{Full Compute. This function iterates through all elements, including filtered out ones.}
  \label{fig:rationale_full}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%  
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleMapBitmap.png}
 \caption{Selective Compute with a Bitmap. The iteration logic is adapted from \cite{bitmap_iteration}. The logic is not amenable to SIMD vectorization.}
  \label{fig:rationale_bitmap}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%  
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RationaleMapSelVec.png}
 \caption{Selective Compute with a SelVec. Reading an element involves an indirection to first obtain its index. In the SIMD code, the \texttt{gather} instruction handles this indirection. The last \texttt{scatter} is only necessary to maintain consistent indices between the input and output vector during multi-step Map operations. It should otherwise be avoided due to its slowness.}
  \label{fig:rationale_selvec}
\end{subfigure}
\hspace*{\fill}%
\vspace*{8pt}%

\caption{\textbf{Implementing Vector Primitives} -- The code in green indicates the core operation (multiplication). The code in red indicates iteration logic.}
\label{fig:rationale_code}
\end{figure}

This section analyzes the performance of primitives according to their implementations. Our primitive implementations follow one of the patterns shown in \cref{fig:rationale_code}. This listing shows the strategies to multiply a vector by a constant (a Map primitive) along with their optimized versions using SIMD instructions: Full Compute (which does not depend on representation), Selective Compute with SelVecs, and Selective Compute with Bitmaps. In each instance, the code is a loop divided in two parts: the \textit{iteration logic} (code in red), and the \textit{core operation} (code in green). The core operation (e.g., multiplication by a constant) is independent of the strategy. On the other hand, the iteration logic is strategy dependent; it determines how to iterate through tuples -- whether selected ones in the case of Selective Compute or all tuples in the case of Full Compute -- to perform the core operation and store its result.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/PerfEq.png}
    \caption{\textbf{The Primitive Performance Equation}}
    \label{fig:perf_equation}
\end{figure}

The equation in \cref{fig:perf_equation} shows the formula to compute the running time of a primitive. The first equality indicates that the total running time is equal to the number of tuples processed multiplied by the time spent per tuple. The second equality splits the time spent per tuple in two: the time dedicated to the iteration logic, and the time dedicated to the core operation. This formula provides a framework to analyze any vectorized primitive. Let us consider how each of the three factors in the second equality relates to implementation strategies.

\textbf{Number of Tuples Processed}: Selective Compute processes only selected tuples; Full Compute processes every tuple. Thus, the lower the selectivity (i.e., the ratio of selected tuples), the more wasteful Full Compute is.

\textbf{Iteration Logic Time per Tuple}: Full Compute, while processing more tuples, has the simplest iteration logic because it does not aim to identify the selected ones. For Selective Compute, iterating over a Bitmap (e.g., \cref{fig:rationale_bitmap}) is more expansive than iterating over a SelVec (e.g., \cref{fig:rationale_selvec}) because of the bit operations required by Bitmaps. Optimizations like loop unrolling and SIMD vectorization can further decrease the iteration logic time per tuple, but they are only available for Full Compute and SelVecs. The Bitmap's selective iteration logic is too complex to perform such optimizations.

\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{images/BranchingCodeSISD.png}
 \caption{SISD code.}
  \label{fig:branching_sisd}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{images/BranchingCodeSIMD.png}
 \caption{SIMD code.}
  \label{fig:branching_simd}
\end{subfigure}
\caption{\textbf{SISD Branching Code versus SIMD} -- Both listings select elements from one of \texttt{vec1}, or \texttt{vec2} depending on the \texttt{condition}. In the SISD code, only one branch is executed every iteration. In the SIMD code, we rely on masking to select which vector to load from, but all instructions are executed at every iteration.}
\label{fig:branching_code}
\end{figure}

\textbf{Core Operation Time per Tuple}: The core operation is the same for all strategies. The only way to reduce its contribution to the running time is through data-parallel SIMD instructions, which are only available with Full Compute and Selective Compute with SelVecs. There are, however, situations where SIMD vectorization provides little gain or even hurts performance. Consider the example in \cref{fig:branching_code}. This function selects an element from one of two vectors depending on the value in a third boolean vector. It corresponds to a relational \texttt{CASE} statement. For each loop iteration, the SISD code (\cref{fig:branching_sisd}) only executes one branch. The SIMD code (\cref{fig:branching_simd}) executes all branches and relies on SIMD masking to set the correct vector lanes; it ends up executing more instructions that the SISD code, which reduces the benefits data-parallelism. In general, core operations benefit most from SIMD vectorization when they only contain straight-line (i.e., no branches), arithmetic, and bitwise instructions.

The explanation above raises three questions that can allow us to determine the optimal strategy for any primitive: What is the input vector's selectivity? Can we use data-parallelism to perform the core operations on multiple elements at a time? Does the core operation contain instructions that make data-parallelism inefficient (e.g., branching)? The next chapter experimentally derives the best strategies according to the answers to these questions.

\chapter{Optimal Strategies}
\label{chapter3}
In this chapter, we aim to derive the optimal execution strategy for any given primitive experimentally. In doing so, we will provide a decision tree that developers can use to guide the implementation of vectorized primitives.

\begin{table}[t!]
\centering
\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Strategy Name} & \textbf{Representation} & \textbf{Compute} & \textbf{SIMD}      & \textbf{Compatibility} \\ \hline
SelVecPartial         & Selection Vectors       & Selective        & None               & Update and Map         \\ \hline
SelVecManual          & Selection Vectors       & Selective        & Manually Written   & Update and Map         \\ \hline
BitmapPartial         & Bitmaps                 & Selective        & None               & Update and Map         \\ \hline
BitmapFull            & Bitmaps                 & Full             & Auto Vectorization & Update and Map         \\ \hline
BitmapFullManual      & Bitmaps                 & Full             & Manually Written   & Update and Map         \\ \hline
Full                  & Either                  & Full             & Auto Vectorization & Map                    \\ \hline
FullManual            & Either                  & Full             & Manully Written    & Map                    \\ \hline
\end{tabular}
\caption{\textbf{Implemented Strategies} -- For the SIMD columns, `Auto Vectorization' indicates that we rely on compiler auto vectorization, whereas `Manually written' indicates that we write SIMD code ourselves using intrinsics. On the compatibility column, Update operations cannot use both Full Compute and Selection Vectors, which explains the last two entries.}
\label{tab:strategies}
\end{table}

The chapter is divided into a series of experiments based on the questions raised in \cref{prim_perf_section} to determine the optimal strategy for a given primitive. \cref{tab:strategies} summarizes all the strategies we implemented. As noted before, there is no Full Compute strategy for Updates that works with SelVecs. Update experiments will thus always use a Bitmap for Full Compute; For Map experiments, the representation is irrelevant because Full Compute neither reads nor updates the filter. In addition to relying on compiler auto-vectorization, we also experimented with manual vectorization (e.g., with SelVecManual) for straight-line arithmetic and bitwise operations (as in \cref{sldpsection}).

We run the experiments on an Intel Xeon Platinum 8124M CPU @ 3.00GHz with AVX512 support. We use the NoisePage DBMS \cite{noise_page}, with a read-only column store as its storage engine, compiled with clang version 9 \cite{clang}.

In each experiment, we synthetically generate the data and manually vary the selectivity of each primitive's input vector from $0.0$ to $1.0$ in increments of $0.05$ to show the impact selectivity has on what the optimal strategy is. For example, we set a random $25\%$ of bits in a Bitmap to obtain filter with a selectivity of $0.25$. Each experiment executes its primitive on enough vectors to obtain a stable average running time. For example, fast primitives are executed in the order of $10^6$ times, whereas the slower ones are executed in the order of $10^4$ or $10^5$ times. Map primitives materialize their results on a vector in-memory, but Update primitives only modify the input filter without materializing a new vector.

\section{Non Data-Parallel Core Operations}
\label{nondataparallel}
First, we consider core operations that are not data-parallel. These are the operations where it is impossible to use SIMD instructions to process multiple tuples at a time. We experiment on two kinds of such operations: those that operate on variable-length data (e.g., string operations) and those that use instructions without a SIMD counterpart (e.g., integer division, modulo).

\subsection{Variable Length Data}
\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_string_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE country < `Japan'}}
  \label{fig:varlen_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_string_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT POSITION(`a' in country)}.}
  \label{fig:varlen_map}
\end{subfigure}
\caption{\textbf{Variable Length Data} -- Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
\label{fig:varlen_map_update}
\end{figure}

We first analyze the performance of operations on variable-length data, as is often the case for string operations (e.g., string comparison, sub-string). Let us consider the factors in \cref{fig:perf_equation}. Here, the core operation executes a variable number of instructions depending, for example, on the length of the input strings. As we show in the next experiment, it dominates the iteration logic time, even for short strings. Because data parallelism is not available, the core operation time per tuple is the same across all strategies. The principal factor that differentiates strategies is, therefore, the number of tuples processed, which favors Selective Compute strategies over Full Compute ones.

To confirm this intuition, we run a micro-benchmark using simple string operations on short strings. The Update micro-benchmark performs string comparison, and the Map micro-benchmark performs character location. The strings are all short country names (e.g., USA, Senegal, China). The purpose is to show that the number of tuples processed is the principal performance factor, even for relatively cheap variable-length operations.

The results are shown in \cref{fig:varlen_map_update}. We see that Full Compute (with BitmapFull or Full) consistently performs worst because it processes more tuples, except at full selectivity. At full selectivity, Full Compute and Selective Compute process the same number of tuples, but the former slightly benefits from a simpler iteration logic. We also see that the performance of SelVecPartial is similar to that of BitmapPartial, despite the former's simpler iteration logic, meaning that the core operation is the dominating cost.

The optimal strategy in this scenario is SelVecPartial. It processes fewer tuples than Full Compute strategies and has a slightly cheaper iteration logic than BitmapPartial (though the performance difference is small). Full Compute is only marginally competitive at full selectivity; its benefits are mostly negligible; the more expansive the core operation, the worse Full Compute becomes.

\subsection{Integer Division}
\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_division_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 \% col2 < val}.}
  \label{fig:division_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/simple_division_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 / col2}.}
  \label{fig:division_map}
\end{subfigure}
\caption{\textbf{Integer Division.}}
\label{fig:division_map_update}
\end{figure}
Second, we analyze the integer division instruction as it does not have a SIMD counterpart. Let us consider the factors in \cref{fig:perf_equation}. Integer division is so expansive that the iteration logic time is negligible compared to the core operation time. Because data parallelism is not available, its cost per tuple is the same across all strategies. Once again, the main factor that differentiates strategies is the number of tuples processed, which favors Selective Compute.

The micro-benchmarks of this section evaluate an Update and a Map primitive that both contain an integer division. The results are shown in \cref{fig:division_map_update}. The explanation of its results is similar to \cref{fig:varlen_map_update}'s: the cost of the integer division and the unavailability of data-parallelism make Full Compute inefficient. SelVecPartial is, once again, the optimal strategy, but only has a slight edge over BitmapPartial due to the former's simpler iteration logic.

We conclude that the number of tuples processed in the dominant performance factor for non-data-parallel primitives, making Full Compute impractical. Because of the small impact of iteration logic time, SelVecPartial has a small performance edge over BitmapPartial, but developers can choose either representation without much affecting performance.

\section{Inefficient Data Parallelism}
\label{inefficientsimd}
\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_and_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \&\& col2 < val2}.}
  \label{fig:logical_and_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_and_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \&\& col2 < val2}.}
  \label{fig:logical_and_map}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_or_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 || col2 < val2}.}
  \label{fig:logical_or_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/logical_or_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 || col2 < val2}.}
  \label{fig:logical_or_map}
\end{subfigure}

\caption{\textbf{Branching Operations} -- Each operation contains one branch because of boolean short-circuiting.}
\label{fig:logical_map_update}
\end{figure}

\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_and_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \& col2 < val2}.}
  \label{fig:bitwise_and_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_and_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \& col2 < val2}.}
  \label{fig:bitwise_and_map}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_or_update.pdf}
 \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 | col2 < val2}.}
  \label{fig:bitwise_or_update}
\end{subfigure}%
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/bitwise_or_map.pdf}
 \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 | col2 < val2}.}
  \label{fig:bitwise_or_map}
\end{subfigure}

\caption{\textbf{Bitwise Operations} -- The increase in performance compared to \cref{fig:logical_map_update} is due to the lack of branching.}
\label{fig:bitwise_map_update}
\end{figure}

To study the effect of inefficient use of SIMD instructions, we now consider core operations with branching code. \cref{chapter4} will discuss primitives that contain pointer manipulation and several memory accesses. SIMD strategies (e.g., BitmapFull) benefit from processing multiple tuples at a time, but suffer from executing more code than SISD strategies (e.g., SelVecPartial). Their core operation time per tuple may even decrease depending on the branching structure (e.g., having to execute all branches of a switch statement).  Thus, if Full Compute can outperform Selective Compute at all, it will do so at the highest selectivities because the small reduction in time spent per tuple cannot compensate for the higher number of tuples processed. Among SISD strategies, SelVecPartial will outperform BitmapPartial due to its simple iteration logic.

To determine the impact of a single branch within the core operation, we implemented primitives that perform logical and/or operations. Due to boolean short-circuiting, these primitives will contain exactly one branch. We have confirmed that the compiler manages to auto-vectorize Full Compute strategies. The results are shown in \cref{fig:logical_map_update}. We can see that Full Compute is either always the worst strategy (e.g., in \cref{fig:logical_or_update}), or only competitive at high selectivities ($>= 0.85$ in \cref{fig:logical_or_map}). To confirm that branching is indeed responsible for the poor performance of Full Compute, we implemented branch-less primitives with bitwise and/or operations rather than logical ones. The results are in \cref{fig:bitwise_map_update}. Full Compute now outperforms Selective Compute for selectivities $<= 0.2$, confirming that branching code significantly hurts SIMD strategies' performance.

In conclusion, though Full Compute can sometimes be competitive at the highest selectivities, we recommend, for the sake of simplicity, the SelVecPartial strategy.

\section{Straight-Line and Data-Parallel Core Operations}
\label{sldpsection}

We now consider core operations with Straight-line and data-parallel (SLDP) code, i.e., non-branching code that can efficiently leverage SIMD instructions. Primitives that only perform arithmetic (without integer division) and bitwise instructions fall under this category. Let us consider the equation in \cref{prim_perf_section} to analyze the performance of each strategy in this scenario. Full Compute strategies (e.g., BitmapFull, Full) can leverage SIMD instructions to significantly reduce the iteration logic and core operation time per tuple. AVX512 \cite{intel2019} integer addition instructions, for example, performs 8 parallel 64-bit additions in a single cycle. We, therefore, expect Selective Compute strategies to only be competitive when the selectivity is very low, meaning that Full Compute processes far more tuples. SelVecManual is the exception: it is a Selective Compute strategy that uses SIMD instructions. Its gain in iteration logic time is, however, not as high as that of Full Compute strategies because it uses a \texttt{gather} instruction \cite{intel2019} to collect the elements at the selected indices (shown in \cref{fig:rationale_selvec}). The higher the core operations time per tuple, the less important the \texttt{gather} overhead because iteration logic time becomes more and more insignificant. Thus, we expect SelVecManual to be competitive with Full Compute at medium or below selectivities depending on the core operation. The next experiments will quantify the differences between each strategy.

\begin{figure}[t!]
    \centering
    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_update.pdf}}
    \caption{\textbf{SNBS Update Operations} -- For Updates, SelVecManual relies on the \texttt{compress\_store} SIMD instruction rather than on \texttt{scatter}, which reduces its overhead.}
    \label{fig:snbs_update}
\end{figure}

\begin{figure}[t!]
    \centering
    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_map.pdf}}
    \caption{\textbf{SNBS Map Operations} -- For Maps, SelVecManual relies on the \texttt{scatter} SIMD instruction, which explains why it initially does not perform better than SelVecPartial.}
    \label{fig:snbs_map}
\end{figure}

\begin{figure}[t!]
\captionsetup[subfigure]{justification=justified}
\centering
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/snbs_thres_update.pdf}
 \caption{Update thresholds between BitmapFullManual and Selective Compute strategies.}.
  \label{fig:snbs_thres_update}
\end{subfigure}
\begin{subfigure}[t]{.49\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/snbs_thres_map.pdf}
 \caption{Map thresholds between FullManual and Selective Compute strategies.}
  \label{fig:snbs_thres_map}
\end{subfigure}
\caption{\textbf{Mixed Compute Thresholds} -- The Full Compute strategies are BitmapFullManual and FullManual because compiler auto-vectorization does not consistently use AVX512 registers.}
\label{fig:snbs_thres}
\end{figure}

We implemented Update and Map primitives in which the core operation contains single-cycle arithmetic and bitwise instructions. We progressively increase the number of such instructions to increase the core operations time per tuple and quantify the selectivity thresholds above which Full Compute strategies become faster than Selective Compute ones. The results are shown in \cref{fig:snbs_update,fig:snbs_map} (for Updates and Maps). The selectivity thresholds for each graph are respectively in \cref{fig:snbs_thres_update,fig:snbs_thres_map}. Let us analyze the results in detail:
\begin{itemize}
    \item In all experiments, SelVecPartial performs better than BitmapPartial. As explained in the previous chapter, this is only due to a simpler iteration logic.
    \item SelVecManual is always the best Selective Compute strategy because it uses data-parallelism to reduce the time spent per tuple.
    \item Manual Vectorization (with FullManual and BitmapFullManual) sometimes performs much better than Auto-Vectorization (with Full and BitmapFull). For example, in \cref{fig:snbs_update}'s graph for \texttt{NumOps=13}, BitmapFullManual is $1.8$ times faster than BitmapFull. With further investigation of the generated code, we discovered that the compiler is overly conservative when it comes to using AVX512. AVX512 registers result in decreased CPU frequency \cite{avx512_registers}, so the compiler is careful not always to use them. It often uses AVX2 registers instead. We, on the other hand, always use AVX512.
    \item The thresholds for SelVecPartial and BitmapPartial, shown in \cref{fig:snbs_thres_update,fig:snbs_thres_map}, prove that these strategies are only competitive with Full Compute at low selectivities ($<= 0.2$) because, as explained above, they do not use SIMD instructions.
    \item The threshold for SelVecManual, on the other hand, increases with the number of operations because its iteration logic overhead becomes more and more insignificant as the core operation time increases. SelVecManual thus benefits from data-parallelism while processing fewer tuples than Full Compute strategies. The iteration logic difference is never entirely negligible, though; the threshold is medium at best ($0.5$ at 15 operations in \cref{fig:snbs_thres_update}). 
\end{itemize}

This analysis shows the importance of Mixed Compute: we should use Selective Compute below the selectivity thresholds, and Full Compute otherwise. The next section details our implementation of Mixed Compute.

\section{Implementing Mixed Compute}
\label{mixedcompute}
The previous section showed that for a given SLDP primitive, there is a threshold below which SelVecManual is optimal and above which BitmapFullManual and FullManual are optimal for Updates and Maps respectively. To find this threshold, we can run a micro-benchmark similar to the one in \cref{fig:snbs_thres}. The lowest selectivity at which the runtime of Full Compute strategies exceeds that of Selective Compute strategies represents the Mixed Compute threshold for a given primitive. In summary, the optimal SLDP strategy is as follows:
\begin{itemize}
\item For Maps: Full or FullManual when the selectivity is above the threshold, and SelVecManual when the selectivity is below the threshold.
\item For Updates: BitmapFull or BitmapFullManual when the selectivity is above the threshold, and SelVecManual when the selectivity is below the threshold. Note that the BitmapFull and SelVecManual have different filter representations.  There is a small cost associated with the conversion from Bitmap to SelVec, but, in practice, other operations always follow an update (e.g., there is a projection after a scan filter). The subsequent operations often amortize the conversion cost. For small queries, in which the amortization does not neutralize the conversion cost, BitmapPartial can slightly outperform SelVecManual because it maintains the Bitmap representation. Nonetheless, as we show in \cref{chapter4}, its gains are mostly negligible.
\end{itemize}

\section{Vectorization Decision Tree}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DecisionTree.png}
    \caption{\textbf{The Summary Decision Tree} -- It summarizes the results in this chapter.}
    \label{fig:filter_decision_tree}
\end{figure}

The decision tree in \cref{fig:filter_decision_tree} summarizes the results of this chapter. It uses the experiments above to derive the optimal implementation strategy depending on the core operation and the input selectivity of a given primitive. The \textbf{Data-Parallel} node, the \textbf{Effecient Data-Parallelism} respectively use the results in \cref{nondataparallel} and \cref{inefficientsimd} that designated SelVecPartial as the optimal strategy. The \textbf{Operation Type} and \textbf{Above Threshold} nodes use the results in \cref{sldpsection} and \cref{mixedcompute} that designed a Mixed Compute strategies for each primitive type (Update or Map).

The tree assumes that the system developer implements all representations and compute strategies. The total number of variants of each primitive will be extensive, which constitutes a costly engineering endeavor. To remedy this cost, we note that the BitmapPartial strategy does not perform significantly worse than SelVecPartial or SelVecManual at most selectivities. Developers can, thus, focus on focus on the Bitmap representation without losing too much performance. Switching between Full and Selective Compute is critical, though; the performance gain is too significant.

\chapter{Experimental Evaluation}
\label{chapter4}

We now evaluate the benefits of the decision tree in \cref{chapter3}. For this evaluation, we implemented the strategies for a subset of queries from the TPCH benchmark, a decision support system workload that simulates an OLAP environment~\cite{tpch}. We use a scale factor of 10 ($\sim$10~GB). The hardware setup is unchanged from the previous chapter. We use the columnar storage engine of the NoisePage DBMS \cite{noise_page}. The string columns are all dictionary compressed for faster processing \cite{abadicompression}. Whenever the compiler (clang \cite{clang}) failed to SIMD vectorize a primitive that should be data-parallel or used AVX2 instead of AVX512, we hand-wrote SIMD code using AVX512 intrinsics \cite{intel2019}.

Each query contains primitive with side-effects (e.g., hash table insertion), and primitives without side-effects (e.g., multiplication of two columns). Given that this thesis focuses on side-effect-free primitives, we report two metrics for each query: the running time of the whole query and the running time of the side-effect free primitives within the query. The output of each query is materialized in memory and printed to the standard output; no network transfer occurs.

\section{Q1: High Selectivity SLDP primitives}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/TPCH-Q1.png}
    \caption{\textbf{The TPCH Q1 query.}}
    \label{fig:tpch_q1}
\end{figure}

\label{q1section}
\begin{figure}[t!]
\centering
\hspace*{\fill}% 
\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/q1_optimized.pdf}
 \caption{Running time of side-effect free primitives.} \label{fig:q1_optimized}
\end{subfigure}
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/q1_bench.pdf}
 \caption{Total running time.}
  \label{fig:q1_bench}
\end{subfigure}%
\caption{\textbf{Q1 Performance}}
\label{fig:q1_perf}
\end{figure}

\cref{fig:tpch_q1} shows the SQL code of the Q1 query. It contains a single SLDP filter in its \texttt{WHERE} clause; its selectivity is consistently above $0.95$ for the vectors from the \texttt{lineitem} table. Its \texttt{SELECT} clause contains multiple arithmetic SLDP operations followed by side-effect-full aggregation and order-by operators that dominate the running time. Before the aggregation, there is an SLDP primitive that hashes the \texttt{GROUP BY} keys. Because the selectivity is high and the query is SLDP-heavy, our decision tree favors SIMD vectorized Full Compute strategies. Besides, switching between strategies is not required.

The results for the strategies we implemented are shown in \cref{fig:q1_perf}. We omit the Mixed strategies' measurements because no switching ever occurs in the query plan; they have the same running times as BitmapFull. \cref{fig:q1_optimized} shows the running time of the primitives we optimized. As predicted by our decision tree, the SIMD Full Compute strategy, BitmapFull, best optimizes these primitives. It outperforms BitmapPartial and SelVecPartial (which have SISD primitives) by $1.6\times$, and SelVecManual (which has manually written SIMD primitives) by $1.3\times$.

The total running times (\cref{fig:q1_bench}) are mostly the same for all queries for two reasons. First, the side-effect-full aggregation, which we do not try to optimize, is the query's dominant component. Second, Intel's implementation of AVX512 instructions decreased CPU speed. This effect explains why clang limits their usage to avoid performance degradations \cite{avx512_registers}. Compare, for example, the BitmapPartial and BitmapFull strategies. They both have the same code for the side-effect-full primitives, but BitmapFull performs them in $838\text{ms}$, whereas BitmapPartial performs them in $739\text{ms}$. Naively avoiding AVX512 instructions is not the solution; AVX2 uses vectors that have half the width of AVX512's, which slows down the execution of SLDP primitives. Careful analysis is thus required to balance the benefits and disadvantages of AVX512 instructions. This analysis is left as future work.

\section{Q6: Mixed Selectivity SLDP Primitives}
\label{q6section}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/TPCH-Q6.png}
    \caption{\textbf{The TPCH Q6 query.}}
    \label{fig:tpch_q6}
\end{figure}

\begin{figure}[t!]
\centering
\hspace*{\fill}% 
\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/q6_optimized.pdf}
 \caption{Running time of side-effect free primitives.}
  \label{fig:q6_optimized}
\end{subfigure}
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/q6_bench.pdf}
 \caption{Total running time.}
  \label{fig:q6_bench}
\end{subfigure}%

\caption{\textbf{Q6 Performance}. The \texttt{+} notation indicates mixed strategies.}
\label{fig:q6_perf}
\end{figure}

\cref{fig:tpch_q6} shows the SQL code of the Q6 query. Its \texttt{WHERE} contains five SLDP filters, and its \texttt{SELECT} one arithmetic SLDP projection followed by an aggregation. For all vectors, the second filter leads to a selectivity below $0.15$, triggering a threshold-based switch. Thus, our decision tree favors a mixed strategy: BitmapFull at high selectivities, and SelVecManual at low selectivities.

The results are shown in \cref{fig:q6_perf}. \cref{fig:q6_optimized} shows the running time of the primitives we optimized. This graph confirms the necessity of the caveat we raised in \cref{mixedcompute}: the cost of converting from Bitmaps to SelVecs makes BitmapFull+SelVecManual slightly more expansive than BitmapFull+BitmapPartial by $0.3\%$, even though SelVecManual performs better than BitmapPartial on the individual primitives in this query. This difference is insignificant, so our decision tree does not take it into account. The BitmapFull+SelVecPartial strategy also suffers from the conversion overhead, but the difference only amounts to $2\%$. All mixed strategies perform better than non-mixed ones. BitmapFull+BitmapPartial and BitmapFull+SelVecManual are faster than BitmapFull by $1.3\times$ because Full Compute becomes wasteful after the selectivity drops below $0.15$. The SIMD SelVecManual strategy is slower by $1.04\times$ because it is only suboptimal in the first two filters of the \texttt{WHERE} clause. The SISD Selective Compute strategies BitmapPartial and SelVec Partial are slower by $1.6\times$ and $1.4\times$ because they do not take advantage SIMD instructions for SLDP operations.

\cref{fig:q6_bench} shows the total running time that includes the side-effect-full aggregation primitive. This primitive only accounts for a small portion of the running time, so we do not observe the slowdown caused by AVX512 registers. In SLDP-dominated queries like Q6, the benefits of AVX512 outweigh its disadvantages.

\section{Q4: Low Selectivity, Inefficient Data Parallelism}
\label{q4section}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/TPCH-Q4.png}
    \caption{\textbf{The TPCH Q4 query.}}
    \label{fig:tpch_q4}
\end{figure}

\begin{figure}[t!]
\centering
\hspace*{\fill}% 
\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/q4_optimized.pdf}
 \caption{Running time of side-effect free primitives.}
  \label{fig:q4_optimized}
\end{subfigure}
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}[t]{.8\linewidth}
 \centering
 \includegraphics[width=0.9\linewidth]{eval/q4_bench.pdf}
 \caption{Total running time.}
  \label{fig:q4_bench}
\end{subfigure}%
\caption{\textbf{Q4 Performance}}
\label{fig:q4_perf}
\end{figure}

\cref{fig:tpch_q4} shows the SQL code of the Q4 query. It is a join of two tables (\texttt{lineitem} and \texttt{orders}) followed an aggregation and an order-by operator. The join's left side (from the \texttt{orders} table) has two SLDP filters that reduce the selectivity to $0.3$ and $0.1$, respectively, and an SLDP hashing primitive before insertion into the join hash table. Its right side (from the \texttt{lineitem} table) has a filter with a selectivity $>0.6$ and a join probe that contains the multiple primitives shown in \cref{fig:probe_example}. Although the probe's hashing primitive is an SLDP operation, its other primitives contain multiple indirect lookups (e.g., accessing hash table entries or the keys within those entries for exact comparison). Because these operations are called in a loop to find all potential matches during the join, they constitute the bulk of the running time. Unfortunately, their data-parallel versions are inefficient. For example, the comparison primitive, which we manually implemented for the SelVecManual strategy, contains three \texttt{gather} instructions: one to collect the keys from the right side of the join, one to collect the hash table entries, and one to collect the keys within the hash table entries for comparison. SIMD instructions can cause performance degradations in these primitives. Thus, we predict that we should use BitmapFull to perform the SLDP filters and hashing, then switch to using SelVecPartial, not the SIMD SelVecManual, for the complex primitives within the join probe.

The results are shown in \cref{fig:q4_perf}. \cref{fig:q4_optimized} shows the running time of the primitives we optimized. Our predicted strategy, BitmapFull+SelVecPartial, is indeed the optimal one, closely followed by BitmapFull+BitmapPartial, which also switches from SIMD to SISD code for complex primitives. Notice the performance degradation caused by SIMD vectorization, as mentioned in \cref{inefficientsimd}. Unlike the previous queries' results, BitmapFull+SelVecPartial performs better than the BitmapFull+SelVecManual by $1.1\times$ because the latter relies on SIMD vectorization for complex primitives. BitmapFull is also slightly worse (by $1.04\times$). Its performance degradation is attenuated by the fact that it requires fewer \texttt{gather} instructions than SelVecManual (e.g., one instead of three in the comparison primitive), relying instead on faster \texttt{vector\_load} instructions to load all elements in a vector, whether they are selected or not. The SISD Selective Compute strategies BitmapPartial and SelVecPartial are, respectively, $1.1\times$ and $1.05\times$ worse because they do not optimize SLDP primitives.

The results in \cref{fig:q4_optimized} mostly reflect those in \cref{fig:q1_optimized}. AVX512 registers, once again, cause a slowdown in SIMD implementations, causing, for example, BitmapFull to become slower than BitmapPartial even though they share the same code for side-effect-full primitives. Here, the disadvantages of AVX512 outweigh its benefits on the full query.

\section{Summary}
Our analysis shows that our results in \cref{chapter3} provide us with the best implementation for each query. The downside is that it requires a wide variety of variants for each primitive because any strategy can be optimal under specific scenarios. We see, however, that Bitmap only strategies (e.g., BitmapFull, or BitmapFull+BitmapPartial) perform nearly as well as the optimal ones in the queries above. Developers can thus discard the SelVec representation to save engineering effort without losing too much performance.

\chapter{Related Works}
\label{chapter5}
\section{Optimizing Analytical Queries for DBMSs}
Besides the Vectorization and Materialization model introduced in \cref{chapter1}, much work has been dedicated to optimizing analytical queries on column store DBMSs. \citeauthor{materializationstrats} studied the effect of materialization strategies (i.e. late or early materialization) on query performance \cite{materializationstrats}. All experiments in this paper use late materialization and rely on filters to keep track of valid tuples. Nonetheless, \citeauthor{materializationstrats} found that the optimal materialization strategy depends on selectivity and on the exact operations performed on tuples \cite{materializationstrats}. \citeauthor{cstore} studied compression as way to optimize both data storage and query execution in column stores \cite{cstore}. \citeauthor{abadicompression} performed an extensive evaluation of various compressions schemes \cite{abadicompression}. The experiments in \cref{chapter4} rely only on dictionary compression.

Researchers have also used SIMD vectorization to optimize relational algorithms. \citeauthor{simd1} optimized scans and nested loop joins using techniques most similar to our BitmapFullManual implementation. \citeauthor{simd2} further applied SIMD vectorization hash tables probes \cite{simd2}. \citeauthor{simdsort1} and \citeauthor{simdsort2} proposed sorting algorithms using SIMD instructions \cite{simdsort1,simdsort2}. The relational order-by operators, sort-merge joins, and sorting aggregations can all use these efficient sorting algorithms.

\citeauthor{ibm_blu} and \citeauthor{simdscan} combined this body of work and built full commercial DBMSs using column stores with compression and SIMD vectorization techniques to optimize table scans \cite{ibm_blu,simdscan}. \citeauthor{orestis_bitmap} further built a system using SIMD vectorization on all relational operators \cite{orestis_bitmap}. These systems make a static filter representation decision: Bitmaps for table scans, and SelVecs for other operations.

Along with the Vectorization Model, the current state-of-the-art execution technique is the Data-Centric Query Compilation technique pioneered by \citeauthor{hyper_llvm} \cite{hyper_llvm}.  This technique splits a given query plan into several pipelines and compiles them individually. A pipeline is a sequence of relational operators that do not materialize their result set followed by a pipeline-breaker -- an operator that does materialize its result set (e.g., to build a hash table or sort tuples). This approach's benefits are: (1) eliminate interpretation overhead by compiling specialized code and (2) maximize cache locality by keeping tuples in the registers or the L1 cache for entire pipelines. Our system uses the relaxed operator fusion technique \cite{rof}, which combines data-centric code generation with the Vectorization Model. It generates code that operates on vectors of tuples, thereby eliminating the interpretation overhead and optimizing cache locality while maintaining all the Vectorization Model's benefits.

\section{Primitive Optimization}
There has been previous work dedicated to optimizing primitives in the Vectorization Model. They mostly differ from this thesis in that they do not consider the impact of filter representation in their performance. We now discuss this work and how they relate to our proposed method.

\citeauthor{sompolski_vec} compared the cost of vectorization to data-centric query compilation \cite{sompolski_vec} and combined the two approaches to generate new vectorized primitive at runtime, which is the method we use in \cref{sldpsection}. Using only the SelVec representation, the authors compare vectorized primitives' performance under various compute strategies (e.g., Full versus Selective Compute, Conditional versus Unconditional Store). They, however, do not provide a way to switch between these strategies.

\citeauthor{everything_vectorized} performs a similar analysis for whole queries in addition to individual primitives \cite{everything_vectorized}. They developed a SIMD implementation of primitives using Selection Vectors, just like our SelVecManual strategy. Like ours, their implementation was effective on compute-heavy queries (e.g., Q6 in \cref{q6section}) and ineffective on memory access-heavy queries (e.g., Q4 in \cref{q4section}).

\citeauthor{miro_adapt} recognized the importance of switching compute strategy at run-time \cite{miro_adapt}. For each operation, the authors implement several \textit{flavors} (i.e., different ways to perform the same operations). They then use reinforcement learning (RL) to switch between the best \textit{flavors} dynamically at runtime. This approach is more general than our micro-benchmarks because it takes into account changes in system load. SelVec is the only representation considered, limiting the range of adaptivity on Update tasks. Besides, our decision tree provides more explainability than RL techniques. As future work, we could employ a similarly dynamic strategy to determine our thresholds.

\chapter{Conclusion}
\label{chapter6}
This work analyzed the impact of filter representation (i.e., Bitmap vs. SelVec) and compute strategy (i.e., Full vs. Selective Compute) on the performance of the vectorized primitives in an in-memory analytical DBMS. We identified the factors that influence performance: number of tuples processed, iteration logic, and core operation time per tuple. We explained how each combination of representation and compute strategy balances between these three factors. Full Compute has the cheapest iteration logic, processes all tuples, but spends less time on each tuple when SIMD vectorization is possible. Full Compute is, however, only available with Bitmaps on Update primitives. Selective Compute with SelVecs has a cheaper iteration logic than Selective Compute with Bitmaps, and is more amenable to SIMD vectorization. We confirmed these observations with several micro-benchmarks, and created a decision tree to help systems developers find the optimal implementation of arbitrary side-effect-free primitives. Finally, we showcased the benefits of our analysis on OLAP queries with multiple primitives and consistently achieved the best performance. Our performance gains over the best techniques that do not adapt filter representation and compute strategy can be up to $1.3\times$.

%\appendix
%\include{appendix}

\backmatter

%\renewcommand{\baselinestretch}{1.0}\normalsize

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}}
%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{plainnat}
\bibliography{citations} %your bib file

\end{document}

### 2021-07-22 old1.tex
Updates from Overleaf

@@
 \title{ %% {\it \huge Thesis Proposal}\\
-{\bf Filter Representation in Vectorized Query Execution}}
-\author{Amadou Latyr Ngom}
+{\bf Log Replay for Code Generation in OLTP Systems}}
+\author{Tianlei Pan}
 \date{June}
-\Year{2020}
+\Year{2021}
 \trnumber{}
@@
 Andy Pavlo, Chair \\
-Todd C. Mowry \\
+Wenting Ye \\
 }
@@
 
-\keywords{Vectorization, Compilation, Filter, SIMD}
+\keywords{Code Generation, Log Replay, Recovery, Query Compilation}
 
@@
 \begin{abstract}
-In-memory Database Management Systems (DBMSs) use vectorization to reduce interpretation overhead, improve cache locality, and perform loop optimizations.
-For each \textit{vector} (i.e.
-batch of tuples), the DBMS needs to track the set of valid tuples after a predicate application.
-Systems employ one of two representations: Selection Vectors (SelVecs) or Bitmaps.
-In this work, we analyze each approach's strengths and weaknesses to provide recommendations on how to implement vectorized operations.
-We determine, through a wide range of micro-benchmarks, that the optimal implementation strategy is a function of many factors: the cost of the iterating through tuples, the cost of the operation itself and how amenable it is to SIMD vectorization.
-Our analysis shows that Bitmaps perform better for operations that can be vectorized using SIMD instructions but that SelVecs otherwise perform better due to a cheaper iteration logic.
+Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records.
+We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
 \end{abstract}
@@
 \begin{acknowledgments}
-My advisor is cool.
+yes
 \end{acknowledgments}
@@
 
-\chapter{Introduction and Background}
-DBMSs typically handle two broad classes of applications: Online Transaction Processing (OLTP) and Online Analytical Processing (OLAP) applications \cite{abadi_phd}.
-OLTP applications tend to be short-running, write-heavy queries that only touch a few entries in a database.
-OLAP applications, which are the focus of this thesis, are long-running, read-mostly queries that analyze large amounts of data.
-Historically, this data was disk-resident; disk accesses were the bottleneck of OLAP queries.
-As DRAM sizes increase, and the cost per GB decreases, large databases begin to fit in main memory.
-Rather than optimize for disk access, modern DBMSs must optimize for CPU efficiency by reducing instruction count and cycles per instruction when executing queries.
+\chapter{Introduction}
+A \textit{database management system (DBMS)} is a software that is responsible for storing data, analyzing data, and interacting with applications.
+Depending on the use case, a DBMS can be designed to focus more on either capturing data, or analyzing data.
+\textit{On-line Analytical Processing (OLAP)} applications focus on reading, analyzing and aggregating cold data that is less likely to be modified.
+On the other hand, \textit{On-line Transaction Processing (OLTP)} applications work with write-heavy transactions that modify the database frequently.
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/SampleQuery.png}
-    \caption{\textbf{Sample SQL Query and Plan.}}
-    \label{fig:intro_query}
-\end{figure}
+Consequently, recovery and durability in OLTP applications are more of a pressing issue compared to OLAP systems.
+Naive or manual backups are usually sufficient for read-only OLAP applications.
+However, since OLTP databases are being updated by transactions constantly, naive backups will either cause the database to lose significant amount of data, or fail to preserve the \textit{ACID (atomicity, consistency, isolation, and durability)} properties of a DBMS.
+This requires OLTP applications to develop specialized recovery mechanism to handle these problems.
+\textit{Write-ahead logging (WAL)} is currently the most widely adopted recovery mechanisms for OLTP applications that preserves ACID properties.
 
-The traditional execution technique used in most DBMSs \cite{mysql, posrgres}, the Iterator Model \cite{volcano}, became ill-suited for modern OLAP applications.
-It executes a query by interpreting the query plan.
-Each database operator implements a virtual \texttt{next()} function that returns the next tuple it can generate; each expression (i.e.
-addition, multiplication) also provides an virtual \texttt{evaluate()} function that takes in a tuple and returns a value.
-\cref{fig:volcano_graph} showcases the Iterator model with the sample query in \cref{fig:intro_query}.
-The operators \texttt{Scan A} and \texttt{Scan B} iterate through the table and emit the tuples to the parent operator.
-The filter only returns tuples that satisfy the \texttt{WHERE} clause.
-The join first builds a hash table, then probes it to find matches.
+In addition, OLTP systems also require specialized algorithms and implementations to execute write transactions.
+Query compilation is an important optimization technique used in OLTP systems to greatly reduce the number of instructions that need to be executed.
+A DBMS uses a processing model (e.g.
+Iterator, Vectorization, Materialization) to determine how the DBMS executes a query plan.
+The processing model is responsible for translating a query into an intermediate expression.
+The execution engine then breaks down the expression into various tasks for execution.
+To speed up this process, the DBMS can use an optimization technique known as \textit{Query Compilation}.
+Query compilation requires the DBMS to generate specialized code for execution tasks to increase execution throughput.
+These specialized code are often created using an off-shelf compiler into a C/C++ program that implements the functionality of a certain query.
+For instance, HyPer \cite{hyper_llvm} uses the LLVM compiler framework to translate queries into machine code.
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/Volcano.png}
-    \caption{\textbf{The Iterator Model on the Sample SQL Query.}}
-    \label{fig:volcano_graph}
-\end{figure}
+While query compilation has proved successful in increasing the efficiency of a DBMS, it comes with several drawbacks.
+Implementing a query compilation system can be quite an ordeal.
+It requires additional knowledge of compiler systems (e.g.
+LLVM) and huge amounts of engineering work to translate different execution tasks.
+Moreover, the DBMS also need to modify other components to accommodate with an execution engine reliant on query compilation.
 
-We can see why the Volcano model is ill-suited for in-memory DBMSs.
-For every tuple -- of which there can be billions in an OLAP setting -- the virtual function calls (marked in red in the figure) constitute a significant overhead.
+One of such components is the recovery system.
+The structure of log records are vastly different from query plans.
+Therefore, to implement a recovery system in a query compilation environment, the DBMS would need to implement an additional execution engine to interpret the log records.
+Instead, we propose a system design that will extend the query execution engine to create custom programs for replaying log records.
 
-To minimize this overhead for in-memory DBMSs, researchers have developed the Vectorization Model.
+\chapter{Background}
+\section{Log Replay}
+Database recovery systems are responsible for preserving the ACID principles of a DBMS in case of failure.
+Early recovery systems assume that there is no non-volatile memory.
+Mention Tandem here ($https://link.springer.com/chapter/10.1007/3-540-51085-0_43$).
+Modern DBMS can utilize a much larger memory.
+This allows logging protocols to be simplified (e.g.
+tracking dirty pages is no longer required).
+A modern recovery system consists of two parts: logging and recovery.
 
-\iffalse
-\section{Compilation}
+Logging is the action of storing transactional information on disk.
+These information are stored in a special data structure called a \textit{log record}.
+A log record can either contain physical information (e.g.
+changes made to a specific physical address) or higher-level information (e.g.
+user-input query).
+Mention MSSQL (time-travel table).
+
+Recovery is the action of restoring the database system into a consistent state.
+If available, the DBMS first attempts to restore its previously recorded checkpoint.
+The database then initiates the log replay process that goes through every log record stored since that checkpoint.
+SiLO (checkpoint).
+
+Describe how log replay works in Noisepage. (Big Picture)
+Logging: Transactions are logged prior to commit.
+Recovery: Independent recovery thread that loops over all the log records.
+The log record is either a redo record or a delete record.
+The contents of the record are then interpreted into either a staged write or delete.
+Initialization of catalog is hard-coded.
+
+\section{Query Compilation}
+part for Talk about query compilation / code generation. Mention Efficiently Compiling Efficient Query Plans
 \begin{figure}[H]
@@
 
-Query compilation \cite{system_r, hique, hyper_llvm} aims to eliminate the cost of interpretation by compiling the query plan down to machine code.
-{
-\color{red}
-Describe what compilation means.
-}
+Query compilation aims to eliminate the cost of interpretation by compiling the query plan down to machine code.
 
+Talk about related work. (Revise this section).
 The data-centric compilation technique pioneered by Hyper \cite{hyper_llvm} splits the plan into several pipelines and compiles them individually.
@@
 Compilation thus reduces both the number of instructions executed and the number cycles per instructions.
-\fi
 
-\section{Vectorization}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/Vectorization.png}
-    \caption{\textbf{The Vectorization Model with the Sample Query} -- The code is nearly identical to that in \cref{fig:volcano_graph}}
-    \label{fig:vectorization_graph}
-\end{figure}
+Describe how query compilation works in Noisepage. (Big Picture).
 
-The vectorization model, pioneered by Vectorwise \cite{vectorwise}, is almost identical to the Volcano model, except that it deals with batches of tuples or \textit{vectors} (e.g., 2048 tuples at a time).
-The \texttt{next()} function returns a vector of tuples, and the \texttt{evaluate()} function returns a vector of values.
-\cref{fig:vectorization_graph} showcases vectorization model on the sample query in \cref{fig:intro_query}; it is near identical to \cref{fig:volcano_graph}.
+\chapter{Method}
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/VectorMul.png}
-    \caption{\textbf{A Vectorized Multiplication Function} -- The code in green indicates compiler or programmer optimizations to exploit Intel's AVX512 features.}
-    \label{fig:vector_mul}
-\end{figure}
-Vectorization provides two main benefits.
-First, it amortizes the cost of interpretation by the size of vectors.
-Second, most functions that operate on vectors are highly amenable to loop optimizations (e.g., loop unrolling, SIMD vectorization) by the compiler or by the programmer.
-Consider, for a example, the function shown in \cref{fig:vector_mul}.
-The iterations of the loop are independent, and the floating-point multiplication possesses a SIMD variant, which means the loop can be unrolled and vectorized by the compiler.
+Our proposal is to extend the recovery layer and query compilation layer to support the interpretation of log records. This approach can be realized in three steps:
+\begin{enumerate}
+\item Extract information from log records.
+\item Convert the information into a format that can be accepted and compiled by the code generation engine.
+\item Execute the compiled code.
+\end{enumerate}
 
-\iffalse
-\begin{figure}[H]
-    \centering
-    \includegraphics[scale=0.5]{images/VectorCompilation.png}
-    \caption{A hybrid vectorization and compilation model with the sample query..}
-    \label{fig:mix_compilation_vectorization}
-\end{figure}
-Our system \cite{rof, noise_page} combines both data-centric code generation and vectorization : it generates code that operates on vectors.
-This approach eliminates interpretation overhead and still enables loop optimizations.
-\cref{fig:mix_compilation_vectorization} showcases the mixed model.
-It differs from \cref{fig:compilation_graph} only in that it operates on vectors.
-\fi
+\subsection{Log Record}
+The log record structure in Noisepage follows closely to the typical format of a write-ahead log.
+The header of a log record stores metadata about the log record itself (e.g.
+record type, record size).
+The header is then followed by a byte buffer that stores the persisted contents.
+The size of the byte buffer is also aligned to 8 bytes to ensure the entire log record is aligned to 8 bytes.
 
-Vectorization introduces a new challenge.
-Individual tuples within a vector may be \textit{filtered out}, meaning that should not be processed.
-Only \textit{selected} tuples -- those not filtered out -- should be processed.
-For example, in \cref{fig:vectorization_graph}, only tuples that satisfy the \texttt{WHERE} clause should be inserted in the hash table.
-Next, we discuss the methods used to represent filters, i.e., to identify selected tuples within a vector.
+During a step of the conventional recovery process, the recovery manager only needs to check the type of the log record and proceed with an insert, update or delete operation.
+For our method, besides the log record type, we need to find out the contents of the log record as well.
+This requires additional access to the database catalog.
 
-\section{Filter Representation}
-\cref{fig:repr_intro} displays the two common ways to represent filters: Selection Vectors (SelVecs) and Bitmaps.
-SelVecs index into the selected tuples; tuples without an index are implicitly filtered out.
-For Bitmaps, on the other hand, the set bits indicate the selected tuples; the unset bits indicated filtered out ones.
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.7]{images/FilterRepresentationIntro.png}
-    \caption{\textbf{Filter Representation} -- Selection Vectors (left) index into the tuple vector. Bitmaps (right) are only set on selected rows.}
-    \label{fig:repr_intro}
-\end{figure}
-Previous work has assumed one of these two representations, without providing a reason for that choice.
-Vectorwise \cite{vectorwise}, and work derived from it \cite{miro_adapt, everything_vectorized, sompolski_vec},  rely on SelVecs, whereas the more recent VIP \cite{orestis_bitmap} relies on Bitmaps.
-This paper aims to analyze the trade-offs between these two representations and derive insights for developers implementing vectorized query processing.
+[Should I expand on the specific types here or later???]
+\subsection{Compilation Pipeline}
+In Noisepage, a query is first converted into a plan node.
+The execution engine then translates the plan node into Table Producing Language (TPL) code via a translator object.
+Finally, the TPL code is compiled into native C++ code.
+[Also need to explain execution context].
 
-\chapter{Computing on Filtered Vectors}
-\section{Operations on Filtered Vectors}
-\begin{figure}[t!]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/UpdateIntro.png}
- \caption{Update operation.}
-  \label{fig:update_intro}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/MapIntro.png}
- \caption{Map operation.}
-  \label{fig:map_intro}
-\end{subfigure}
-\caption{\textbf{Operations on Filtered Vectors} -- Updates change the set of selected tuple. Maps apply a function to the set of selected tuples without updating the set.}
-\label{fig:update_map_intro}
-\end{figure}
-\cref{fig:update_map_intro} shows the two kinds of operations on vectors of tuples: Updates and Maps.
-Updates apply a new predicate function to the vector and update the selected tuples accordingly.
-A scan filter (i.e., WHERE clause) is an example of an update operation.
-Maps, on the other hand, keep the selected set but compute a new vector using a mapping function, or reduce the vector to a single value (e.g., for aggregation).
-A projection (i.e., SELECT clause) is an example of a map operation.
-All relational operations considered in this paper are decomposable into a set of maps and updates.
-\cref{fig:probe_example} shows the example of join probes.
-First, we compute the hashes of the input tuples; the result is a vector of hash values (Map).
-Then, we perform a hash table lookup using the hashes to obtain a vector of hash table entries (Map).
-Due to hashing collisions, the entries may not match the input, so we need to compare the join keys to filter out inexact matches (Update).
-We then advance the hash table chains (Map) and repeat the comparison until entries are exhausted.
+The structure of a log record is different from a query in a number of ways.
+Firstly, the log record only stores physical information of a committed transaction.
+For example, an insert query requires the execution engine to retrieve a new tuple slot.
+For an insert log record, the tuple slot is already been determined and stored within the log record.
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/JoinProbeExample.png}
-    \caption{\textbf{Multi-step Join Probe} -- Bold squares indicate the hash table entries currently pointed to.}
-    \label{fig:probe_example}
-\end{figure}
+These differences require subtle modifications to the compilation process.
+Describe Plannodes, Translators that are required for Query Compilation. Can use a graph here.
 
-Other relational operators are similarly decomposable. The next section briefly discusses transitioning between steps before we move on to the core of the thesis: optimizing each step.
+Describe Execution Context, VM, Bytecode Emitter/Handlers. Interpret vs Compiled.
 
-\section{Transitioning Between Steps}
-In this section, we briefly talk about the cost of transitioning between steps.
-Most of the time, this cost is non-existent because the filter: the transition is \textit{implicit}.
-The hashing phase and the lookup phase in \cref{fig:probe_example} have the same selection vector.
-On the other hand, the Chain Following phase operates on entries that do not satisfy the comparison: the transition involves a set difference.
-Transitions in decomposed disjunctive filters \cite{pcq} (shown in \cref{fig:disjunctive_filter_howto}) involve both a set difference for short-circuiting and a union to get the final result.
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/DisjunctionHowto.png}
-    \caption{\textbf{Transitioning Between Steps} -- We apply the first clause on all input tuples.
-In the second step, however, we only apply the second clause on those tuples that don't pass the first clause.
-In the third step, we take the union between the two outputs.}
-    \label{fig:disjunctive_filter_howto}
-\end{figure}
-To the best of our knowledge, all transitions take the form of set differences (for elements that do not pass the previous predicate) and unions (to obtain the final result).
-These operations are cheap to implement with Bitmaps because we can easily rely on SIMD instructions.
-Efficiently implementing set operations on SelVecs is more complicated.
-Whenever necessary, Updates can write to one additional SelVec for tuples that do not pass the filter; this second SelVec implicitly corresponds to the result of a set difference.
-Note that it does not overlap with the first SelVec, so a union of any of their subsets is just an array append, a fast operation.
-\cref{fig:disjunctive_filter_perf} shows that the overhead of set operations is negligible for both bitmaps and selection vectors.
-The rest of the paper, therefore, focuses on optimizing the individual steps.
+\subsection{Reinterpreting Log Records}
+Describe the process of extracting contents from the log records and converting them into plannodes, translators.
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{eval/transition.pdf}
-    \caption{\textbf{Transition Overheads} -- The overheads are $3.55\%$ and $4.06\%$ for Bitmaps and SelVecs respectively.}
-    \label{fig:disjunctive_filter_perf}
-\end{figure}
+For all records: Find out the mapping from physical column ids to logical column ids.
 
-\section{Compute Strategies}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/ComputeIntro.png}
-    \caption{\textbf{Compute Strategies} -- Selective Compute (left) only considers selected tuples. An \texttt{x} indicates an uninitialized value. Full Compute (right) considers all tuples.}
-    \label{fig:compute_intro}
-\end{figure}
- A compute strategy determines how to perform the Update or Map operations.
-We consider three compute strategies: Selective, Full, and Mixed Compute.
-As their names indicate, Selective Compute only applies the given functions to selected tuples, whereas Full Compute applies them to all tuples.
-\cref{fig:compute_intro} shows each approach's effect.
-While Full Compute usually performs more work, it greatly benefits from SIMD vectorization, simple loop structure (for loop unrolling and interleaving), and easy branch prediction.
-To exploit this trade-off, the third strategy, namely Mixed Strategy, switches from Full Compute to Selective Compute when the selectivity (i.e., the ratio of selected tuples) goes below a certain threshold.
-The next sections will show how to obtain this threshold.
+\subsubsection{Insert Redo Record}
+Iterate over the columns of a redo record.
+Find the value types for each column.
+Get the value using the previously found value types and column id mappings.
+Transform each value into an expression value type (ConstantValueExpression/ParameterValueExpression).
+All the values and metadata are then passed into an Insert Plannode, which is then compiled into an executable query.
+The tuple slot of the insert operation is retrieved from the execution context, which is then passed to the recovery manager.
+\subsubsection{Update Redo Record}
+TBD. Should be similar to insert/delete.
+\subsubsection{Delete Record}
+How to delete index? No information is given on the index, except for the tuple slot to be deleted.
 
-Note that for Update operations, Full Compute is not nicely compatible with SelVecs because a slow intersection of sorted sets \cite{sorted_set} would be required to compute the final selection.
-This limitation will become important when choosing a strategy to implement.
-Furthermore, operations with side-effects (e.g., hash table build, aggregation) cannot use Full Compute for correctness reason.
-This thesis focuses on side-effect free operations.
+Unless we store a Delta in the delete record, there is no way to get the values.
 
-Furthermore, for SelVecs, there is another dimension we did not consider: Branching Store versus Non-Branching Store.
-We do not focus on this dimension, for it has been studied extensively before \cite{miro_adapt}.
-All experiments below use the optimal branching strategy.
+Pass the delete tuple slot into the execution context, which can be processed in the Bytecode Handler to force the delete operation on the designated tuple slot.
 
-\section{The Filter Decision Tree}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/DecisionTree.png}
-    \caption{\textbf{The Filter Decision Tree.}}
-    \label{fig:filter_decision_tree}
-\end{figure}
-\cref{fig:filter_decision_tree} shows the filter decision tree.
-For every kind of operation on filtered tuples, the tree indicates what the best representation and compute strategy pair is.
-It assumes that the developer implements all representations and compute strategies, which is a fairly costly endeavor engineering-wise.
-When this is not affordable, systems designers can still use this tree as a starting point before making compromises.
-The remaining part of this section will give a high-level rationale behind the tree.
-Chapter 4 will provide practical case studies that justify the shape of this tree.
-Chapter 5 will use the insights provided by this tree to execute an end-to-end multi-step query and show the performance gains.
+\section{Caching Optimization}
+Caching Identifier: Database Oid + Table Oid.
+\subsection{Metadata}
+Metadata can be cached. Acquire metadata from Catalog. Contents need to be invalidated on drop table?
+\subsubsection{Physical to Logical Column Mapping}
+Required since the recovery protocol is physical, while query compilation uses logical ids.
+\subsubsection{Column Value Types}
+This is only for Redo Records. This is required to memcpy the correct values out of a log record into a plan node.
+\subsection{Compiled Query}
+The compiled query can be cached. Mention that there are two execution modes: Interpret vs Compiled.
 
-\section{High Level Rationale}
-\begin{figure}[t!]
-\centering
-\hspace*{\fill}%
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RationaleFull.png}
- \caption{Full Compute.}
-  \label{fig:rationale_full}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
+For Insert Redo Records: Use ParameterValueExpressions for insert plannodes. Then after compilation, swap in the values to insert into the compiled query into the parameter slots.
 
-\hspace*{\fill}%  
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RationaleBitmap.png}
- \caption{Selective Compute with a Bitmap. The iteration logic is adapted from \cite{bitmap_iteration}. The SIMD version is equivalent to Full Compute.}
-  \label{fig:rationale_bitmap}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
+For Update Redo Records: ???
 
-\hspace*{\fill}%  
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RationaleSelVec.png}
- \caption{Selective Compute with a SelVec.
-The last \texttt{scatter} is only necessary to maintain consistent indices between the input and output vector during multi-step Map operations.
-It should otherwise be avoided due to its slowness.}
-  \label{fig:rationale_selvec}
-\end{subfigure}
-\hspace*{\fill}%
-\vspace*{8pt}%
+For Delete Records: Tuple slot is carried by Execution Context, so not much needs to be done here.
+\chapter{Experimental Evaluation}
+Mention how the benchmark is implemented.
+\section{Throughput Measurement}
+Use proper names. Machine specifications. FIO benchmarks. Run experiments with different perspectives.
 
-\caption{\textbf{Implementing Vector Primitives} -- The code in green indicates the core operation (multiplication). The code in red indicates iteration logic.}
-\label{fig:rationale_code}
-\end{figure}
+1. Standard Insert Workload. Single database, single table. 100 percent inserts. Max Columns of 5. 5 statement per txn. Transaction length 5. Allows Varlen.
 
-Every operation on filtered tuples is a loop, the body of which performs the actual computation.
-Three factors influence this loop's performance: the iteration logic, number of processed tuples, and time per processed tuple.
-\cref{fig:rationale_code} shows the three ways to multiply a filtered vector by a constant (a Map operation).
-Full Compute (e.g., \cref{fig:rationale_full}), regardless of representation, has the cheapest iteration logic because it does not need to differentiate selected indices.
-For Selective Compute, iterating over a Bitmap (e.g., \cref{fig:rationale_bitmap}) is more expansive than iterating over a SelVec (e.g., \cref{fig:rationale_selvec}) because of the bit operations required.
-On the other hand, Full Compute processes more tuples than Selective Compute, especially when the selectivity is low.
-Finally, SIMD instructions usually reduce the number of iterations and the number of cycles per processed tuples.
-SIMD code is cheapest with Full Compute; Selective Compute requires \texttt{gather} and possibly \texttt{scatter} to differentiate selected tuples.
+2. HighStress Insert Workload. Single database, single table. 100 percent inserts. 5 statement per txn. Max Column of 1. Does not allow Varlen.
 
-We can use these these three factors to explain the shape of the decision tree.
+\chapter{Related Work}
+1.
+A Recovery Algorithm for A High-Performance Memory-Resident Database System.
+An early paper for disk-based DBMS recovery.
+Proper log replay infrastructure and checkpoint recoveries.
+https://15721.courses.cs.cmu.edu/spring2018/papers/12-logging/p104-lehman.pdf
 
-\textbf{The root node}: The root of the tree addresses the case of operations on variable-length data, as is often the case for string operations (e.g., string comparison, sub-string).
-The number of cycles dedicated to each tuple usually far dominates the total cost.
-In this scenario, Full Compute is highly inefficient because it processes more tuples than Selective Compute.
-Also, the cost iteration logic is negligible, making representation irrelevant.
-We, therefore, choose to use Selective Compute, with either SelVecs or Bitmaps.
+2.
+Different Logging Schemes.
+Physical Logging vs Logical Logging.
+Noisepage uses physical logging.
+Examples: SiloR (https://15721.courses.cs.cmu.edu/spring2020/papers/10-recovery/zheng-osdi14.pdf).
+It also uses checkpoints, which does not exist in Noisepage.
 
-\textbf{The branching node}: We then move on to the Branching node, which addresses operations with branches in them (e.g., logical operations, \texttt{CASE} statements, \texttt{IN} statements).
-Branching increases the cost of SIMD relative to SISD: SISD skips unnecessary instructions, SIMD executes all of them using bit-masks.
-Furthermore, branches within a loop complicate loop optimizations (e.g., unrolling, interleaving).
-As a result, Selective Compute beats Full Compute.
-SelVecs only win over Bitmaps because of the simplicity of their iteration logic.
+3.
+Different Recovery Schemes.
+Time-travel table (Azure SQL Database https://15721.courses.cs.cmu.edu/spring2020/papers/10-recovery/p2143-antonopoulos.pdf).
+Os-fork Snapshots (Hyper, https://dl.acm.org/doi/10.1109/ICDE.2011.5767867)
 
-\textbf{The non-simdable node}: We now consider the non-simdable node.
-The most (only?) relevant operations that qualify are integer division and modulo.
-Complex pointer manipulations, while technically simdable using expensive scatters and gathers, also fall under this category.
-No compiler will auto-vectorize them, and few programmers will write and maintain such SIMD code.
-Like with variable-length instructions, Full Compute would introduce too much extra computation, and the representation does not matter.
-
-\textbf{The SNBS nodes}: The most interesting nodes of this tree are those dedicated to scalar non-branching simdable (SNBS) operations; all three factors simultaneously play an essential role.
-Besides, for Updates, Full Compute is incompatible with SelVecs; it may be necessary to switch both compute strategy and filter representation depending on the task.
-The next section will justify the claims in these nodes and provide an efficient way to switch compute strategy and representation.
-
-\chapter{Experimental Evaluation}
-In this section, we experimentally justify the shape of the tree.
-In doing so, we will also provide a general technique that systems can use optimally switch between strategies.
-\cref{tab:strategies} summarizes all the strategies we implemented.
-We run all experiments on an Intel Xeon Platinum 8124M CPU @ 3.00GHz with AVX512 support.
-
-\begin{table}[t!]
-\centering
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-\textbf{Strategy Name} & \textbf{Representation} & \textbf{Compute} & \textbf{SIMD}      & \textbf{Compatibility} \\ \hline
-SelVecPartial         & Selection Vectors       & Selective        & None               & Update and Map         \\ \hline
-SelVecManual          & Selection Vectors       & Selective        & Manually Written   & Update and Map         \\ \hline
-BitmapPartial         & Bitmaps                 & Selective        & None               & Update and Map         \\ \hline
-BitmapFull            & Bitmaps                 & Full             & Auto Vectorization & Update and Map         \\ \hline
-BitmapFullManual      & Bitmaps                 & Full             & Manually Written   & Update and Map         \\ \hline
-Full                  & Either                  & Full             & Auto Vectorization & Map                    \\ \hline
-FullManual            & Either                  & Full             & Manully Written    & Map                    \\ \hline
-\end{tabular}
-\caption{\textbf{Implemented Strategies} -- For the SIMD columns, `Auto Vectorization' indicates that we rely on compiler auto vectorization, whereas `Manually written' indicates that we write SIMD code ourselves using intrinsics.
-On the compatibility column, Update operations cannot use both Full Compute and Selection Vectors, which explains the last two entries.}
-\label{tab:strategies}
-\end{table}
-
-\section{Variable Length Data}
-\begin{figure}[t!]
-\captionsetup[subfigure]{justification=justified}
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_string_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE country < `Japan'}}
-  \label{fig:varlen_update}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_string_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT POSITION(`a' in country)}.}
-  \label{fig:varlen_map}
-\end{subfigure}
-\caption{\textbf{Variable Length Data} -- Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
-\label{fig:varlen_map_update}
-\end{figure}
-
-In this section, we analyze the performance profile of operations on variable-length data.
-We choose cheap Update and Map operations -- string comparison, and character location on short strings (country names) -- to show that time spent per tuple typically dominates the cost of string operations.
-Here, the primitives process one tuple at a time; it is impossible to use SIMD registers on a vector with variable-length data.
-
-The profiles are shown in \cref{fig:varlen_map_update}.
-The results confirm the high-level intuition described in the previous section.
-The string operations themselves are so expensive that representation matters little, and full compute is not viable.
-As a result, we decide to use that Selective Compute with either representation.
-
-\section{Branching Operations}
-\begin{figure}[t!]
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_and_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \&\& col2 < val2}}.
-  \label{fig:logical_and_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_and_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \&\& col2 < val2}.}
-  \label{fig:logical_and_map}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_or_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 || col2 < val2}}.
-  \label{fig:logical_or_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_or_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 || col2 < val2}.}
-  \label{fig:logical_or_map}
-\end{subfigure}
-
-\caption{\textbf{Branching Operations} -- Each operation contains one branch because of boolean short-circuiting.}
-\label{fig:logical_map_update}
-\end{figure}
-
-\begin{figure}[t!]
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_and_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \& col2 < val2}}.
-  \label{fig:bitwise_and_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_and_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \& col2 < val2}.}
-  \label{fig:bitwise_and_map}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_or_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 | col2 < val2}}.
-  \label{fig:bitwise_or_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_or_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 | col2 < val2}.}
-  \label{fig:bitwise_or_map}
-\end{subfigure}
-
-\caption{\textbf{Bitwise Operations} -- The increase in performance is due to the lack of branching.}
-\label{fig:bitwise_map_update}
-\end{figure}
-
-We now consider operations involving branching execution paths.
-We choose the logical or/and operations because boolean short-circuiting involves a single branch to avoid evaluating unnecessary conditions.
-To isolate the effect of branching, we also evaluate branch-less bitwise or/and operations.
-We have confirmed that the compiler manages to auto-vectorize the operations.
-
-The results for logical operations are shown in \cref{fig:logical_map_update}.
-We see that PartialSelVec performs better thanks to a cheaper iteration logic and that Full Compute is not viable because SIMD does not interact well with branching.
-\cref{fig:bitwise_map_update} confirms that branching is truly responsible for the difference in performance.
-The only difference between bitwise and logical operations is the short-circuiting that takes place.
-While Full Compute can sometimes be competitive in these scenarios (see \cref{fig:logical_or_map} at high selectivities), we decide, for the sake of simplicity, to use Selective Compute with SelVecs.
-
-\section{Non Simdable Operations: Integer Division}
-\begin{figure}[t!]
-\captionsetup[subfigure]{justification=justified}
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_division_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 \% col2 < val}}.
-  \label{fig:division_update}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_division_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 / col2}.}
-  \label{fig:division_map}
-\end{subfigure}
-\caption{Simple operations on a variable-length column. Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
-\label{fig:division_map_update}
-\end{figure}
-
-For this section, we consider operations involving integer division. There is no SIMD instruction for integer division; the primitives process one tuple at a time.
-
-The results are shown in \cref{fig:division_map_update}.
-As predicted in the high-level rationale, the cost of integer division is so high that we recommend Selective Compute with either representation.
-\section{SNBS Operations}
-\label{snbs_section}
-\begin{figure}[t!]
-    \centering
-    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_update.pdf}}
-    \caption{\textbf{SNBS Update Operations} -- For Updates, SelVecManual relies on the \texttt{compress\_store} SIMD instruction rather than on \texttt{scatter}, which reduces its overhead.}
-    \label{fig:snbs_update}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_map.pdf}}
-    \caption{\textbf{SNBS Map Operations} -- For Maps, SelVecManual relies on the \texttt{scatter} SIMD instruction, which explains why it initially does not perform better than SelVecPartial.}
-    \label{fig:snbs_map}
-\end{figure}
-
-As mentioned in the rationale, the performance of SNBS operations will depend on all three performance factors: iteration logic, number of tuples processed, time per tuple.
-To isolate each component's effect, we experimentally evaluate the running time of Maps and Updates as a function of the selectivity and the number of single-cycle instructions (e.g., addition, subtraction, bitwise and, or, xor) in each operation.
-
-The experimental profile of operations with multi-cycle instructions (e.g., multiplication, load, store) can be approximated by treating such operations as multiple single-cycle.
-The multiplying factor can be found in various instruction tables \cite{intel2019}.
-For example, in the machine we use for testing, a Map operation with a multiplication should have a profile similar to that of a Map operation with three additions.
-This method is likely not optimal, though; finding a better one is left as future work.
-
-\cref{fig:snbs_update} and \cref{fig:snbs_map} show the performance profile of our different implementations as the number of single-cycle operations varies.
-The profiles correspond to what we expect based on the three performance factors, but a detailed explanation of the main features in order.
-\begin{itemize}
-    \item SelVecPartial generally performs better than BitmapPartial.
-We attribute this fact to the cost of the iteration logic.
-This difference is particularly relevant for cheap operations, but become irrelevant as complexity increases.
-    \item Full Compute is always worst when the selectivity is low; it processes far more tuples than Selective Compute. 
-    \item Full Compute is always best when the selectivity is high thanks to cheaper iteration logic and cheaper SIMD code.
-For example, SelVecManual always needs a gather instruction, whereas BitmapFullManual does not.
-    \item SIMD beats SISD because of less time spent per tuple, as exemplified by the difference between SelVecPartial and SelVecManual.
-    \item Manual Vectorization (with FullManual and BitmapFullManual) sometimes performs much better than Auto-Vectorization (with Full and BitmapFull).
-With further investigation of the generated code, we discovered that the compiler is overly conservative when it comes to using AVX512.
-AVX512 registers result in decreased CPU frequency \cite{avx512_registers}, so the compiler is careful not always to use them.
-It often uses AVX2 registers instead.
-We, on the other hand, always use AVX512.
-\end{itemize}
-
-\section{Implementing Mixed Compute}
-\begin{figure}[t!]
-\captionsetup[subfigure]{justification=justified}
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/snbs_thres_update.pdf}
- \caption{Update thresholds between BitmapFullManual and Selective Compute strategies.}.
-  \label{fig:snbs_thres_update}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/snbs_thres_map.pdf}
- \caption{Map thresholds between FullManual and Selective Compute strategies.}
-  \label{fig:snbs_thres_map}
-\end{subfigure}
-\caption{\textbf{Mixed Compute Thresholds} -- The Full Compute strategies are BitmapFullManual and FullManual because compiler auto-vectorization does not consistently use AVX512 registers (as explained in the previous section).}
-\label{fig:snbs_thres}
-\end{figure}
-
-Both \cref{fig:snbs_update} and \cref{fig:snbs_map} explain the necessity of Mixed Compute.
-Depending on the operation, there is a selectivity threshold at which we should switch from Full Compute to Selective Compute.
-As \cref{fig:snbs_thres} shows, this threshold a function of the cost of the operation.
-The curves for BitmapPartial and SelVecPartial are relatively flat because they lack the SIMD capabilities.
-For example, Full Compute dedicates fewer cycles per tuple even as it processes around five times more tuples at a selectivity of $0.2$.
-As for SelVecManual, the \texttt{gather} instruction (show in \cref{fig:rationale_selvec}) dominates its cost when the number of operations is low.
-As the number of operations increases, this \texttt{gather} plays a less important role in the overall performance; SelVecManual becomes more and more competitive with Full Compute because they both utilize SIMD instructions.
-The threshold, therefore, continually increases.
-
-For a new query, the optimal thresholds can either be obtained by running micro-benchmarks, or by using the pre-computed thresholds in \cref{fig:snbs_thres}. The optimal strategy is as follows: 
-\begin{itemize}
-\item For Maps: use Full when the selectivity is above the threshold, and use SelVecManual when the selectivity is below the threshold.
-\item For Update: use BitmapFull or BitmapFullManual when the selectivity is above the threshold, and use SelVecManual when the selectivity is below the threshold.
-There is a small cost associated with the conversion from Bitmap to SelVec, but, in practice, other operations always follow an update (e.g., there is a projection after a scan filter).
-The subsequent operations amortize the conversion cost.
-\end{itemize}
-
-\section{End-To-End Evaluation: TPCH Q6}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/Q6.png}
-    \caption{\textbf{The TPCH Q6 query.}}
-    \label{fig:tpch_q6}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[width=.9\linewidth]{eval/q6_bench.pdf}
-    \caption{\textbf{End-To-End Evaluation: TPCH Q6}.
-The scale factor is 1 (around $600000$ tuples).
-The horizontal line indicates the performance of the optimal strategy (BitmapFull + SelVecManual).}
-    \label{fig:tpch_q6_perf}
-\end{figure}
-
-We now study the benefits of using the decision tree on a full query: TPCH Q6 \cite{tpch} (shown in \cref{fig:tpch_q6}).
-The query consists of five \texttt{WHERE} clause filters (Updates), and one aggregation (Map).
-All operations are SNBS, and all transitions are implicit.
-We implemented three mixed strategies: BitmapFull + BitmapPartial, BitmapFull + SelVecPartial, BitmapFull + SelVecManual.
-In all these strategies, the second filter triggers the threshold-based switch, which amortizes the conversion cost by a factor of four.
-
-We compare these mixed strategies to the fixed ones of previous systems.
-Vectorwise and its related systems \cite{vectorwise, miro_adapt, everything_vectorized, sompolski_vec} exclusively rely on SelVecs: they cannot use on Full Compute for the five Update filters.
-\cite{miro_adapt} does use Full Compute on Map tasks with high selectivity, but the selectivity of Q6's final vector is in the order of $0.01$; Full Compute is too wasteful in this scenario.
-SelVecPartial is, therefore, representative of these systems.
-The recent \cite{orestis_bitmap} is more similar to BitmapFull.
-
-The results are shown in \cref{fig:tpch_q6_perf}.
-As predicted in the previous chapter, BitmapFull + SelVecManual is indeed the optimal strategy.
-We can also see that the fixed strategies are all worse than the mixed ones.
-For example, SelVecPartial is $2\times$ worse that the optimal strategy; BitmapFull is $1.37\times$ worse.
-
-For software engineering reasons, it may be more practical to use the BitmapFull + BitmapPartial strategy.
-It involves only one representation (i.e., no duplicate functionality), does not require manual vectorization, and is the second-best strategy (worse by $1.2\times$).
-Software engineers should balance performance and maintainability.
-
-\iffalse
-\chapter{End-To-End Evaluation: TPCH Q6}
-
-\chapter{End-To-End Evaluation: The Join Probe}
-We now study the benefits of using the decision tree above on a multi-step operator involving both Maps and Updates: the join probe.
-
-\section{Hashing}
-\cref{fig:hashing} shows the cost of hashing for both strings and integers.
-XXHash operates on variable length data.
-As the root of decision tree indicates, using selective compute is the best strategy.
-Murmur Hash, on the other hand, is an SNBS operations.
-As such, Mixed Compute is the best strategy.
-
-\section{Hash Table Lookup}
-SIMD Hash table lookup can be implemented as vector gather using the hash table array as the sbase address, and the lower bits of the hash values as the indices.
-\cref{fig:lookup} shows the performance profile of hash table lookup.
-The compiler is somehow unable to perform auto-vectorization, so we only consider Manually written SIMD code.
-Note that the cost of a vector gather is at least 16 times that of the single-cycle instructions in the previous chapter which explains why the threshold is fairly high.
-Again Mixed Compute is the best strategy.
-
-\section{Key Equality Check}
-To perform SIMD equality check, we need to first gather the values that we will compare because they are part of an arbitrary hash table payload (which may contain a pointer to the next entry, other join keys and values).
-In addition, we need to avoid operating on null entries by using a extra SIMD mask.
-\cref{fig:key_check} shows that the BitmapFullManual + SelVecManual is best.
-We can ignore the conversion overhead because the equality check is always followed by more operations.
-
-\section{Chain Following}
-The final step we consider is chain following.
-The SIMD implementation, again involves a gather of the \texttt{next} pointer of hash table entries.
-The profile is shown in \cref{fig:chain_follow}.
-Unsurprisingly the Mixed Strategy described above wins again.
-\fi
-
-\chapter{Related Works}
-There is a large corpus of previous work dedicated to vectorized query execution.
-They mostly differ from this thesis in that they do not consider the impact of filter representation in the performance of vectorized primitives.
-We now discuss this work and how they relate to our proposed method.
-
-\cite{sompolski_vec} compares the cost of vectorization to data-centric query compilation.
-It combines the two approaches to generate new vectorized primitive at runtime, which is the method we use in \labelcref{snbs_section}.
-Using only the SelVec representation, the authors compare the performance of vectorized primitives under various compute strategies (e.g., Full v Selective Compute, Branching v Non-Branching) but do not provide a way to switch between these strategies.
-\cite{everything_vectorized} performs a similar analysis for whole queries rather than individual primitives.
-
-\cite{miro_adapt} recognizes the importance of switching compute strategy at run-time.
-For each operation, the authors implement several \textit{flavors}, i.e., different ways to perform the same operations.
-They then use reinforcement learning (RL) to switch the best \textit{flavors} dynamically at run-time.
-This approach is more general than our static micro-benchmarks because it takes into account changes in system load.
-SelVec is the only representation considered, limiting the range of adaptivity on Update tasks.
-Besides, our decision tree provides more explainability that RL techniques.
-As future work, we could employ a similarly dynamic strategy to determine our thresholds.
-
-\cite{orestis_bitmap} builds a query engine from the ground up using manually written SIMDed primitives.
-Its implementation of SNBS operations is most similar to our BitmapFullManual strategy.
-It considers a more extensive range of tasks than our side-effect-free Update/Map tasks (e.g., partitioning, building hash tables, sorting).
-As future work, we should perform our analysis on side-effect-full operations.
+4. Efficiently Compiling Efficient Query Plans
+for Modern Hardware. Early paper abouto query compilation using LLVM framework. https://15721.courses.cs.cmu.edu/spring2020/papers/14-compilation/p539-neumann.pdf
 
 \chapter{Conclusion}
-This work analyzed the impact of filter representation (i.e., Bitmap v SelVec) and compute strategy (i.e., Full v Selevective Compute) on the performance of vectorized primitives.
-We identified the factors that influence performance: iteration logic, number of tuples processed, and time spent per tuple.
-We explained how each combination of representation and compute strategy balances between these three strategies.
-Full Compute has the cheapest iteration logic, processes all tuples, but spends less time on each tuple when SIMD vectorization is possible.
-Full Compute is, however, only available with Bitmaps on Update tasks.
-Selective Compute with SelVecs has a cheaper iteration logic than Selective Compute with Bitmaps, and is more amenable to SIMD vectorization.
-Using these observations, we created a decision tree to help systems developers find the optimal representation for a given primitive and performed an empirical evaluation to confirm the shape of this decision tree.
-Finally, we showcased the benefits of our analysis using an end-to-end query involving multiple primitives.
 

### 2021-07-26 paper.tex
Updates from Overleaf

@@
 
+\newcommand{\dbSQL}[1]{\texttt{\textbf{#1}}\xspace}
+
 \DeclareUnicodeCharacter{0097}{~}
@@
 % Provides a draft mark at the top of the document. 
-%\draftstamp{\today}{DRAFT}
+\draftstamp{\today}{DRAFT}
 
@@
 \title{ %% {\it \huge Thesis Proposal}\\
-{\bf Filter Representation in Vectorized Query Execution}}
-\author{Amadou Latyr Ngom}
-\date{July}
-\Year{2020}
+{\bf Code Generation Log Replay for In-memory Database Management Systems}}
+\author{Tianlei Pan}
+\date{June}
+\Year{2021}
 \trnumber{}
@@
 \begin{abstract}
-Advances in memory capacity have allowed Database Management Systems (DBMSs) to store large amounts of data in memory, thereby shifting the performance bottleneck of query execution from disk accesses to CPU efficiency (i.e., instruction count and cycles per instruction).
-One technique used to achieve such efficiency in analytical applications is batch-oriented processing or \textit{vectorization}: it reduces interpretation overhead, improves cache locality, and allows for efficient loop optimizations (e.g., loop unrolling, SIMD vectorization).
-For each \textit{vector} (i.e., a batch of tuples), the DBMS needs to track the set of valid tuples after a predicate application.
-To that end, systems employ one of two data structures, or \textit{filter representations}: Selection Vectors (SelVecs) or Bitmaps.
-In this work, we analyze each approach's strengths and weaknesses to provide recommendations on how to implement vectorized operations.
-Through a wide range of micro-benchmarks, we determine that the optimal implementation strategy is a function of many factors: the cost of iterating through tuples, the cost of the operation itself, and how amenable it is to SIMD vectorization.
-Our analysis shows that Bitmaps perform better for operations that can be efficiently vectorized using SIMD instructions but that SelVecs perform better on all other operations due to a cheaper iteration logic.
+Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records.
+We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
 \end{abstract}
@@
 \begin{acknowledgments}
-I would like to thank my advisor Andy Pavlo for his guidance and help during my last two years at CMU.
-He has developed my interest in databases and computer science research in general and made my experience at CMU very educational and fun.
-I would also like to thank my amazing peers of the CMUDB group that helped me in this and other projects: Prashanth Menon, Matt Buttrovich, Wan Shen Lim, Lin Ma, Tanuj Nayak, Rohan Aggarwal, Gustavo Angulo, Tianyu Li, John Rollinson.
-Special shoutout to Prashanth Menon, whose help was invaluable in this project.
-Finally, I would like to thank my parents and siblings who guided and supported their son/little bro all throughout his life.
+yes
 \end{acknowledgments}
@@
 
-\chapter{Introduction and Background}
-\label{chapter1}
-DBMSs typically handle two broad classes of applications: Online Transaction Processing (OLTP) and Online Analytical Processing (OLAP) applications.
-OLTP applications tend to be short-running, write-heavy queries that only touch a few entries in a database.
-OLAP applications are long-running, read-mostly queries that analyze large amounts of data.
-Due to the unique characteristics of each class, storage engines (i.e., the component of the DBMS stores the data), and query engines (i.e., the component that executes queries) must specialize storage formats and database algorithms to efficiently handle OLAP or OLTP applications \cite{one_size_fits_all}.
+\chapter{Introduction}
+A \textit{database management system (DBMS)} is a software that is responsible for storing data, analyzing data, and interacting with applications.
+Depending on the use case, a DBMS can be designed to focus more on either capturing data, or analyzing data.
+\textit{On-line Analytical Processing (OLAP)} applications focus on reading, analyzing and aggregating cold data that is less likely to be modified.
+On the other hand, \textit{On-line Transaction Processing (OLTP)} applications work with write-heavy transactions that modify the database frequently.
 
-OLTP applications aim for fast access and update on individual entries.
-Therefore, to improve cache performance, OLTP-specialized storage engines \cite{mysql,postgres,aurora,voltdb}, use a row-store (i.e., a storage format that stores all the attributes of an entry contiguously in memory and in cache).
-OLAP applications, on the other hand, optimize for fast reads and computing on attributes across multiple entries (e.g., to find the average value of an attribute across all tuples).
-OLAP-specialized storage engines \cite{redshift,vertica,saphana,duckdb}, thus, use a column-store (i.e., a storage format that stores similar attributes of different entries contiguously).
-Though most traditional DBMSs \cite{postgres,mysql} focused on OLTP applications and thus used row stores, column stores have resulted in significant performance gains for OLAP applications \cite{abadi_phd,one_size_fits_all}.
+Consequently, recovery and durability in OLTP applications are more of a pressing issue compared to OLAP systems.
+Naive or manual backups are usually sufficient for read-only OLAP applications.
+However, since OLTP databases are being updated by transactions constantly, naive backups will either cause the database to lose significant amount of data, or fail to preserve the \textit{ACID (atomicity, consistency, isolation, and durability)} properties.
+This requires OLTP applications to develop specialized recovery mechanism to handle these problems.
+Commercial DBMS often combines Write-ahead Logging (WAL) with checkpoints to preserve database durability.
 
-Besides the storage engine, disk-accesses were, historically,  the performance bottleneck of OLAP applications.
-For example, it takes around $1ns$ to perform random access in the  L1 cache, $100ns$ for DRAM, and $17000ns$ for SSDs \cite{numbersbyyear}.
-Disk accesses, thus, dominate the execution time of any query whose data is primarily disk-resident.
-Recent advances in DRAM manufacturing have exponentially decreased the price and increased the capacity of main-memory.
-The cost of DRAM per GB was around \$1000 in 2000, \$10 in 2010, and less than \$1 in 2020 \cite{dramprices}.
-This dramatic decrease in cost, and the corresponding increase in capacity, have both obeyed Moore's Law \cite{moorelaw}.
-It is now possible to store large databases entirely in DRAM.
-Rather than optimize for disk access, the query engines of in-memory DBMSs (i.e., DBMSs whose data fits in memory) must optimize for CPU efficiency by reducing instruction count and cycles per instruction when executing queries.
+In addition, OLTP systems also require specialized algorithms and implementations to execute write transactions.
+Query compilation is an important optimization technique used in OLTP systems to greatly reduce the number of instructions that need to be executed.
+A DBMS uses a processing model (e.g.
+Iterator, Vectorization, Materialization) to determine how the DBMS executes a query plan.
+The processing model is responsible for translating a query into an intermediate expression.
+The execution engine then breaks down the expression into various tasks for execution.
+To speed up this process, the DBMS can use an optimization technique known as \textit{Query Compilation}.
+Query compilation requires the DBMS to generate specialized code for execution tasks to increase execution throughput.
+These specialized code are often created using an off-shelf compiler into a C/C++ program that implements the functionality of a certain query.
 
-\begin{figure}[t!]
+While query compilation has proved successful in increasing the efficiency of a DBMS, it comes with several drawbacks.
+Implementing a query compilation system can be quite an ordeal.
+It requires additional knowledge of compiler systems (e.g.
+LLVM) and huge amounts of engineering work to translate different execution tasks.
+Moreover, the DBMS also need to modify other components to accommodate with an execution engine reliant on query compilation.
+
+One of such components is the recovery system.
+Since the execution engine has adapted to executing compiled native code, the DBMS needs to implement an additional execution engine to interpret the log records.
+Instead, we propose a system design that will extend the query execution engine to create custom programs for replaying log records.
+
+\chapter{Background}
+\section{Database Recovery}
+\begin{figure}[H]
 \centering
-    \includegraphics[scale=0.5]{images/SampleQuery.png}
-    \caption{\textbf{Sample SQL Query and Plan.}}
-    \label{fig:intro_query}
+\includegraphics[width=0.9\textwidth]{images/Logging.png}
+\caption{Write Ahead Logging.}
+\label{fig:logging}
+\end{figure}
+Database recovery systems are responsible for preserving the ACID principles of a DBMS in case of failure.
+Prior to 1980s, early recovery systems rely heavily on non-volatile memory storage.
+Most of the DBMSs were disk-based and the recovery protocols assume the existence of dirty pages.
+System R\cite{system_r}, for instance, creates shadow copy of updated pages and maintains log record updates.
+Nowadays, both the capacity and reliability of main memory have increased significantly.
+This greatly improved the capabilities of in-memory DBMSs, but their performance is still be threatened by the need to synchronize non-volatile memory with persistent storage.
+Modern recovery systems aim to reduce the synchronization overhead.
+
+Recovery systems consist of two components: logging and recovery.
+Logging is the action of storing transactional information on disk.
+These information are stored in a special data structure called a log record.
+A log record can either contain physical information (e.g.
+changes made to a specific physical address) or higher-level information (e.g.
+user-input query).
+In Write-ahead logging \cite{aries}, the DBMS records database changes in a log file before the changes can be persisted on disk.
+Each log record in the log file usually consists of redo records (install the effects of a committed transaction) and undo records (remove the effects on an incomplete or aborted transaction).
+Since in-memory DBMSs do not persist uncommitted changes, undo records are no longer required for the recovery system.
+
+Recovery is the action of restoring the database system into a consistent state.
+If available, the DBMS first attempts to restore its previously recorded checkpoint.
+The database then initiates the log replay process to go through every log record stored since that checkpoint.
+SiloR\cite{silo_r} uses physical logging, checkpoints and an epoch-based recovery system.
+Transactions are assigned with different epochs and can only be persisted when their epochs are smaller then the current persistent epoch.
+VoltDB\cite{malviya14} uses a form of physiological logging that reduces transaction execution overhead.
+SAP HANA\cite{lee18}, however, uses a log-less approach through parallel table replication.
+
+\section{Query Compilation / Code Generation}
+When a query arrives, the DBMS parses the query into an expression tree.
+Eventually, the optimizer of the DBMS converts the expression tree into a physical plan tree.
+The physical plan represents how the query will be executed.
+It usually consists of operators that specify physical operations on the DBMS (e.g.
+inserting values into a certain memory location).
+For instance, \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100} can be represented by an update plan tree with a sequential plan as a child.
+The execution engine then uses the plan tree for execution.
+
+The Iterator model\cite{volcano} is widely used in many disk-based DBMS (e.g.
+ORACLE RDBMS, MySQL, Microsft SQL Server) to interpret plan trees.
+The model "iterates" over each operator by implementing a \dbSQL{next()} function in each operator that returns the next tuples to be operated.
+The expressions inside the operators are evaluated using the provided tuples and generates output tuples.
+This approach suffers from frequent virtual function calls for in-memory DBMS.
+The Materialization model reduces this overhead by processing all the input tuples and emitting all the outputs at once ("materializing" the result as a single row or column).
+The Materialization model is suitable for OLTP DBMS since each OLTP query only accesses a small number of tuples at a time.
+However, since the Materialization results are stored in an in-memory buffer, the execution throughput is then bound by the size of the available memory space.
+While disk-based DBMSs are less concerned about the in-memory access since most of their operations are disk-based, this is especially critical for in-memory DBMSs.
+
+Vectorization and Query Compilation are two state-of-the-art techniques employed for in-memory DBMSs to reduce memory overhead\cite{kersten18}.
+The Vectorization model is a variant of the Iterator model.
+Instead of emitting a single tuple at a time, the Vectorization model emits a batch of tuples at a time to amortize the cost of memory access.
+The functions that operate on these tuples are also specialized to be able to take in a batch of tuples at a time and more efficient compared to iterating over a single tuple at a time.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{images/QueryCompilation.png}
+\caption{Query Compilation architecture.}
+\label{fig:query_compilation}
 \end{figure}
 
-\begin{figure}[t!]
+Query Compilation eliminates the cost of interpretation by compiling the physical plan to machine code.
+The machine code of the corresponding physical plan can then be executed repeatedly by the execution engine, removing any need for virtual function calls.
+Before being compiled to machine code, the physical plan is often first converted into an intermediate representation (e.g.
+C/C++ code, LLVM IR, or custom Domain Specific Language) that follows the grammar of an imperative language.
+This simplifies the process of converting the physical plan into machine code, as the intermediate representation form is designed to resembles closely to SQL statements.
+
+Hyper\cite{neumann11} is one of the pioneers of query compilation.
+It uses a data-centric compilation approach that splits the query plan into pipelines and compiles each pipeline separately.
+Vectorwise\cite{raducanu13} chooses to pre-compile huge amounts of micro function primitives that perform simple operations on typed data.
+During runtime, the DBMS can directly invoke these compiled primitives for query execution.
+
+\chapter{Method}
+\begin{figure}[H]
 \centering
-    \includegraphics[scale=0.5]{images/Volcano.png}
-    \caption{\textbf{The Iterator Model on the sample SQL query} -- The code in red indicates virtual function calls.}
-    \label{fig:volcano_graph}
+\includegraphics[width=0.9\textwidth]{images/Proposal.png}
+\caption{Code Generation Recovery Mechanism.}
+\label{fig:codegen_recovery}
 \end{figure}
 
-The traditional query execution technique, the Iterator Model \cite{volcano}, is ill-suited for in-memory DBMSs due to CPU efficiency reasons.
-It executes a query by interpreting the query plan.
-In this model, each database operator implements a virtual \texttt{next()} function that returns the next tuple it can generate; each expression (i.e.
-addition, multiplication) also provides an virtual \texttt{evaluate()} function that takes in a tuple and returns a value.
-\cref{fig:volcano_graph} showcases the Iterator model with the sample query in \cref{fig:intro_query}.
-The operators \texttt{Scan A} and \texttt{Scan B} iterate through their respective tables and output the tuples to their parent operators.
-The filter only outputs tuples that satisfy the \texttt{WHERE} clause.
-The join first builds a hash table, then probes it to find matches.
-For every tuple -- of which there can be billions in an OLAP setting -- the virtual function calls (marked in red in the figure) constitute a significant overhead, which decreases CPU efficiency.
-Since the 2000s, there has been much work dedicated to optimizing the CPU efficiency of in-memory column stores for high-performance OLAP applications.
+Our proposal is to extend the recovery layer and query compilation layer to support the interpretation of log records. This approach can be realized in three steps:
 
-\citeauthor{monetdb2} proposed an alternative to the Iterator Model: the Materialization Model \cite{monetdb2}.
-In this model, rather than process and output one tuple at a time, each operator processes all its input tuples, then outputs all its results to the next operator.
-The authors also introduce specialized algorithms for each relational operator to exploit the columnar format of their system.
-Although the Materialization Model removes the Iterator Model's interpretation overhead, it can become memory-bound due to frequent materializations (e.g., copies) of large columns.
-To solve this issue, \citeauthor{boncz_vector} proposed a middle ground between the Materialization and Iterator Model: the Vectorization Model \cite{boncz_vector}.
-The thesis explores strategies for speeding up vectorized queries (i.e., queries executed using the Vectorization Model).
-As such, we will dedicate the rest of this chapter to an overview of the Vectorization Model.
-We discuss alternative query engine designs and optimizations for in-memory analytical DBMSs in \cref{chapter5}.
+\begin{enumerate}
+\item Extract information from log records.
+\item Convert the information into expressions that can be accepted and compiled by the code generation engine (Code Generation Step).
+\item Run the compiled machine code using the execution engine.
+\end{enumerate}
 
-\section{The Vectorization Model}
-\begin{figure}[t!]
+This code generation approach brings around several benefits.
+Firstly, it significantly reduces engineering overhead.
+Compiling contents of the log records into machine code allows the recovery manager to seamlessly integrate with the built-in execution engine.
+This removes the need to specifically construct code to handle tasks that are not compiled to machine code.
+
+\begin{figure}[H]
 \centering
-    \includegraphics[scale=0.5]{images/Vectorization.png}
-    \caption{\textbf{The Vectorization Model with the Sample Query} -- The code is nearly identical to that in \cref{fig:volcano_graph}, except that it processes vectors.}
-    \label{fig:vectorization_graph}
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=0.9\textwidth]{images/RecoverIndex.png}
+ \caption{Index Update During Recovery.}
+  \label{fig:pipeline_graph}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+ \centering
+ \includegraphics[width=0.9\textwidth]{images/RecoverConstraint.png}
+ \caption{Constraint Checking During Recovery.}
+  \label{fig:pipeline_code}
+\end{subfigure}
+\caption{Problems faced by conventional recovery in code generation DBMS.}
+\label{fig:log_record}
 \end{figure}
 
-The Vectorization Model is almost identical to the Iterator model except that it operates on batches of $T$ tuples or \textit{vectors} (e.g., 2048 tuples at a time).
-The \texttt{next()} function returns vectors and the \texttt{evaluate()} function operates on vectors too.
-\cref{fig:vectorization_graph} showcases Vectorization Model on the sample query in \cref{fig:intro_query}; it is near identical to \cref{fig:volcano_graph}.
-The difference is that processing occurs one vector at at time rather than one tuple at a time.
+Secondly, it also resolves tricky problems that will appear when writing specialized code for log records.
+An example of this is index updates.
+Without code generation, the recovery manager needs to specifically reconstruct the index given an insert or update record.
+With code generation, the index updates are taken care directly by the execution engine that already contains high-performance specialized code for index maintenance.
+Similar problems, such as constraint checking
 
-\begin{figure}[t!]
+Our proposed architecture is implemented and tested in \textit{Noisepage} \cite{noisepage}, a self-driving in-memory DBMS developed at Carnegie Mellon University.
+Noisepage is built in C++ and uses the PostgreSQL wire protocol for user communication.
+The DBMS supports ACID transactions and \texttt{SNAPSHOT-ISOLATION}.
+Noisepage uses the conventional WAL recovery scheme with physical logging and depends on query compilation for query execution.
+We will dive into the details of Noisepage's recovery structure in the following sections.
+
+\section{Recovery System Architecture}
+\subsection{Transaction}
+Transactions are the atomic operations of any DBMS.
+Noisepage uses a transaction to perform physical updates on the database and physical logging to disk.
+Noisepage uses a multi-versioned delta store transaction engine that ensures \texttt{SNAPSHOT-ISOLATION}.
+The transaction engine allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis.
+A transaction's lifecycle begins when the transaction begins, and ends after the log manager has serialized its changes to disk.
+
+Each data tuple in noisepage is uniquely identified through a tuple slot that stores the offset of a tuple.
+When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table.
+Subsequent updates to the tuple does not create direct copies of the tuple, but stores delta information (redo records) about the updates instead.
+These redo records are essential for executing log replay correctly.
+The structure of a redo record is shown later in \ref{fig:log_record}.
+
+A transaction uses a buffer (redo buffer) that stores all the delta records that will appear in its lifetime.
+Each redo buffer has a fixed size of 4096 bytes and is allocated from a centralized buffer pool.
+Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to dump a corresponding delta record of the change.
+If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the buffer pool.
+The redo buffer allows the logging components to process the changes before the transaction is completed.
+
+When a transaction commits, the transaction engine creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager.
+The commit record is appended to the redo buffer.
+The redo buffer is then carried over to the log manager, where the changes will be serialized to disk.
+When a transaction aborts, the transaction creates an abort record that will prevent corresponding records to be replayed during recovery.
+Abort records are essential; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk.
+Aborted transactions that are persisted on disk are rare, however, since it would be difficult for a typical OLTP transaction to fill up the entire redo buffer.
+
+\subsection{Write-Ahead Logging}
+\begin{figure}[H]
 \centering
-\begin{subfigure}[t]{.5\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/MulByConstSISD.png}
- \caption{Unoptimized version.}
-  \label{fig:vector_mul_sisd}
-\end{subfigure}%
-\begin{subfigure}[t]{.5\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/MulByConstSIMD.png}
- \caption{Optimization version. There are two optimizations here: SIMD vectorization and loop unrolling with a factor of 2. The code in green indicates Intel's AVX512 instructions.}
-  \label{fig:vector_mul_simd}
-\end{subfigure}
-\caption{\textbf{A Vectorized Multiplication Function} -- The programmer usually writes only the code on the left; the compiler can perform the optimizations on the right.}
-\label{fig:vector_mul}
+\includegraphics[width=0.9\textwidth]{images/WAL.png}
+\caption{Logging Architecture. Log manager uses the log serializer and log consumers to convert redo buffers into log records.}
+\label{fig:wal}
 \end{figure}
 
-This batch-oriented processing provides several benefits.
-First, virtual function calls occur for every $T$ tuples, which amortizes their overhead by a factor of $T$.
-Second, materializing $T$ tuples at a time instead of all tuples eliminates the memory pressure of the Materialization Model.
-Third, functions that operate on vectors, also known as \textit{vectorized primitives}, are usually amenable to loop optimizations (e.g., loop unrolling, SIMD vectorization) applied by either the compiler or the programmer.
-Consider, for a example, the primitive shown in \cref{fig:vector_mul_sisd}.
-It multiplies the elements in a vector by a constant number, and stores the result in a new vector.
-The loop of this primitive has independent iterations.
-Furthermore, the body of the loop only contains a load, a multiplication, and a store, which can all operate on SIMD vectors.
-As a result, the compiler or programmer can unroll and SIMD vectorize the loop.
-\cref{fig:vector_mul_simd} shows one way to optimize the loop.
-The body now contains AVX512 instructions \cite{intel2019} that perform that load, the multiplication, and the store, each operating on eight elements at a time.
-Besides, the loop unrolling (by a factor of two) optimization reduces the cost of loop maintenance.
+The log manager is responsible for performing write-ahead logging in Noisepage.
+All the changes are serialized to disk as logs before they are performed on the DBMS storage.
+Each log record is self-contained and does not require any additional metadata to be replayed.
+The log manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.
 
-Despite these benefits, the Vectorization Model introduces a new challenge.
-Individual tuples within a vector may be \textit{filtered out}, meaning that subsequent operators should not process them.
-For example, in \cref{fig:vectorization_graph}, the \texttt{Join} operator should not insert tuples that do no satisfy the \texttt{WHERE} clause in the \texttt{JoinHashTable}.
-It should only insert \textit{selected} tuples -- those not filtered out.
-Next, we discuss the methods used to identify selected tuples within a vector.
+\subsubsection{Log Serializer}
+The log serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit.
+The log serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers of 4096 bytes.
+These buffers will be consumed by the Disk Log Consumer for persistence.
 
-\section{Filter Representation}
-\begin{figure}[t!]
+The log records must be serialized in the correct order for recovery to perform correctly.
+Different transactions can share a same redo buffer that is filled with each of their own log records.
+Moreover, it is possible for transactions to appear in a non-serial order relative to their begin timestamp.
+However, the log records performed by an individual transaction is guaranteed to appear in the same order as they were created.
+This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of Noisepage during log replay.
+\subsubsection{Log Consumer}
+Buffers supplied by the log serializer can be consumed by different log consumer tasks.
+Each log consumer is given a copy of the original buffer to achieve parallelism.
+Noisepage uses two type of log consumers: disk log consumer and replication log consumer.
+
+The disk log consumer repeatedly polls for new buffers from the log serializer and writes the changes to disk.
+To improve performance, disk log consumer uses group commit that allows a batch of changes to be committed over a time period specified by the user.
+Under default settings, the disk log consumer writes down changes every 10 microseconds of if more than 1 MB of data has been written since the last write.
+
+The replication log consumer also polls for new buffers, but instead of persisting them to disk, sends the new buffers over across the network to database replicas.
+
+\subsection{Log Replay}
+\begin{figure}[H]
 \centering
-    \includegraphics[scale=0.7]{images/FilterRepresentationIntro.png}
-    \caption{\textbf{Filter Representation} -- Selection Vectors (left) index into selected tuples. Bitmaps (right) are only set on selected rows.}
-    \label{fig:repr_intro}
+\includegraphics[width=0.9\textwidth]{images/RecoveryManager.png}
+\caption{How recovery operates in Noisepage.}
+\label{fig:recoverymanager}
 \end{figure}
 
-A \textit{filter representation} is the data structure used to identify selected tuples with a vector.
-\cref{fig:repr_intro} displays the two common representations: Selection Vectors (SelVecs) and Bitmaps.
-A SelVec stores a list of indices into the selected tuples.
-Tuples at an index stored within the SelVec are selected; tuples without an index are implicitly filtered out.
-A Bitmap, on the other hand, stores a list of bits.
-Its set bits indicate the selected tuples; its unset bits indicate filtered out ones.
-In the rest of this thesis, we will use the noun \textit{filter} to denote a particular instance of either data structure.
+Noisepage uses four different record types: \textit{Redo, Delete, Commit, Abort}.
+Redo and Delete records are not processed immediately.
+Instead, they are placed inside a deferred changes buffer, ordered by their transaction timestamp.
+An Abort record removes the corresponding Redo/Delete record from the buffered changes.
+A Commit record initiates a transaction to iterate over the deferred changes buffer.
+Once all the deferred changes are processed, the buffer is cleared and the transaction is committed.
+The transaction aborts if any log record is malformed or any recovery operation fails.
 
-\subsection{Contribution}
-Previous works on vectorized execution have chosen a filter representation strategy without providing an experimental comparison of the two approaches.
-Vectorwise \cite{vectorwise}, and works derived from it \cite{miro_adapt,everything_vectorized,sompolski_vec},  rely on SelVecs.
-IBM DB2's BLU \cite{ibm_blu} and the more recent VIP \cite{orestis_bitmap} rely on Bitmaps for the intermediary results of a table scan's \texttt{WHERE} clause filters, and SelVecs for other relational operators.
-Nonetheless, we find that supporting both representations, and dynamically choosing between them results in better performance than static implementations.
-Depending on the specific primitive and on the selectivity (i.e., the ratio of selected tuples) of its input vector, SelVecs can outperform Bitmaps and vice-versa.
-This work's main contribution is a methodology on how to optimize the performance of arbitrary vectorized primitives by taking into account filter representation, selectivity, and loop optimizations.
-We also provide recommendations for developers working on vectorized query engines on how to best implement primitives.
-
-The remainder of this thesis is structured as follows.
-In \cref{chapter2}, we analyze the factors that influence the performance of vectorized primitive.
-We then derive insights for developers implementing vectorized query processing in \cref{chapter3}.
-Next, we use these insights on the OLAP queries of the TPCH \cite{tpch} benchmark and show the performance gains we obtain in \cref{chapter4}.
-Finally, we discuss related work in \cref{chapter5} and conclude in \cref{chapter6}.
-
-\chapter{Computing on Filtered Vectors}
-\label{chapter2}
-The purpose of this chapter is to discuss the factors that influence the performance of vectorized primitives.
-We first categorize the kinds of operations that primitives perform and the possible ways to implement them.
-Afterward, we will present and explain the equation that quantifies the performance of each implementation.
-Finally, we raise the main questions whose answers, provided in \cref{chapter3}, will determine how to implement primitives optimally.
-
-\section{Operations on Filtered Vectors}
-\begin{figure}[t!]
+\subsubsection{Data Structure}
+\begin{figure}[H]
 \centering
-\hspace*{\fill}%
-\begin{subfigure}{.8\textwidth}
+\begin{subfigure}{.5\textwidth}
  \centering
- \includegraphics[width=0.9\textwidth]{images/UpdateIntro.png}
- \caption{Update operation.}
-  \label{fig:update_intro}
+ \includegraphics[width=0.9\textwidth]{images/RedoRecord.png}
+ \caption{Redo Record Structure.}
+  \label{fig:pipeline_graph}
 \end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}{.8\textwidth}
+\begin{subfigure}{.5\textwidth}
  \centering
- \includegraphics[width=0.9\textwidth]{images/MapIntro.png}
- \caption{Map operation.}
-  \label{fig:map_intro}
+ \includegraphics[width=0.9\textwidth]{images/DeleteRecord.png}
+ \caption{Delete Record Structure.}
+  \label{fig:pipeline_code}
 \end{subfigure}
-\caption{\textbf{Operations on Filtered Vectors} -- Updates change the set of selected tuple.
-Maps apply a function to the set of selected tuples without updating the set.
-Selected input tuples are in green.
-Selected output tuples are in blue.}
-\label{fig:update_map_intro}
-\end{figure}
-\cref{fig:update_map_intro} shows the two kinds of primitives we seek to optimize: Updates and Maps.
-Updates apply a predicate function to the vector and modify the set of selected tuples accordingly.
-A scan filter (i.e., \texttt{WHERE} clause) is an example of an update operation.
-On the other hand, Maps keep the selected set but compute a new vector using a mapping function.
-A projection (i.e., \texttt{SELECT} clause) is an example of a map operation.
-There are other primitives, but these often have side-effects (e.g., insertion in a hash table for joins or an array for sorting) and are, therefore, not amenable to our optimizations for correctness reasons (e.g., Full Compute introduced in \cref{computestrategies}).
-As such, we focus on side-effect-free primitives.
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.45]{images/NewJoinProbeExample.png}
-    \caption{\textbf{Multi-Step Join Probe} -- The color green indicates selected items in the input vector.
-Orange indicates hash table lookup hits.
-Blue indicates exact matches after key comparison; these represent the output of the probe.
-Greyed out squares are filtered out tuples, or empty hash table chains.}
-    \label{fig:probe_example}
-\end{figure}
-All relational operators are decomposable into a set of consecutive primitives.
-As mentioned above, simple relational operators like scan filters or projections consist of one individual primitive.
-Complex operators, like joins and aggregations, consist of multiple primitives.
-As a proof of concept, we showcase, in detail, the example of join probes in \cref{fig:probe_example}.
-The first primitive computes the hashes of the input tuples; the result is a vector of hash values.
-It is, therefore, a Map primitive.
-The second primitive performs a hash table lookup using the hashes to obtain a vector of hash table entries.
-It is also a Map primitive.
-Due to hashing collisions, the entries may not match the input, so the algorithm compares the join keys to filter out inexact matches.
-For example, key 90's hit in the hash table is an exact match, whereas key 73's is not.
-The third primitive performs this comparison; it is an Update primitive.
-The fourth primitive advances the hash table chains to obtain the next entries (e.g., 85 follows 37); it is a Map primitive.
-The probe then loops back to the comparison primitive to find exact matches (e.g., 85 now has an exact match).
-The set of exact matches constitutes the output of the join probe.
-
-Other relational operators (e.g., aggregations) are similarly decomposable.
-
-\section{Transitions}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/DisjunctionHowto.png}
-    \caption{\textbf{Decomposed Disjunctions} -- We apply the first clause (in orange) on all input tuples.
-In the second step, however, we only apply the second clause (in red) on those tuples that do not pass the first clause.
-In the third step, we take the union between the two outputs.
-The output tuples are in blue.}
-    \label{fig:disjunctive_filter_howto}
+\caption{The internal log structures.}
+\label{fig:log_record}
 \end{figure}
 
-In between each primitive, operators may need to perform a \textit{transition} (i.e., modifications to the filter passed into the next primitive).
-We now discuss transitions and their associated cost.
-Most of the time, this cost is non-existent because the same filter is passed on to the next primitive: the transition is \textit{implicit}.
-For example, the hashing primitive and the lookup primitive in \cref{fig:probe_example} have the same filter; the transition performs no modifications.
-On the other hand, the Chain Following primitive operates on entries that do not satisfy the comparison: the transition involves a set difference.
-Set unions are another type of transition.
-Decomposed disjunctions \cite{pcq} (shown in \cref{fig:disjunctive_filter_howto}), for example, involve both set differences and unions.
-They decompose disjunctive \texttt{WHERE} clauses into a sequence of primitives -- one each for each conditional term.
-Each primitive operates on only tuples that have not satisfied the previous conditional terms, hence the set differences.
-The final filter consists of tuples that satisfy at least one primitive, hence the unions.
+A physical log record consists of a header followed by a record body.
+The header stores metadata about the log record itself, including its record type, size and begin timestamp of the transaction that generated the log record as mentioned in Section 3.1.1.
+The record body contains physical information that is necessary for recovery and varies depending on the record type.
+Depending on the value type, the values may bee stored in-line or not in-inline.
+Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may stores the pointer in the values column that points to another location in the record.
 
-To the best of our knowledge, transitions are either implicit or take the form of set differences (for elements that do not pass the previous predicate) and unions (to obtain the final result).
-These operations are efficient to implement with Bitmaps by leveraging single-cycle bitwise AVX512 instructions \cite{intel2019} that operate on 512 bits at a time.
-Efficiently implementing set operations on SelVecs is more complicated.
-Whenever necessary, Updates can write to one additional SelVec for tuples that do not pass the filter; this second SelVec implicitly corresponds to the result of a set difference.
-Note that it does not overlap with the first SelVec, so a union of any of their subsets is just an array append, a fast operation that can leverage SIMD instructions.
+Tuple slots in the log records will no longer be valid memory locations during recovery.
+Instead, the recovery manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted.
+The mapping allows the recovery manager to apply changes to the correct memory locations in the new memory environment.
 
-To show that the overhead of these set operations is similar and negligible for both Bitmaps and SelVecs, we run an experiment that performs the decomposed disjunction shown in \cref{fig:disjunctive_filter_howto}.
-We synthetically generate the data so that each tuple has a $50\%$ chance of satisfying either conditional terms, the average case scenario.
-The data contains around $2000000$ tuples split into $1000$ vectors.
-The results are shown in \cref{fig:disjunctive_filter_perf}.
-The overheads are negligible; they are also nearly equal ($3.55\%$ and $4.06\%$ for Bitmaps and SelVecs, respectively).
-The representation, therefore, matters little when it comes to transitions.
-Given this, we will focus on optimizing primitives and will ignore transition costs.
+Since only Redo and Delete records will effectively modify physical storage, we only need to analyze the record body of those two record types.
+Modifications of a query to the DBMS can be resolved down to the modification of tuple slots on the physical level.
 
-\begin{figure}[t!]
+\begin{figure}[H]
 \centering
-    \includegraphics[scale=0.5]{eval/transition.pdf}
-    \caption{\textbf{Transition Overheads} -- The overheads are $3.55\%$ and $4.06\%$ for Bitmaps and SelVecs respectively.}
-    \label{fig:disjunctive_filter_perf}
+\includegraphics[width=0.9\textwidth]{images/Delta.png}
+\caption{Redo Record Delta.}
+\label{fig:logging}
 \end{figure}
 
-\section{Compute Strategies}
-\label{computestrategies}
-\begin{figure}[t!]
+ While storing tuple slots is sufficient for delete records, Redo Records require additional data to perform Recovery Insert and Update operations.
+These data are kept in the Record Delta, which stores a mapping from column ids to values.
+
+\subsubsection{Recovery Operation}
+A Noisepage recovery step has three different operations:
+\begin{itemize}
+    \item \textbf{Insert}: Represented by a Redo Record. The recovery manager inserts the values stored within the redo record into the given tuple slot and updates the indexes.
+    \item \textbf{Delete}: Represented by a Delete Record. The recovery manager deletes the tuple slot specified by the delete record and updates the indexes.
+    \item \textbf{Update}: Represented by a Redo Record. The recovery manager updates the tuple slot using the values within the redo record and updates the indexes.
+\end{itemize}
+
+While both Insert and Update are represented using the same Redo Record structure, they can be differentiated by keeping track of all the tuple slots found in log records.
+If a tuple slot inside a Redo Record is unseen before, then the Redo Record represents an Insert operation.
+Otherwise, the Redo Record represents an Update operation.
+
+\section{Recovery Code Generation}
+The code generation structure in Noisepage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \ref{fig:query_compilation}.
+A query is interpreted, optimized and converted into a physical plan tree.
+A plan tree consists of plan nodes and specifies how the query should be executed on the physical level of the database.
+
+A physical plan tree/node is then converted into an operator pipeline through a plan translator.
+Each plan translator correspond to a plan node in the original plan tree.
+A pipeline contains an ordered collection of relational operators that are not yet materialized, ended with a pipeline breaker (e.g.
+sort tuples, creating a hash table).
+The pipeline is only materialized whenever a pipeline breaker has been encountered.
+
+\begin{figure}[H]
 \centering
-\hspace*{\fill}%
-\begin{subfigure}{.8\textwidth}
+\includegraphics[width=0.9\textwidth]{images/LLVM.png}
+\caption{Lifecycle of a TPL code fragment. Execution can run in either interpret or compile mode.}
+\label{fig:llvm}
+\end{figure}
+Noisepage then compiles each operator pipeline into a domain-specific language referred to as TPL.
+Each function of a query inside the TPL code is then compiled to bytecode called TBC.
+TBC can be interpreted by a built-in virtual machine, or further compiled into an LLVM module.
+The execution engine can either interpret TBC, or run the compiled LLVM module.
+
+\begin{figure}[H]
 \centering
- \includegraphics[width=0.9\textwidth]{images/SelectiveCompute.png}
- \caption{Selective Compute. An \texttt{x} indicates an uninitialized value.}
-  \label{fig:compute_intro_selective}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/FullComputeMap.png}
- \caption{Full Compute. There are no uninitialized values}
-  \label{fig:compute_intro_full}
-\end{subfigure}
-\caption{\textbf{Compute Strategies}}
-\label{fig:compute_intro}
+\includegraphics[width=0.9\textwidth]{images/CodegenExample.png}
+\caption{Code Generation of \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}}
+\label{fig:codegen_example}
 \end{figure}
 
-A compute strategy determines the implementation of Map and Update primitives.
-We consider three compute strategies: (1) Selective, (2) Full, and (3) Mixed Compute.
-Selective Compute only applies the given functions on selected tuples, whereas Full Compute applies them on all tuples.
-\cref{fig:compute_intro} shows each approach's effect.
-The Selective Compute strategy (\cref{fig:compute_intro_selective}) leaves undefined values because it does not operate on filtered out entries, whereas the Full Compute strategy (\cref{fig:compute_intro_full}) operates of every tuple and thus produces a result for each.
-As we show in \cref{chapter3}, although Full Compute usually performs more work, it benefits from SIMD vectorization, simple loop structure (for loop unrolling and interleaving), and easy branch prediction.
-To exploit this trade-off, the third strategy, Mixed Compute, switches from Full Compute to Selective Compute when the selectivity (i.e., the ratio of selected tuples) goes below a threshold.
-The next chapter will show how to derive this threshold experimentally.
+We further demonstrate the process of query compilation through the example in \ref{fig:codegen_example}.
+To run the query \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}, the DBMS needs to first do a sequential scan to find out the columns, and then do a delete operation on the found columns.
+These two steps are manifested in the delete plan node that includes a sequential scan plan node as a child.
+The plan node is translated into an operator pipeline that specifies how each operation will be exactly executed in the DBMS.
+The operator pipeline is then converted to TPL code.
+Each function inside the TPL code correspond to a piece of function inside the execution engine that will be evetually invoked once the compiled native code is run.
+For instance, \texttt{@tableDelete} corresponds to a C++ function \texttt{OpStorageInterfaceTableDelete} (a delete operation on a tuple slot of a table) in the execution engine.
 
-\begin{figure}[t!]
+\begin{figure}[H]
 \centering
-    \includegraphics[scale=0.5]{images/FullComputeUpdate.png}
-    \caption{\textbf{Update with Full Compute} -- The leftmost vector is the input vector; its selected elements are shown in green.
-The middle vector is the intermediary output of the full compute; its selected elements are shown in orange.
-The last vector has the intersection of the input filter and the intermediary filter; its selected elements are shown in blue.}
-    \label{fig:full_compute_update}
+\includegraphics[width=0.9\textwidth]{images/ExpressionGeneration.png}
+\caption{Process for converting log records into native code.}
+\label{fig:logging}
+\end{figure}
+We imagine each recovery operation as a form of SQL statement that can be converted into native code through the code generation process discussed above.
+The code generation process for a log record can be divided into two steps: plan generation (creating the physical plan node) and plan translation (translating the plan into pipelines).
+Each recovery operation (\textbf{Insert}, \textbf{Update}, \textbf{Delete}) will be matched to a corresponding plan node.
+
+\subsection{Insert Operation Conversion}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{images/InsertConversion.png}
+\caption{Conversion from an Insert Redo Record into an Insert Plan Node. The tuple slot provided by the redo record is discarded.}
+\label{fig:insert_conversion}
+\end{figure}
+Insert plan nodes correspond to \texttt{INSERT} statements.
+For an insert redo record, we can represent it as the SQL statement \dbSQL{INSERT INTO x VALUES y}.
+The insert plan node for this statement does not require a child plan node to function.
+This makes it straightforward to convert from an insert redo record to a plan node.
+Index information and column ids can be retrieved from the catalog.
+Values stored in the redo records are in their most basic form (e.g.
+integer, float, double) as constants and can be converted into expressions to be accepted by the insert plan node.
+The operation of transferring values, however, require one-to-one copy of each value from the log record into the plan node.
+
+\subsection{Delete Operation Conversion}
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{images/DeleteConversion.png}
+\caption{Conversion from a Delete Redo Record into a Delete Plan Node. Tuple slot that will be used in the pipeline is "smuggled" from the redo record through execution context.}
+\label{fig:delete_conversion}
 \end{figure}
 
-On Update primitives, Full Compute first finds the set of all tuples in the vector that pass the predicate, regardless of the input filter.
-The final filter is the intersection of this intermediary filter and the input filter.
-This process is shown in \cref{fig:full_compute_update}.
-Notice how the intermediary filter contains elements filtered out of the input vector (e.g., 100 and 200).
-The intersection is, thus, necessary to remove these elements from the final output.
-Bitmap intersection is efficient thanks to the AVX512 \texttt{and} instruction that intersects 512 bits in one cycle \cite{intel2019}, but SelVecs require a slow intersection of sorted sets \cite{sorted_set}.
-Thus, Full Compute is not efficiently compatible with SelVecs on Update primitives.
-This incompatibility will become important when choosing a strategy to implement.
+Delete plan nodes correspond to \texttt{DELETE} statements and require child nodes to function.
+A delete plan node uses the child plan node to figure out which tuple slot to schedule for deletion.
+This is different for delete records.
+The mapping from recovery manager combined with the original tuple slot already points to the tuple slot to delete.
+In this case, there will be no need for the plan node to go through the child node and resolve the delete tuple slot.
+This is because, eventually, the delete plan node will attempt to create a Delete Record, which is structurally identical to the Delete Record from the recovery manager.
+Other operations, such as index deletions, are still necessary.
 
-\begin{figure}[t!]
+\subsection{Update Operation Conversion}
+Update plan nodes correspond to \texttt{Update} statements and require child nodes to function.
+Unfortunately, update plan nodes are not very compatible with update redo records.
+After an update plan is translated into an update operator pipeline, it expects set clauses returned from the child node (e.g.
+sequential scan).
+These set clauses, however, describe higher-level logical operations and cannot be constructed easily as creating primitive expressions for insert plan nodes.
+\begin{figure}[H]
 \centering
-\hspace*{\fill}%
-\begin{subfigure}[t]{.8\linewidth}
+\includegraphics[width=0.9\textwidth]{images/UpdateConversion.png}
+\caption{Conversion from an Update Redo Record into an Update Plan Node. Set clauses and child plan nodes inside the update plan node are discarded.}
+\label{fig:update_conversion}
+\end{figure}
+The update plan node therefore needs to be extended to accept primitive values just like an insert plan node.
+We then modify the operator pipeline to process the primitive values instead of set clauses during a recovery step.
+Functions inside the operator pipeline will now apply to the primitive value expression.
+Similar to delete record transformations, there is no need for the update pipeline to resolve the update tuple slot or creating a new update record, as these information already exist in an update redo record and the tuple slot mapping in the recovery manager.
+
+\section{Execution}
+The compiled query is finally run by the execution engine.
+The compiled query invokes different operation codes (e.g.
+\texttt{OpStorageInterfaceTableDelete} in \ref{fig:codegen_example}) to complete the corresponding operator in the TPL code.
+
+We explained earlier that the for physical plan nodes, they are expected to retrieve relevant tuple slots from the recovery manager, instead of supplying the tuple slots for an operation.
+We can circumvent this rule by modifying operation code functions.
+
+\begin{figure}[H]
 \centering
- \includegraphics[width=0.9\linewidth]{images/ConditionalStoreConditional.png}
- \caption{Conditional Store. We only update the SelVec when the condition is satisfied (code in green).}
-  \label{fig:conditional_store}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{images/ConditionalStoreUnconditional.png}
- \caption{Unconditional Store. We always update the SelVec regardless of the condition (code in green). The condition determines whether to increment the write index or not.}
-  \label{fig:unconditional_store}
-\end{subfigure}
-\caption{\textbf{Conditional versus Unconditional Store} -- Both listings select the elements within the input vector (first parameter) that are less than the second parameter.
-Initially, the SelVec (third parameter) contains selected indices of the input vector.
-In the end, it only contains those that pass the predicate.}
-\label{fig:conditional_selvec}
+\includegraphics[width=0.9\textwidth]{images/RecoveryExecution.png}
+\caption{Information from log record can be retrieved from the execution context.}
+\label{fig:recovery_execution}
 \end{figure}
 
-Furthermore, for Updates with SelVecs, there is another implementation decision: Conditional versus Unconditional Store.
-\cref{fig:conditional_selvec} shows these options.
-Both primitives attempt to select the values less than a given input constant.
-They differ only in how they update the SelVec (code in green).
-The Conditional Store (\cref{fig:conditional_store}) updates the SelVec when the condition is true.
-The Unconditional Store (\cref{fig:unconditional_store}) always the updates the SelVecs, but only increments the write index when the condition is true.
-We do not focus on this decision, for it has been studied extensively before \cite{miro_adapt, sompolski_vec}.
-Previous work identified branch mispredictions costs as the main factor influencing this decision.
-The Conditional Store implementation is competitive when the same branch is taken around $90\%$ of the time or more, meaning that less than $10\%$, or more than $90\%$ of tuples satisfy the condition.
-All the experiments in this thesis use the optimal strategy on Update primitives that use SelVecs.
+An execution context encapsulates information that is supplied by upper layers of the DBMS.
+These information include access to the database catalog, the transaction running the query, etc.
+For our purposes, we can add tuple slot information into the execution context.
+Since the execution context can be accessed throughout the execution of a compiled query, we can modify the behavior of operation codes to override with the recovery tuple slots if necessary.
 
-Given a compute strategy and a filter representation, the next section provides a framework for analyzing the performance of any vectorized primitive.
+For Delete and Update operations, we need to supply the execution context with the log record itself.
+Since we modified the delete translator to skip the process of looking through child plan node for objects to delete, we will deliver the Delete Record from the recovery manager directly to the operation codes that will be expecting a Delete Record generated by a regular delete plan node.
 
-\section{Primitive Performance}
-\label{prim_perf_section}
+\section{Caching Optimization}
 
-\begin{figure}[t!]
+The naive implementation for code generation searches for table metadata and re-compiles the native code every time the recovery manager reads in a new log record.
+This is very inefficient and requires huge amounts of memory access and copy.
+We can use caching techniques to reduce performance overhead.
+
+\subsection{Metadata}
+Most of the metadata overhead is generated during the conversion step from a redo record into an insert plan node.
+For each table, the DBMS needs to figure out its relevant metadata.
+These operations require frequent lookup into the database catalog.
+As long as those metadata are bound to some table id, the DBMS can cache those metadata and re-use them for the same table.
+Some significant metadata are listed below:
+
+\subsubsection{Physical to Logical Column Mapping}
+The recovery manager needs to maintain a mapping from physical column ids to logical column ids for each table.
+This is required because the column ids recorded in the log records are physical ids, while all the query compilation components expect logical column ids.
+The mappings can be generated by looking into the DBMS catalog.
+\subsubsection{Column Value Type Mapping}
+For redo records, the recovery manager also needs to maintain a mapping from each column to its corresponding SQL value type (e.g.
+\dbSQL{INTEGER, FLOAT, VARLEN}).
+This mapping is used to correctly copy the values out of each redo record into a physical plan node.
+
+\subsection{Compiled Query}
+A major benefit of Query Compilation is that the compiled query can be reused by the execution engine.
+If the contents of a physical plan node remains largely unchanged, then there is no need to reconstruct the physical plan every time the recovery manager processes a log record.
+
+We can cache each type of recovery operation separately and use a combination of database and table id as unique identifiers.
+The cached query can be re-usesd as long as the table schema remains unchanged.
+
+\subsubsection{Parameterization}
+\begin{figure}[H]
 \centering
-\hspace*{\fill}%
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RationaleMapFull.png}
- \caption{Full Compute. This function iterates through all elements, including filtered out ones.}
-  \label{fig:rationale_full}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%  
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RationaleMapBitmap.png}
- \caption{Selective Compute with a Bitmap. The iteration logic is adapted from \cite{bitmap_iteration}. The logic is not amenable to SIMD vectorization.}
-  \label{fig:rationale_bitmap}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%  
-\begin{subfigure}{.8\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RationaleMapSelVec.png}
- \caption{Selective Compute with a SelVec.
-Reading an element involves an indirection to first obtain its index.
-In the SIMD code, the \texttt{gather} instruction handles this indirection.
-The last \texttt{scatter} is only necessary to maintain consistent indices between the input and output vector during multi-step Map operations.
-It should otherwise be avoided due to its slowness.}
-  \label{fig:rationale_selvec}
-\end{subfigure}
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\caption{\textbf{Implementing Vector Primitives} -- The code in green indicates the core operation (multiplication). The code in red indicates iteration logic.}
-\label{fig:rationale_code}
+\includegraphics[width=0.9\textwidth]{images/ParameterizedRedoRecord.png}
+\caption{Parameterized insert plan.}
+\label{fig:parameterization}
 \end{figure}
 
-This section analyzes the performance of primitives according to their implementations.
-Our primitive implementations follow one of the patterns shown in \cref{fig:rationale_code}.
-This listing shows the strategies to multiply a vector by a constant (a Map primitive) along with their optimized versions using SIMD instructions: Full Compute (which does not depend on representation), Selective Compute with SelVecs, and Selective Compute with Bitmaps.
-In each instance, the code is a loop divided in two parts: the \textit{iteration logic} (code in red), and the \textit{core operation} (code in green).
-The core operation (e.g., multiplication by a constant) is independent of the strategy.
-On the other hand, the iteration logic is strategy dependent; it determines how to iterate through tuples -- whether selected ones in the case of Selective Compute or all tuples in the case of Full Compute -- to perform the core operation and store its result.
+Caching is trickier for recovery operations that need to transfer values into plan nodes.
+A primitive expression will change every time its underlying value changes.
+A solution to this is to parameterize the primitive expression given to a insert/update plan node.
+During run-time, the execution context pass in actual values into the parameterized expressions.
+This approach allows a compiled query to keep value type information for each table column and remain flexible enough to swap in actual values without recompilation.
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/PerfEq.png}
-    \caption{\textbf{The Primitive Performance Equation}}
-    \label{fig:perf_equation}
-\end{figure}
+\chapter{Experimental Evaluation}
+We run our experiments on Ubuntu 20.04.
+Dual-socket 10-core Intel Xeon E5-2630v4 CPU, 128 GB of DRAM, and a 500 GB Samsung 970 EVO Plus SSD.
 
-The equation in \cref{fig:perf_equation} shows the formula to compute the running time of a primitive.
-The first equality indicates that the total running time is equal to the number of tuples processed multiplied by the time spent per tuple.
-The second equality splits the time spent per tuple in two: the time dedicated to the iteration logic, and the time dedicated to the core operation.
-This formula provides a framework to analyze any vectorized primitive.
-Let us consider how each of the three factors in the second equality relates to implementation strategies.
-
-\textbf{Number of Tuples Processed}: Selective Compute processes only selected tuples; Full Compute processes every tuple.
-Thus, the lower the selectivity (i.e., the ratio of selected tuples), the more wasteful Full Compute is.
-
-\textbf{Iteration Logic Time per Tuple}: Full Compute, while processing more tuples, has the simplest iteration logic because it does not aim to identify the selected ones.
-For Selective Compute, iterating over a Bitmap (e.g., \cref{fig:rationale_bitmap}) is more expansive than iterating over a SelVec (e.g., \cref{fig:rationale_selvec}) because of the bit operations required by Bitmaps.
-Optimizations like loop unrolling and SIMD vectorization can further decrease the iteration logic time per tuple, but they are only available for Full Compute and SelVecs.
-The Bitmap's selective iteration logic is too complex to perform such optimizations.
-
-\begin{figure}[t!]
-\centering
-\hspace*{\fill}%
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{images/BranchingCodeSISD.png}
- \caption{SISD code.}
-  \label{fig:branching_sisd}
-\end{subfigure}%
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{images/BranchingCodeSIMD.png}
- \caption{SIMD code.}
-  \label{fig:branching_simd}
-\end{subfigure}
-\caption{\textbf{SISD Branching Code versus SIMD} -- Both listings select elements from one of \texttt{vec1}, or \texttt{vec2} depending on the \texttt{condition}.
-In the SISD code, only one branch is executed every iteration.
-In the SIMD code, we rely on masking to select which vector to load from, but all instructions are executed at every iteration.}
-\label{fig:branching_code}
-\end{figure}
-
-\textbf{Core Operation Time per Tuple}: The core operation is the same for all strategies.
-The only way to reduce its contribution to the running time is through data-parallel SIMD instructions, which are only available with Full Compute and Selective Compute with SelVecs.
-There are, however, situations where SIMD vectorization provides little gain or even hurts performance.
-Consider the example in \cref{fig:branching_code}.
-This function selects an element from one of two vectors depending on the value in a third boolean vector.
-It corresponds to a relational \texttt{CASE} statement.
-For each loop iteration, the SISD code (\cref{fig:branching_sisd}) only executes one branch.
-The SIMD code (\cref{fig:branching_simd}) executes all branches and relies on SIMD masking to set the correct vector lanes; it ends up executing more instructions that the SISD code, which reduces the benefits data-parallelism.
-In general, core operations benefit most from SIMD vectorization when they only contain straight-line (i.e., no branches), arithmetic, and bitwise instructions.
-
-The explanation above raises three questions that can allow us to determine the optimal strategy for any primitive: What is the input vector's selectivity?
-Can we use data-parallelism to perform the core operations on multiple elements at a time?
-Does the core operation contain instructions that make data-parallelism inefficient (e.g., branching)?
-The next chapter experimentally derives the best strategies according to the answers to these questions.
-
-\chapter{Optimal Strategies}
-\label{chapter3}
-In this chapter, we aim to derive the optimal execution strategy for any given primitive experimentally.
-In doing so, we will provide a decision tree that developers can use to guide the implementation of vectorized primitives.
-
-\begin{table}[t!]
-\centering
-\begin{tabular}{|l|l|l|l|l|}
+\section{Throughput Measurement}
+\subsection{Insert}
+We first measure the throughput (number of transactions executed per time period) of insert operations against the baseline recovery performance on a single table.
+The recovery manager uses only our insert implementation for recovery.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ |c|c|c| } 
  \hline
-\textbf{Strategy Name} & \textbf{Representation} & \textbf{Compute} & \textbf{SIMD}      & \textbf{Compatibility} \\ \hline
-SelVecPartial         & Selection Vectors       & Selective        & None               & Update and Map         \\ \hline
-SelVecManual          & Selection Vectors       & Selective        & Manually Written   & Update and Map         \\ \hline
-BitmapPartial         & Bitmaps                 & Selective        & None               & Update and Map         \\ \hline
-BitmapFull            & Bitmaps                 & Full             & Auto Vectorization & Update and Map         \\ \hline
-BitmapFullManual      & Bitmaps                 & Full             & Manually Written   & Update and Map         \\ \hline
-Full                  & Either                  & Full             & Auto Vectorization & Map                    \\ \hline
-FullManual            & Either                  & Full             & Manully Written    & Map                    \\ \hline
+Statements Ratio & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\ 
+ \hline
+ 100\% Insert & $1.12 \times 10^4$ txns/s & $5.60 \times 10^3$ txns/s\\
+ \hline
+ 50\% Insert, 50\% Select & $1.32 \times 10$ txns/s & $6.60 \times 10^3$ txns/s \\ 
+ \hline
 \end{tabular}
-\caption{\textbf{Implemented Strategies} -- For the SIMD columns, `Auto Vectorization' indicates that we rely on compiler auto vectorization, whereas `Manually written' indicates that we write SIMD code ourselves using intrinsics.
-On the compatibility column, Update operations cannot use both Full Compute and Selection Vectors, which explains the last two entries.}
-\label{tab:strategies}
+\caption{\textbf{Insert throughput measurement setup on a single table}. The table uses an initial table size of $10^6$. Each transaction runs 5 statements. A total of $10^5$ transactions are run.}
+\label{tab:throughput_exp_insert}
+\end{center}
 \end{table}
 
-The chapter is divided into a series of experiments based on the questions raised in \cref{prim_perf_section} to determine the optimal strategy for a given primitive.
-\cref{tab:strategies} summarizes all the strategies we implemented.
-As noted before, there is no Full Compute strategy for Updates that works with SelVecs.
-Update experiments will thus always use a Bitmap for Full Compute; For Map experiments, the representation is irrelevant because Full Compute neither reads nor updates the filter.
-In addition to relying on compiler auto-vectorization, we also experimented with manual vectorization (e.g., with SelVecManual) for straight-line arithmetic and bitwise operations (as in \cref{sldpsection}).
+2x difference. Memory copies.
 
-We run the experiments on an Intel Xeon Platinum 8124M CPU @ 3.00GHz with AVX512 support.
-We use the NoisePage DBMS \cite{noise_page}, with a read-only column store as its storage engine, compiled with clang version 9 \cite{clang}.
+\subsection{Delete}
+We then measure throughput of delete operations against the baseline. The recovery manager uses only our delete implementation for recovery.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ |c|c|c| } 
+ \hline
+Statements Ratio & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\
+ \hline
+ 90\% Insert, 10\% Delete & $1.12 \times 10^4$ txns/s & $1.06 \times 10^4$ txns/s \\ 
+ \hline
+ 80\% Insert, 20\% Delete & $1.11 \times 10^4$ txns/s & $1.07 \times 10^4$ txns/s\\
+  \hline
+ 75\% Insert, 25\% Delete & $1.13 \times 10^4$ txns/s & $1.03 \times 10^4$ txns/s\\
+ \hline
+\end{tabular}
+\caption{\textbf{Delete throughput measurement setup on a single table}. The table uses the same setup in \ref{tab:throughput_exp_insert}.}
+\label{tab:throughput_exp_delete}
+\end{center}
+\end{table}
 
-In each experiment, we synthetically generate the data and manually vary the selectivity of each primitive's input vector from $0.0$ to $1.0$ in increments of $0.05$ to show the impact selectivity has on what the optimal strategy is.
-For example, we set a random $25\%$ of bits in a Bitmap to obtain filter with a selectivity of $0.25$.
-Each experiment executes its primitive on enough vectors to obtain a stable average running time.
-For example, fast primitives are executed in the order of $10^6$ times, whereas the slower ones are executed in the order of $10^4$ or $10^5$ times.
-Map primitives materialize their results on a vector in-memory, but Update primitives only modify the input filter without materializing a new vector.
+\subsection{Update}
+Finally, we measure throughput of delete operations against the baseline. The recovery manager uses only our delete implementation for recovery.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ |c|c|c| } 
+ \hline
+Statements Ratio & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\ 
+ \hline
+ 90\% Insert, 10\% Update & $???$ txns/s & $???$ txns/s\\
+  \hline
+ 75\% Insert, 25\% Update & $???$ txns/s & $5.6 \times 10^3$ txns/s\\
+ \hline
+ 50\% Insert, 50\% Update & $???$ txns/s & $10^4$ txns/s \\ 
+ \hline
+\end{tabular}
+\caption{\textbf{Update throughput measurement setup on a single table}. The table uses the same setup in \ref{tab:throughput_exp_insert}.}
+\label{tab:throughput_exp_update}
+\end{center}
+\end{table}
 
-\section{Non Data-Parallel Core Operations}
-\label{nondataparallel}
-First, we consider core operations that are not data-parallel.
-These are the operations where it is impossible to use SIMD instructions to process multiple tuples at a time.
-We experiment on two kinds of such operations: those that operate on variable-length data (e.g., string operations) and those that use instructions without a SIMD counterpart (e.g., integer division, modulo).
+\section{Throughput Measurement With Indexes}
+In the baseline recovery architecture, the recovery manager implements a custom index update function that is independent from the query execution engine.
+This index update function is largely hard-coded and contains large amounts of memory copy operations.
+The code generation approach, however, use the built-in execution engine for index updates.
+We believe as the number of indexes increases, the performance gap between baseline and the code generation approach will shorten.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ |c|c|c|c|c| } 
+ \hline
+Statements Ratio & \#Indexes & \textbf{Baseline Throughput} & \textbf{Codegen Throughput} & \textbf{Baseline:Codegen} \\ 
+ \hline
+ 100\% Insert & 5 & $3.7 \times 10^4$ txns/s & $2.4 \times 10^4$ txns/s & 1.54\\
+ \hline
+ 100\% Insert & 10 & $2.4 \times 10^4$ txns/s & $1.7 \times 10^4$ txns/s & 1.41\\
+ \hline
+ 100\% Insert & 100 & $3.2 \times 10^3$ txns/s & $2.4 \times 10^3$ txns/s & 1.3\\
+ \hline
+\end{tabular}
+\caption{\textbf{Insert throughput measurement setup on a single table with indexes}. The table uses the same setup in \ref{tab:throughput_exp_insert}.}
+\label{tab:throughput_exp_index_insert}
+\end{center}
+\end{table}
 
-\subsection{Variable Length Data}
-\begin{figure}[t!]
-\captionsetup[subfigure]{justification=justified}
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_string_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE country < `Japan'}}
-  \label{fig:varlen_update}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_string_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT POSITION(`a' in country)}.}
-  \label{fig:varlen_map}
-\end{subfigure}
-\caption{\textbf{Variable Length Data} -- Country names are all short strings that highlight the fact that even cheap variable-length operations should avoid full compute.}
-\label{fig:varlen_map_update}
-\end{figure}
+\section{Throughput Measurement With Constraints}
 
-We first analyze the performance of operations on variable-length data, as is often the case for string operations (e.g., string comparison, sub-string).
-Let us consider the factors in \cref{fig:perf_equation}.
-Here, the core operation executes a variable number of instructions depending, for example, on the length of the input strings.
-As we show in the next experiment, it dominates the iteration logic time, even for short strings.
-Because data parallelism is not available, the core operation time per tuple is the same across all strategies.
-The principal factor that differentiates strategies is, therefore, the number of tuples processed, which favors Selective Compute strategies over Full Compute ones.
-
-To confirm this intuition, we run a micro-benchmark using simple string operations on short strings.
-The Update micro-benchmark performs string comparison, and the Map micro-benchmark performs character location.
-The strings are all short country names (e.g., USA, Senegal, China).
-The purpose is to show that the number of tuples processed is the principal performance factor, even for relatively cheap variable-length operations.
-
-The results are shown in \cref{fig:varlen_map_update}.
-We see that Full Compute (with BitmapFull or Full) consistently performs worst because it processes more tuples, except at full selectivity.
-At full selectivity, Full Compute and Selective Compute process the same number of tuples, but the former slightly benefits from a simpler iteration logic.
-We also see that the performance of SelVecPartial is similar to that of BitmapPartial, despite the former's simpler iteration logic, meaning that the core operation is the dominating cost.
-
-The optimal strategy in this scenario is SelVecPartial.
-It processes fewer tuples than Full Compute strategies and has a slightly cheaper iteration logic than BitmapPartial (though the performance difference is small).
-Full Compute is only marginally competitive at full selectivity; its benefits are mostly negligible; the more expansive the core operation, the worse Full Compute becomes.
-
-\subsection{Integer Division}
-\begin{figure}[t!]
-\captionsetup[subfigure]{justification=justified}
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_division_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 \% col2 < val}.}
-  \label{fig:division_update}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/simple_division_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 / col2}.}
-  \label{fig:division_map}
-\end{subfigure}
-\caption{\textbf{Integer Division.}}
-\label{fig:division_map_update}
-\end{figure}
-Second, we analyze the integer division instruction as it does not have a SIMD counterpart.
-Let us consider the factors in \cref{fig:perf_equation}.
-Integer division is so expansive that the iteration logic time is negligible compared to the core operation time.
-Because data parallelism is not available, its cost per tuple is the same across all strategies.
-Once again, the main factor that differentiates strategies is the number of tuples processed, which favors Selective Compute.
-
-The micro-benchmarks of this section evaluate an Update and a Map primitive that both contain an integer division.
-The results are shown in \cref{fig:division_map_update}.
-The explanation of its results is similar to \cref{fig:varlen_map_update}'s: the cost of the integer division and the unavailability of data-parallelism make Full Compute inefficient.
-SelVecPartial is, once again, the optimal strategy, but only has a slight edge over BitmapPartial due to the former's simpler iteration logic.
-
-We conclude that the number of tuples processed in the dominant performance factor for non-data-parallel primitives, making Full Compute impractical.
-Because of the small impact of iteration logic time, SelVecPartial has a small performance edge over BitmapPartial, but developers can choose either representation without much affecting performance.
-
-\section{Inefficient Data Parallelism}
-\label{inefficientsimd}
-\begin{figure}[t!]
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_and_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \&\& col2 < val2}.}
-  \label{fig:logical_and_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_and_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \&\& col2 < val2}.}
-  \label{fig:logical_and_map}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_or_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 || col2 < val2}.}
-  \label{fig:logical_or_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/logical_or_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 || col2 < val2}.}
-  \label{fig:logical_or_map}
-\end{subfigure}
-
-\caption{\textbf{Branching Operations} -- Each operation contains one branch because of boolean short-circuiting.}
-\label{fig:logical_map_update}
-\end{figure}
-
-\begin{figure}[t!]
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_and_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 \& col2 < val2}.}
-  \label{fig:bitwise_and_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_and_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 \& col2 < val2}.}
-  \label{fig:bitwise_and_map}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_or_update.pdf}
- \caption{Update operation corresponding to \\ \texttt{\footnotesize WHERE col1 < val1 | col2 < val2}.}
-  \label{fig:bitwise_or_update}
-\end{subfigure}%
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/bitwise_or_map.pdf}
- \caption{Map operation corresponding to \\ \texttt{\footnotesize SELECT col1 < val1 | col2 < val2}.}
-  \label{fig:bitwise_or_map}
-\end{subfigure}
-
-\caption{\textbf{Bitwise Operations} -- The increase in performance compared to \cref{fig:logical_map_update} is due to the lack of branching.}
-\label{fig:bitwise_map_update}
-\end{figure}
-
-To study the effect of inefficient use of SIMD instructions, we now consider core operations with branching code.
-\cref{chapter4} will discuss primitives that contain pointer manipulation and several memory accesses.
-SIMD strategies (e.g., BitmapFull) benefit from processing multiple tuples at a time, but suffer from executing more code than SISD strategies (e.g., SelVecPartial).
-Their core operation time per tuple may even decrease depending on the branching structure (e.g., having to execute all branches of a switch statement).
-Thus, if Full Compute can outperform Selective Compute at all, it will do so at the highest selectivities because the small reduction in time spent per tuple cannot compensate for the higher number of tuples processed.
-Among SISD strategies, SelVecPartial will outperform BitmapPartial due to its simple iteration logic.
-
-To determine the impact of a single branch within the core operation, we implemented primitives that perform logical and/or operations.
-Due to boolean short-circuiting, these primitives will contain exactly one branch.
-We have confirmed that the compiler manages to auto-vectorize Full Compute strategies.
-The results are shown in \cref{fig:logical_map_update}.
-We can see that Full Compute is either always the worst strategy (e.g., in \cref{fig:logical_or_update}), or only competitive at high selectivities ($>= 0.85$ in \cref{fig:logical_or_map}).
-To confirm that branching is indeed responsible for the poor performance of Full Compute, we implemented branch-less primitives with bitwise and/or operations rather than logical ones.
-The results are in \cref{fig:bitwise_map_update}.
-Full Compute now outperforms Selective Compute for selectivities $<= 0.2$, confirming that branching code significantly hurts SIMD strategies' performance.
-
-In conclusion, though Full Compute can sometimes be competitive at the highest selectivities, we recommend, for the sake of simplicity, the SelVecPartial strategy.
-
-\section{Straight-Line and Data-Parallel Core Operations}
-\label{sldpsection}
-
-We now consider core operations with Straight-line and data-parallel (SLDP) code, i.e., non-branching code that can efficiently leverage SIMD instructions.
-Primitives that only perform arithmetic (without integer division) and bitwise instructions fall under this category.
-Let us consider the equation in \cref{prim_perf_section} to analyze the performance of each strategy in this scenario.
-Full Compute strategies (e.g., BitmapFull, Full) can leverage SIMD instructions to significantly reduce the iteration logic and core operation time per tuple.
-AVX512 \cite{intel2019} integer addition instructions, for example, performs 8 parallel 64-bit additions in a single cycle.
-We, therefore, expect Selective Compute strategies to only be competitive when the selectivity is very low, meaning that Full Compute processes far more tuples.
-SelVecManual is the exception: it is a Selective Compute strategy that uses SIMD instructions.
-Its gain in iteration logic time is, however, not as high as that of Full Compute strategies because it uses a \texttt{gather} instruction \cite{intel2019} to collect the elements at the selected indices (shown in \cref{fig:rationale_selvec}).
-The higher the core operations time per tuple, the less important the \texttt{gather} overhead because iteration logic time becomes more and more insignificant.
-Thus, we expect SelVecManual to be competitive with Full Compute at medium or below selectivities depending on the core operation.
-The next experiments will quantify the differences between each strategy.
-
-\begin{figure}[t!]
-    \centering
-    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_update.pdf}}
-    \caption{\textbf{SNBS Update Operations} -- For Updates, SelVecManual relies on the \texttt{compress\_store} SIMD instruction rather than on \texttt{scatter}, which reduces its overhead.}
-    \label{fig:snbs_update}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \makebox[\textwidth]{\includegraphics[width=.9\paperwidth]{eval/snbs_map.pdf}}
-    \caption{\textbf{SNBS Map Operations} -- For Maps, SelVecManual relies on the \texttt{scatter} SIMD instruction, which explains why it initially does not perform better than SelVecPartial.}
-    \label{fig:snbs_map}
-\end{figure}
-
-\begin{figure}[t!]
-\captionsetup[subfigure]{justification=justified}
-\centering
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/snbs_thres_update.pdf}
- \caption{Update thresholds between BitmapFullManual and Selective Compute strategies.}.
-  \label{fig:snbs_thres_update}
-\end{subfigure}
-\begin{subfigure}[t]{.49\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/snbs_thres_map.pdf}
- \caption{Map thresholds between FullManual and Selective Compute strategies.}
-  \label{fig:snbs_thres_map}
-\end{subfigure}
-\caption{\textbf{Mixed Compute Thresholds} -- The Full Compute strategies are BitmapFullManual and FullManual because compiler auto-vectorization does not consistently use AVX512 registers.}
-\label{fig:snbs_thres}
-\end{figure}
-
-We implemented Update and Map primitives in which the core operation contains single-cycle arithmetic and bitwise instructions.
-We progressively increase the number of such instructions to increase the core operations time per tuple and quantify the selectivity thresholds above which Full Compute strategies become faster than Selective Compute ones.
-The results are shown in \cref{fig:snbs_update,fig:snbs_map} (for Updates and Maps).
-The selectivity thresholds for each graph are respectively in \cref{fig:snbs_thres_update,fig:snbs_thres_map}.
-Let us analyze the results in detail:
-\begin{itemize}
-    \item In all experiments, SelVecPartial performs better than BitmapPartial. As explained in the previous chapter, this is only due to a simpler iteration logic.
-    \item SelVecManual is always the best Selective Compute strategy because it uses data-parallelism to reduce the time spent per tuple.
-    \item Manual Vectorization (with FullManual and BitmapFullManual) sometimes performs much better than Auto-Vectorization (with Full and BitmapFull).
-For example, in \cref{fig:snbs_update}'s graph for \texttt{NumOps=13}, BitmapFullManual is $1.8$ times faster than BitmapFull.
-With further investigation of the generated code, we discovered that the compiler is overly conservative when it comes to using AVX512.
-AVX512 registers result in decreased CPU frequency \cite{avx512_registers}, so the compiler is careful not always to use them.
-It often uses AVX2 registers instead.
-We, on the other hand, always use AVX512.
-    \item The thresholds for SelVecPartial and BitmapPartial, shown in \cref{fig:snbs_thres_update,fig:snbs_thres_map}, prove that these strategies are only competitive with Full Compute at low selectivities ($<= 0.2$) because, as explained above, they do not use SIMD instructions.
-    \item The threshold for SelVecManual, on the other hand, increases with the number of operations because its iteration logic overhead becomes more and more insignificant as the core operation time increases.
-SelVecManual thus benefits from data-parallelism while processing fewer tuples than Full Compute strategies.
-The iteration logic difference is never entirely negligible, though; the threshold is medium at best ($0.5$ at 15 operations in \cref{fig:snbs_thres_update}).
-\end{itemize}
-
-This analysis shows the importance of Mixed Compute: we should use Selective Compute below the selectivity thresholds, and Full Compute otherwise.
-The next section details our implementation of Mixed Compute.
-
-\section{Implementing Mixed Compute}
-\label{mixedcompute}
-The previous section showed that for a given SLDP primitive, there is a threshold below which SelVecManual is optimal and above which BitmapFullManual and FullManual are optimal for Updates and Maps respectively.
-To find this threshold, we can run a micro-benchmark similar to the one in \cref{fig:snbs_thres}.
-The lowest selectivity at which the runtime of Full Compute strategies exceeds that of Selective Compute strategies represents the Mixed Compute threshold for a given primitive.
-In summary, the optimal SLDP strategy is as follows:
-\begin{itemize}
-\item For Maps: Full or FullManual when the selectivity is above the threshold, and SelVecManual when the selectivity is below the threshold.
-\item For Updates: BitmapFull or BitmapFullManual when the selectivity is above the threshold, and SelVecManual when the selectivity is below the threshold.
-Note that the BitmapFull and SelVecManual have different filter representations.
-There is a small cost associated with the conversion from Bitmap to SelVec, but, in practice, other operations always follow an update (e.g., there is a projection after a scan filter).
-The subsequent operations often amortize the conversion cost.
-For small queries, in which the amortization does not neutralize the conversion cost, BitmapPartial can slightly outperform SelVecManual because it maintains the Bitmap representation.
-Nonetheless, as we show in \cref{chapter4}, its gains are mostly negligible.
-\end{itemize}
-
-\section{Vectorization Decision Tree}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/DecisionTree.png}
-    \caption{\textbf{The Summary Decision Tree} -- It summarizes the results in this chapter.}
-    \label{fig:filter_decision_tree}
-\end{figure}
-
-The decision tree in \cref{fig:filter_decision_tree} summarizes the results of this chapter.
-It uses the experiments above to derive the optimal implementation strategy depending on the core operation and the input selectivity of a given primitive.
-The \textbf{Data-Parallel} node, the \textbf{Effecient Data-Parallelism} respectively use the results in \cref{nondataparallel} and \cref{inefficientsimd} that designated SelVecPartial as the optimal strategy.
-The \textbf{Operation Type} and \textbf{Above Threshold} nodes use the results in \cref{sldpsection} and \cref{mixedcompute} that designed a Mixed Compute strategies for each primitive type (Update or Map).
-
-The tree assumes that the system developer implements all representations and compute strategies.
-The total number of variants of each primitive will be extensive, which constitutes a costly engineering endeavor.
-To remedy this cost, we note that the BitmapPartial strategy does not perform significantly worse than SelVecPartial or SelVecManual at most selectivities.
-Developers can, thus, focus on focus on the Bitmap representation without losing too much performance.
-Switching between Full and Selective Compute is critical, though; the performance gain is too significant.
-
-\chapter{Experimental Evaluation}
-\label{chapter4}
-
-We now evaluate the benefits of the decision tree in \cref{chapter3}.
-For this evaluation, we implemented the strategies for a subset of queries from the TPCH benchmark, a decision support system workload that simulates an OLAP environment~\cite{tpch}.
-We use a scale factor of 10 ($\sim$10~GB).
-The hardware setup is unchanged from the previous chapter.
-We use the columnar storage engine of the NoisePage DBMS \cite{noise_page}.
-The string columns are all dictionary compressed for faster processing \cite{abadicompression}.
-Whenever the compiler (clang \cite{clang}) failed to SIMD vectorize a primitive that should be data-parallel or used AVX2 instead of AVX512, we hand-wrote SIMD code using AVX512 intrinsics \cite{intel2019}.
-
-Each query contains primitive with side-effects (e.g., hash table insertion), and primitives without side-effects (e.g., multiplication of two columns).
-Given that this thesis focuses on side-effect-free primitives, we report two metrics for each query: the running time of the whole query and the running time of the side-effect free primitives within the query.
-The output of each query is materialized in memory and printed to the standard output; no network transfer occurs.
-
-\section{Q1: High Selectivity SLDP primitives}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/TPCH-Q1.png}
-    \caption{\textbf{The TPCH Q1 query.}}
-    \label{fig:tpch_q1}
-\end{figure}
-
-\label{q1section}
-\begin{figure}[t!]
-\centering
-\hspace*{\fill}% 
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/q1_optimized.pdf}
- \caption{Running time of side-effect free primitives.} \label{fig:q1_optimized}
-\end{subfigure}
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/q1_bench.pdf}
- \caption{Total running time.}
-  \label{fig:q1_bench}
-\end{subfigure}%
-\caption{\textbf{Q1 Performance}}
-\label{fig:q1_perf}
-\end{figure}
-
-\cref{fig:tpch_q1} shows the SQL code of the Q1 query.
-It contains a single SLDP filter in its \texttt{WHERE} clause; its selectivity is consistently above $0.95$ for the vectors from the \texttt{lineitem} table.
-Its \texttt{SELECT} clause contains multiple arithmetic SLDP operations followed by side-effect-full aggregation and order-by operators that dominate the running time.
-Before the aggregation, there is an SLDP primitive that hashes the \texttt{GROUP BY} keys.
-Because the selectivity is high and the query is SLDP-heavy, our decision tree favors SIMD vectorized Full Compute strategies.
-Besides, switching between strategies is not required.
-
-The results for the strategies we implemented are shown in \cref{fig:q1_perf}.
-We omit the Mixed strategies' measurements because no switching ever occurs in the query plan; they have the same running times as BitmapFull.
-\cref{fig:q1_optimized} shows the running time of the primitives we optimized.
-As predicted by our decision tree, the SIMD Full Compute strategy, BitmapFull, best optimizes these primitives.
-It outperforms BitmapPartial and SelVecPartial (which have SISD primitives) by $1.6\times$, and SelVecManual (which has manually written SIMD primitives) by $1.3\times$.
-
-The total running times (\cref{fig:q1_bench}) are mostly the same for all queries for two reasons.
-First, the side-effect-full aggregation, which we do not try to optimize, is the query's dominant component.
-Second, Intel's implementation of AVX512 instructions decreased CPU speed.
-This effect explains why clang limits their usage to avoid performance degradations \cite{avx512_registers}.
-Compare, for example, the BitmapPartial and BitmapFull strategies.
-They both have the same code for the side-effect-full primitives, but BitmapFull performs them in $838\text{ms}$, whereas BitmapPartial performs them in $739\text{ms}$.
-Naively avoiding AVX512 instructions is not the solution; AVX2 uses vectors that have half the width of AVX512's, which slows down the execution of SLDP primitives.
-Careful analysis is thus required to balance the benefits and disadvantages of AVX512 instructions.
-This analysis is left as future work.
-
-\section{Q6: Mixed Selectivity SLDP Primitives}
-\label{q6section}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/TPCH-Q6.png}
-    \caption{\textbf{The TPCH Q6 query.}}
-    \label{fig:tpch_q6}
-\end{figure}
-
-\begin{figure}[t!]
-\centering
-\hspace*{\fill}% 
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/q6_optimized.pdf}
- \caption{Running time of side-effect free primitives.}
-  \label{fig:q6_optimized}
-\end{subfigure}
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/q6_bench.pdf}
- \caption{Total running time.}
-  \label{fig:q6_bench}
-\end{subfigure}%
-
-\caption{\textbf{Q6 Performance}. The \texttt{+} notation indicates mixed strategies.}
-\label{fig:q6_perf}
-\end{figure}
-
-\cref{fig:tpch_q6} shows the SQL code of the Q6 query.
-Its \texttt{WHERE} contains five SLDP filters, and its \texttt{SELECT} one arithmetic SLDP projection followed by an aggregation.
-For all vectors, the second filter leads to a selectivity below $0.15$, triggering a threshold-based switch.
-Thus, our decision tree favors a mixed strategy: BitmapFull at high selectivities, and SelVecManual at low selectivities.
-
-The results are shown in \cref{fig:q6_perf}.
-\cref{fig:q6_optimized} shows the running time of the primitives we optimized.
-This graph confirms the necessity of the caveat we raised in \cref{mixedcompute}: the cost of converting from Bitmaps to SelVecs makes BitmapFull+SelVecManual slightly more expansive than BitmapFull+BitmapPartial by $0.3\%$, even though SelVecManual performs better than BitmapPartial on the individual primitives in this query.
-This difference is insignificant, so our decision tree does not take it into account.
-The BitmapFull+SelVecPartial strategy also suffers from the conversion overhead, but the difference only amounts to $2\%$.
-All mixed strategies perform better than non-mixed ones.
-BitmapFull+BitmapPartial and BitmapFull+SelVecManual are faster than BitmapFull by $1.3\times$ because Full Compute becomes wasteful after the selectivity drops below $0.15$.
-The SIMD SelVecManual strategy is slower by $1.04\times$ because it is only suboptimal in the first two filters of the \texttt{WHERE} clause.
-The SISD Selective Compute strategies BitmapPartial and SelVec Partial are slower by $1.6\times$ and $1.4\times$ because they do not take advantage SIMD instructions for SLDP operations.
-
-\cref{fig:q6_bench} shows the total running time that includes the side-effect-full aggregation primitive.
-This primitive only accounts for a small portion of the running time, so we do not observe the slowdown caused by AVX512 registers.
-In SLDP-dominated queries like Q6, the benefits of AVX512 outweigh its disadvantages.
-
-\section{Q4: Low Selectivity, Inefficient Data Parallelism}
-\label{q4section}
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.5]{images/TPCH-Q4.png}
-    \caption{\textbf{The TPCH Q4 query.}}
-    \label{fig:tpch_q4}
-\end{figure}
-
-\begin{figure}[t!]
-\centering
-\hspace*{\fill}% 
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/q4_optimized.pdf}
- \caption{Running time of side-effect free primitives.}
-  \label{fig:q4_optimized}
-\end{subfigure}
-\hspace*{\fill}%
-\vspace*{8pt}%
-
-\hspace*{\fill}%
-
-\begin{subfigure}[t]{.8\linewidth}
- \centering
- \includegraphics[width=0.9\linewidth]{eval/q4_bench.pdf}
- \caption{Total running time.}
-  \label{fig:q4_bench}
-\end{subfigure}%
-\caption{\textbf{Q4 Performance}}
-\label{fig:q4_perf}
-\end{figure}
-
-\cref{fig:tpch_q4} shows the SQL code of the Q4 query.
-It is a join of two tables (\texttt{lineitem} and \texttt{orders}) followed an aggregation and an order-by operator.
-The join's left side (from the \texttt{orders} table) has two SLDP filters that reduce the selectivity to $0.3$ and $0.1$, respectively, and an SLDP hashing primitive before insertion into the join hash table.
-Its right side (from the \texttt{lineitem} table) has a filter with a selectivity $>0.6$ and a join probe that contains the multiple primitives shown in \cref{fig:probe_example}.
-Although the probe's hashing primitive is an SLDP operation, its other primitives contain multiple indirect lookups (e.g., accessing hash table entries or the keys within those entries for exact comparison).
-Because these operations are called in a loop to find all potential matches during the join, they constitute the bulk of the running time.
-Unfortunately, their data-parallel versions are inefficient.
-For example, the comparison primitive, which we manually implemented for the SelVecManual strategy, contains three \texttt{gather} instructions: one to collect the keys from the right side of the join, one to collect the hash table entries, and one to collect the keys within the hash table entries for comparison.
-SIMD instructions can cause performance degradations in these primitives.
-Thus, we predict that we should use BitmapFull to perform the SLDP filters and hashing, then switch to using SelVecPartial, not the SIMD SelVecManual, for the complex primitives within the join probe.
-
-The results are shown in \cref{fig:q4_perf}.
-\cref{fig:q4_optimized} shows the running time of the primitives we optimized.
-Our predicted strategy, BitmapFull+SelVecPartial, is indeed the optimal one, closely followed by BitmapFull+BitmapPartial, which also switches from SIMD to SISD code for complex primitives.
-Notice the performance degradation caused by SIMD vectorization, as mentioned in \cref{inefficientsimd}.
-Unlike the previous queries' results, BitmapFull+SelVecPartial performs better than the BitmapFull+SelVecManual by $1.1\times$ because the latter relies on SIMD vectorization for complex primitives.
-BitmapFull is also slightly worse (by $1.04\times$).
-Its performance degradation is attenuated by the fact that it requires fewer \texttt{gather} instructions than SelVecManual (e.g., one instead of three in the comparison primitive), relying instead on faster \texttt{vector\_load} instructions to load all elements in a vector, whether they are selected or not.
-The SISD Selective Compute strategies BitmapPartial and SelVecPartial are, respectively, $1.1\times$ and $1.05\times$ worse because they do not optimize SLDP primitives.
-
-The results in \cref{fig:q4_optimized} mostly reflect those in \cref{fig:q1_optimized}.
-AVX512 registers, once again, cause a slowdown in SIMD implementations, causing, for example, BitmapFull to become slower than BitmapPartial even though they share the same code for side-effect-full primitives.
-Here, the disadvantages of AVX512 outweigh its benefits on the full query.
-
-\section{Summary}
-Our analysis shows that our results in \cref{chapter3} provide us with the best implementation for each query.
-The downside is that it requires a wide variety of variants for each primitive because any strategy can be optimal under specific scenarios.
-We see, however, that Bitmap only strategies (e.g., BitmapFull, or BitmapFull+BitmapPartial) perform nearly as well as the optimal ones in the queries above.
-Developers can thus discard the SelVec representation to save engineering effort without losing too much performance.
-
-\chapter{Related Works}
-\label{chapter5}
-\section{Optimizing Analytical Queries for DBMSs}
-Besides the Vectorization and Materialization model introduced in \cref{chapter1}, much work has been dedicated to optimizing analytical queries on column store DBMSs.
-\citeauthor{materializationstrats} studied the effect of materialization strategies (i.e.
-late or early materialization) on query performance \cite{materializationstrats}.
-All experiments in this paper use late materialization and rely on filters to keep track of valid tuples.
-Nonetheless, \citeauthor{materializationstrats} found that the optimal materialization strategy depends on selectivity and on the exact operations performed on tuples \cite{materializationstrats}.
-\citeauthor{cstore} studied compression as way to optimize both data storage and query execution in column stores \cite{cstore}.
-\citeauthor{abadicompression} performed an extensive evaluation of various compressions schemes \cite{abadicompression}.
-The experiments in \cref{chapter4} rely only on dictionary compression.
-
-Researchers have also used SIMD vectorization to optimize relational algorithms.
-\citeauthor{simd1} optimized scans and nested loop joins using techniques most similar to our BitmapFullManual implementation.
-\citeauthor{simd2} further applied SIMD vectorization hash tables probes \cite{simd2}.
-\citeauthor{simdsort1} and \citeauthor{simdsort2} proposed sorting algorithms using SIMD instructions \cite{simdsort1,simdsort2}.
-The relational order-by operators, sort-merge joins, and sorting aggregations can all use these efficient sorting algorithms.
-
-\citeauthor{ibm_blu} and \citeauthor{simdscan} combined this body of work and built full commercial DBMSs using column stores with compression and SIMD vectorization techniques to optimize table scans \cite{ibm_blu,simdscan}.
-\citeauthor{orestis_bitmap} further built a system using SIMD vectorization on all relational operators \cite{orestis_bitmap}.
-These systems make a static filter representation decision: Bitmaps for table scans, and SelVecs for other operations.
-
-Along with the Vectorization Model, the current state-of-the-art execution technique is the Data-Centric Query Compilation technique pioneered by \citeauthor{hyper_llvm} \cite{hyper_llvm}.
-This technique splits a given query plan into several pipelines and compiles them individually.
-A pipeline is a sequence of relational operators that do not materialize their result set followed by a pipeline-breaker -- an operator that does materialize its result set (e.g., to build a hash table or sort tuples).
-This approach's benefits are: (1) eliminate interpretation overhead by compiling specialized code and (2) maximize cache locality by keeping tuples in the registers or the L1 cache for entire pipelines.
-Our system uses the relaxed operator fusion technique \cite{rof}, which combines data-centric code generation with the Vectorization Model.
-It generates code that operates on vectors of tuples, thereby eliminating the interpretation overhead and optimizing cache locality while maintaining all the Vectorization Model's benefits.
-
-\section{Primitive Optimization}
-There has been previous work dedicated to optimizing primitives in the Vectorization Model.
-They mostly differ from this thesis in that they do not consider the impact of filter representation in their performance.
-We now discuss this work and how they relate to our proposed method.
-
-\citeauthor{sompolski_vec} compared the cost of vectorization to data-centric query compilation \cite{sompolski_vec} and combined the two approaches to generate new vectorized primitive at runtime, which is the method we use in \cref{sldpsection}.
-Using only the SelVec representation, the authors compare vectorized primitives' performance under various compute strategies (e.g., Full versus Selective Compute, Conditional versus Unconditional Store).
-They, however, do not provide a way to switch between these strategies.
-
-\citeauthor{everything_vectorized} performs a similar analysis for whole queries in addition to individual primitives \cite{everything_vectorized}.
-They developed a SIMD implementation of primitives using Selection Vectors, just like our SelVecManual strategy.
-Like ours, their implementation was effective on compute-heavy queries (e.g., Q6 in \cref{q6section}) and ineffective on memory access-heavy queries (e.g., Q4 in \cref{q4section}).
-
-\citeauthor{miro_adapt} recognized the importance of switching compute strategy at run-time \cite{miro_adapt}.
-For each operation, the authors implement several \textit{flavors} (i.e., different ways to perform the same operations).
-They then use reinforcement learning (RL) to switch between the best \textit{flavors} dynamically at runtime.
-This approach is more general than our micro-benchmarks because it takes into account changes in system load.
-SelVec is the only representation considered, limiting the range of adaptivity on Update tasks.
-Besides, our decision tree provides more explainability than RL techniques.
-As future work, we could employ a similarly dynamic strategy to determine our thresholds.
-
+\chapter{Related Work}
 \chapter{Conclusion}
-\label{chapter6}
-This work analyzed the impact of filter representation (i.e., Bitmap vs.
-SelVec) and compute strategy (i.e., Full vs.
-Selective Compute) on the performance of the vectorized primitives in an in-memory analytical DBMS.
-We identified the factors that influence performance: number of tuples processed, iteration logic, and core operation time per tuple.
-We explained how each combination of representation and compute strategy balances between these three factors.
-Full Compute has the cheapest iteration logic, processes all tuples, but spends less time on each tuple when SIMD vectorization is possible.
-Full Compute is, however, only available with Bitmaps on Update primitives.
-Selective Compute with SelVecs has a cheaper iteration logic than Selective Compute with Bitmaps, and is more amenable to SIMD vectorization.
-We confirmed these observations with several micro-benchmarks, and created a decision tree to help systems developers find the optimal implementation of arbitrary side-effect-free primitives.
-Finally, we showcased the benefits of our analysis on OLAP queries with multiple primitives and consistently achieved the best performance.
-Our performance gains over the best techniques that do not adapt filter representation and compute strategy can be up to $1.3\times$.
 

### 2021-07-27 paper.tex
Updates from Overleaf

@@
 \begin{abstract}
-Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records.
-We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
+Code generation is a widely-used technique for improving query execution throughput by compiling instructions into native code.
+This technique, however, leads to design challenges for the recovery system of a database management system.
+The log replay process will be disjoint from the built-in execution engine that has been adapted to operate efficiently on compiled code.
+This usually leads to the implementation of a separate "execution engine" to deal with the execution of log records, which can be a huge waste of engineering efforts.
+To resolve this design conflict between code generation and database recovery, we propose a code generation design for the recovery system.
+We ask the recovery system to convert log records into compiled code that will be easily accepted by the execution engine.
+Our results show that while our approach creates more performance overhead compared to using a separate execution engine, it also introduces much less engineering overhead and is superior in aspects such as index updates and constraint checking.
 \end{abstract}
@@
 \includegraphics[width=0.9\textwidth]{images/WAL.png}
-\caption{Logging Architecture. Log manager uses the log serializer and log consumers to convert redo buffers into log records.}
+\caption{\textbf{Logging Architecture}. Log manager uses the log serializer and log consumers to convert redo buffers into log records.}
 \label{fig:wal}
@@
 This mapping is used to correctly copy the values out of each redo record into a physical plan node.
+\subsubsection{Catalog Access}
+Retrieving access to a database catalog is costly.
+Profiling results show that a single operation to retrieve catalog access can account for 10\% of total execution time.
+Therefore, a mapping is used to keep track of catalog accessors.
 
@@
 \subsection{Insert}
-We first measure the throughput (number of transactions executed per time period) of insert operations against the baseline recovery performance on a single table.
-The recovery manager uses only our insert implementation for recovery.
 \begin{table}[H]
@@
  \hline
- 100\% Insert & $1.12 \times 10^4$ txns/s & $5.60 \times 10^3$ txns/s\\
+ 100\% Insert & $1.12 \times 10^4$ txns/s & $6.00 \times 10^3$ txns/s\\
  \hline
@@
 \end{table}
+We first measure the throughput (number of transactions executed per time period) of insert operations against the baseline recovery performance on a single table.
+The recovery manager uses only our insert implementation for recovery.
 
-2x difference. Memory copies.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ |c|c|c| } 
+ \hline
+\textbf{IR} & \textbf{Percentage} & \textbf{Functionality} \\ 
+ \hline
+79.35 & \textbf{81.7\%} & Query Execution (Op Codes)\\
+ \hline
+ 9.35 & 9.62\% & Execution Context Initialization \\ 
+ \hline
+ 6.19 & 6.37\% & Retrieving Redo Values \\
+ \hline
+ 2.26 & 2.32 \% & Metadata  Cache Map Access \\
+ \hline
+ 0.03 & 0.03 \% & Compile \\
+ \hline
+\end{tabular}
+
+\begin{tabular}{ |c|c|c| } 
+ \hline
+\textbf{IR} & \textbf{Percentage} & \textbf{Operation Code}\\ 
+ \hline
+29.88 & \textbf{74.4\%} & \texttt{StorageInterfaceInit}\\
+ \hline
+ 6.00 & 14.9\% & \texttt{StorageInterfaceTableInsert} \\ 
+ \hline
+ 2.76 & 6.87\% & \texttt{StorageInterfaceFree} \\
+ \hline
+ 1.40 & 3.48\% & \texttt{StorageInterfaceGetTablePR} \\
+ \hline
+ 0.12 & 0.35\% & \texttt{ExecOUFeatureVectorReset} \\
+ \hline
+\end{tabular}
+\caption{\textbf{KCachegrind Analysis on Insert}. IR refers to the number of instructions executed in total by the selected function during a single recovery step.}
+\label{tab:throughput_exp_insert_overhead_all}
+\end{center}
+\end{table}
+Results show that the code generation throughput is 47\% slower than the baseline implementation.
+Profiling result \ref{tab:throughput_exp_insert_overhead_all} explains that 81.7\% of the slow down comes from the execution of the compiled query.
+This is unexpected, as we imangined that most of the overhead would come from memory copies for extracting values, but those memory copy overhead only accounts for 6\%.
+
+A further dive into the composition of query execution shows that 74.4\% of the overhead comes from \texttt{StorageInterfaceInit}, which is a function that initializes a storage interface.
+The storage interface stores objects that allow manipulation over a certain DBMS table.
+These objects can include the execution context, redo deltas and index objects associated with table.
+The storage interface is reconstructed every time an execution of a query begins, and is destructed after the query has been executed.
 
@@
 \end{table}
+
+Performance of delete is closer to the baseline than insert, with code generation performance aroudn 90\% to 95\% of baseline performance.
+The performance of delete is close to our expectation, as delete does not require any memory copies from a redo record like an insert does.
+Executing a delete is also a lightweight operation, as we have prevented the pipeline from performing extra work as shown in Section 3.2.2.
 
@@
  \hline
- 100\% Insert & 5 & $3.7 \times 10^4$ txns/s & $2.4 \times 10^4$ txns/s & 1.54\\
+ 100\% Insert & 5 & $3.97 \times 10^4$ txns/s & $2.65 \times 10^4$ txns/s & 1.50\\
  \hline
- 100\% Insert & 10 & $2.4 \times 10^4$ txns/s & $1.7 \times 10^4$ txns/s & 1.41\\
+ 100\% Insert & 10 & $2.37 \times 10^4$ txns/s & $1.7 \times 10^4$ txns/s & 1.40\\
  \hline
- 100\% Insert & 100 & $3.2 \times 10^3$ txns/s & $2.4 \times 10^3$ txns/s & 1.3\\
+ 100\% Insert & 100 & $3.30 \times 10^3$ txns/s & $2.49 \times 10^3$ txns/s & 1.32\\
  \hline

### 2021-08-03 paper.tex
Updates from Overleaf

--- 2021-07-26 paper.tex
@@
 % some useful packages
-\usepackage{times}
+\usepackage{libertine}
 \usepackage{fullpage}
 \usepackage{graphicx}
-\usepackage{amsmath}
+\usepackage{amsmath, lipsum, afterpage}
 \usepackage[numbers,sort]{natbib}
@@
 \usepackage{subcaption}
+\usepackage{booktabs}
+\usepackage{caption}
 
@@
 \usepackage[scaled]{inconsolata}
+
+\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={red}}
 
@@
 \title{ %% {\it \huge Thesis Proposal}\\
-{\bf Code Generation Log Replay for In-memory Database Management Systems}}
+{\bf Code Generation Recovery Log Replay for In-Memory Database Management Systems}}
 \author{Tianlei Pan}
@@
 
-\keywords{Code Generation, Log Replay, Recovery, Query Compilation}
+\keywords{Code Generation, Log Replay, Recovery, query compilation}
 
@@
 \begin{abstract}
-Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records.
-We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
+Code generation is an optimization technique for improving query execution throughput by compiling query plans into native code.
+This technique, however, leads to design challenges for the recovery system of a database management system.
+The log replay process will be disjoint from the built-in execution engine that has adapted to operate efficiently on compiled code.
+This leads to the implementation of a separate execution engine to deal with the execution of log records, which can be a huge waste of engineering efforts.
+To resolve this design conflict between code generation and database recovery, we present a unified approach to support both query execution and log replay in a code-generation-based DBMS.
+We ask the recovery system to convert log records into compiled code that will be easily accepted by the execution engine.
+Our results show that while our approach incurs a higher performance overhead compared to using a separate execution engine, it requires much less engineering effort and is potentially superior in aspects such as index updates and constraint checking.
 \end{abstract}
@@
 \chapter{Introduction}
-A \textit{database management system (DBMS)} is a software that is responsible for storing data, analyzing data, and interacting with applications.
-Depending on the use case, a DBMS can be designed to focus more on either capturing data, or analyzing data.
-\textit{On-line Analytical Processing (OLAP)} applications focus on reading, analyzing and aggregating cold data that is less likely to be modified.
-On the other hand, \textit{On-line Transaction Processing (OLTP)} applications work with write-heavy transactions that modify the database frequently.
+Database management system (DBMS) is a category of software that are responsible for storing data, analyzing data, and interacting with applications.
+In 1970, E.F.
+Codd proposed the relational model of data,
 
-Consequently, recovery and durability in OLTP applications are more of a pressing issue compared to OLAP systems.
-Naive or manual backups are usually sufficient for read-only OLAP applications.
-However, since OLTP databases are being updated by transactions constantly, naive backups will either cause the database to lose significant amount of data, or fail to preserve the \textit{ACID (atomicity, consistency, isolation, and durability)} properties.
-This requires OLTP applications to develop specialized recovery mechanism to handle these problems.
-Commercial DBMS often combines Write-ahead Logging (WAL) with checkpoints to preserve database durability.
+Depending on the use case, a DBMS can focus more on either capturing data or analyzing data.
+On-line Analytical Processing (OLAP) DBMSs focus on reading, analyzing and aggregating data that is less likely to be modified.
+On the other hand, On-line Transaction Processing (OLTP) DBMSs support write-heavy transactions that modify the database frequently~\cite{conn05}.
 
-In addition, OLTP systems also require specialized algorithms and implementations to execute write transactions.
-Query compilation is an important optimization technique used in OLTP systems to greatly reduce the number of instructions that need to be executed.
-A DBMS uses a processing model (e.g.
-Iterator, Vectorization, Materialization) to determine how the DBMS executes a query plan.
-The processing model is responsible for translating a query into an intermediate expression.
-The execution engine then breaks down the expression into various tasks for execution.
-To speed up this process, the DBMS can use an optimization technique known as \textit{Query Compilation}.
-Query compilation requires the DBMS to generate specialized code for execution tasks to increase execution throughput.
-These specialized code are often created using an off-shelf compiler into a C/C++ program that implements the functionality of a certain query.
+Regardless of the type of a DBMS, both OLTP and OLAP DBMSs are susceptible to failures.
+The DBMS may shutdown unexpectedly, fail to execute a query, or cease to function due to corrupted data.
+These failures threaten the integrity of the database and make the DBMS unreliable.
+Therefore, it is important for a DBMS to distinguish between different kinds of failures in order to develop mechanisms to maintain integrity.
 
-While query compilation has proved successful in increasing the efficiency of a DBMS, it comes with several drawbacks.
-Implementing a query compilation system can be quite an ordeal.
-It requires additional knowledge of compiler systems (e.g.
-LLVM) and huge amounts of engineering work to translate different execution tasks.
-Moreover, the DBMS also need to modify other components to accommodate with an execution engine reliant on query compilation.
+A DBMS can encounter three types of failures: Transaction Failures, System Failures, and Media Failures\cite{theo83}.
+Transaction Failures are the most common type of DBMS failures.
+Transaction Failures occur when a transaction fails to commit, either at its own request (e.g., logical errors) or on behalf of the DBMS (e.g., resource unavailability).
+When a transaction fails, the system needs to react within its lifetime to prevent data inconsistency~\cite{aries}.
+The DBMS needs to decide whether and how it should undo the changes of the failing transaction.
+Since a DBMS is running transactions rapidly while its active, Transaction Failures can occur as much as 100 times per minute\cite{theo83}.
+System Failures occur due to hardware failures (e.g., power outage), operating system faults (e.g., insufficient memory) or DBMS exceptions.
+Each of these trigger events can cause the DBMS to shutdown unexpectedly and uncontrollably.
+System Failures occur less frequently compared to Transaction Failures, but take a much longer time for the DBMS to recover.
+The database needs to pinpoint time of failure and what changes the DBMS failed to record.
+Media Failures~\cite{sauer17} happen when the underlying storage device of a DBMS fails (e.g., disk head crash, bad sectors).
+This type of failure causes permanent data loss to the DBMS.
+The only way for a DBMS to recover from Media Failures is to restore from a secondary backup storage.
 
-One of such components is the recovery system.
-Since the execution engine has adapted to executing compiled native code, the DBMS needs to implement an additional execution engine to interpret the log records.
-Instead, we propose a system design that will extend the query execution engine to create custom programs for replaying log records.
+To handle different types of failures, a DBMS needs to develop a recovery system that will restart the database correctly in case of failure.
+A widely adopted recovery system is to combine write-ahead logging during transaction execution with crash recovery that uses redo (install changes to the database) and undo (remove changes from the database) processes, represented by ARIES\cite{malviya14}.
+The ARIES (Algorithm for Recovery and Isolation Exploiting Semantics)~\cite{aries} protocol is a recovery method developed by IBM in the 1990s.
+It guarantees database integrity in the fact of Transaction, System and Media Failures~\cite{aries}.
+In ARIES, the DBMS records transaction modifications in durable log records before the DBMS propagates changes to a database page to disk.
+During crash recovery, the DBMS applies changes from the log records to the database with redo and undo processes (i.e., log replay) ~\cite{franklin97}.
+
+However, the original ARIES paper focused on disk-based DBMSs.
+Recently, we have seen a rapidly increasing number of in-memory DBMSs due to technological advancement in semiconductor memory~\cite{gar92}.
+Many in-memory DBMSs avoid strictly adhering to the ARIES protocol for recovery.
+This is due to the fact that a lot of the concepts in ARIES no longer apply to in-memory DBMSs.
+For instance, ARIES uses undo records to revert changes that have been applied to the database.
+However, undo records are excessive for many in-memory DBMS recovery systems~\cite{noisepage, malviya14, hekaton2013, scuba14, silo_r}.
+Performance concerns is another reason why in-memory DBMSs stray from the ARIES protocol.
+Logging requires the DBMS to interact with the disk.
+Therefore, logging I/O is a major bottleneck for an in-memory DBMS.
+This prompts in-memory DBMSs to minimize logging traffic~\cite{faeber17}.
+An example is the recovery of database indexes~\cite{faeber17}.
+Disk-based DBMSs log updates to index structures (e.g., B+ Tree) that allow faster recovery during log replay~\cite{aries}.
+In comparison, in-memory DBMSs do not log index updates~\cite{hekaton2013, noisepage, malviya14, silo_r} and choose to reconstruct indexes from scratch during log replay.
+Despite the differences, many in-memory DBMSs still use some form of logging~\cite{noisepage, malviya14, hekaton2013, silo_r} in their recovery systems.
+
+In addition to changes in recovery systems, in-memory DBMSs adopt various optimization algorithms for their execution engines\cite{kersten18}.
+These optimizations aim to increase execution throughput by reducing either the number of instructions that a DBMS executes to run a query, or the clock cycles per instruction (CPI)~\cite{boncz05}.
+Query compilation/code generation is an important optimization technique used by DBMSs to greatly reduce the number of instructions that the CPU needs to execute\cite{kersten18}.
+During execution, the DBMS breaks down a query into various tasks~\cite{noisepage, neumann11}.
+To speed up the execution of those tasks, the DBMS can compile them into native code (e.g., C/C++) with an off-shelf compiler\cite{boncz05}.
+Query compilation leads to faster query execution because it specializes both the data structures (e.g., hash table) and access methods of a DBMS towards execution efficiency~\cite{neumann11}.
+Moreover, the compiled code can be optimized around locality that increases the chance of a data tuple being propagated between operators in CPU registers\cite{menon17}.
+
+While query compilation succeeds in increasing the execution throughput of a DBMS, it comes with several drawbacks.
+Implementing a query compilation system requires additional knowledge of compiler systems (e.g., LLVM) and huge amounts of engineering work to translate different execution tasks\cite{kersten18}.
+The DBMS also generates low-level machine code that is hard to understand and debug~\cite{kersten18}.
+
+Moreover, the DBMS recovery system is isolated from the query compilation execution engine (i.e., a component of a DBMS that is responsible for execution).
+Most of the query compilation DBMSs we have surveyed~\cite{noisepage, neumann11, paroski16, craig14} have no conduits between their recovery systems and execution pipelines.
+The recovery system implements its own functionalities to update the database, but some of these functionalities already exist in the execution engine.
+For instance, we mentioned that in-memory DBMSs rebuild the indexes during log replay.
+When a redo process install changes on a table, it also retrieves the indexes on that table and update them (index update).
+However, the execution engine has already implemented this index update functionality.
+As a result, the DBMS contains two implementations that provide identical functionalities.
+Similar reasoning applies to other functionalities (e.g., constraint checking, trigger events) that may be used during log replay.
+The number of those functionalities in the recovery system will eventually amount to the scale of a separate execution engine that is built specifically for log replay.
+
+Therefore, if we can find a way to unify a DBMS's recovery system and execution engine, we can drastically reduce engineering overhead.
+Furthermore, we believe this unification also increases replay efficiency.
+Unless the rewritten functionalities in the recovery manager are executed as native machine code, they will be less efficient compared to those in the execution engine that uses query compilation.
+Therefore, if the DBMS allows the recovery system to invoke functionalities within the execution engine, the log replay process should be more efficient.
+
+\section{Contribution}
+We present a system design that will extend the execution engine to create custom programs for replaying physical log records.
+Our design converts log records into a format that the DBMS can compile into native code to be accepted by the execution engine.
+We will show that our approach removes the need for introducing extra implementations into the recovery system for functionalities that already exist in the execution engine.
+We will also verify our assumption that by utilizing the code generation execution engine, the DBMS will have increased log replay efficiency.
+
+We implement our approach in NoisePage\cite{noisepage}, a self-driving in-memory DBMS developed at Carnegie Mellon University.
+NoisePage is built in C++ and uses the PostgreSQL wire protocol for user communication.
+The DBMS also depends on query compilation for its execution engine.
+Our experimental results show that while the code generation recovery approach falls short of recovery throughput compared to baseline recovery, it is more efficient in certain areas (e.g., index maintenance, table constraints) and has higher scalabiltiy for a table with a larger number of columns.
+
+We structure the remainder of the thesis as follows.
+In Chapter 2, we provide more background information on database recovery, code generation, and motivations behind our approach.
+In Chapter 3, we present the recovery and code generation architecture in NoisePage and how we implemented our approach in this DBMS.
+In Chapter 4, we evaluate our implementation by comparing it against NoisePage's builtin recovery system.
+We discuss related works in Chapter 5 and conclude our thesis in Chapter 6.
 
 \chapter{Background}
+In this chapter, we discuss more background information on database recovery and code generation.
+We then explain how the two components interact within a DBMS and what changes need to be made for them to cooperate.
+
+\section{Transactions}
+Transactions are the atomic operations of any DBMSs~\cite{theo83}.
+They appear throughout a DBMS, including the recovery system and the execution engine responsible for code generation.
+A transaction is a sequence of interactions with the database.
+These interactions can be retrieving data, inserting data, removing data, etc.
+
 \section{Database Recovery}
-\begin{figure}[H]
+A DBMS ensures database integrity by satisfying the following conditions:
+\begin{itemize}
+    \item \textbf{Durability of Updates}~\cite{agrawal89}: All the changes made by committed transactions are durable.
+    \item \textbf{Failure Atomicity}~\cite{franklin97}: None of the changes made by aborted or failed transactions are persisted visible after recovery.
+\end{itemize}
+
+Many DBMSs~\cite{noisepage, malviya14, hekaton2013, silo_r, azure, postgres, mysql} combine logging and recovery protocols to preserve these two conditions in case of failure.
+Logging is the action of storing information about committed transactions on disk, while recovery is the action of restoring the database system into a consistent state.
+
+The DBMS stores logging information in a special data structure called a log record.
+A log record contains physical changes (e.g., changes made to a specific physical address) performed on the DBMS, or higher-level information (e.g., user-input query).
+In write-ahead logging \cite{aries}, the DBMS records physical database changes in a log file before the DBMS persists the changes on disk.
+A transaction is not allowed to commit until the DBMS persists its corresponding log record.
+
+For disk-based DBMSs, a log record in the log file can be either a redo record or an undo record~\cite{aries}.
+During recovery, a redo record installs the effects of committed transactions, while an undo record removes the effects on incomplete or aborted transactions.
+In-memory DBMSs, on the other hand, do not persist uncommitted changes on disk~\cite{noisepage, malviya14, hekaton2013, scuba14, silo_r}.
+This is because disk-based DBMSs generate dirty pages.
+These are pieces of modified entries in the database that reside in the memory, but have not yet been persisted to disk by the DBMS.
+However, in-memory DBMSs do not write dirty data to persistent storage~\cite{noisepage, malviya14, hekaton2013, scuba14, silo_r}.
+Therefore, in-memory DBMSs do not generate dirty pages, nor do they need to keep undo records.
+As a result, undo records are no longer required for their recovery systems.
+Moreover, in-memory DBMSs do not need to generate log records for indexes~\cite{faeber17}.
+During recovery, the DBMS reconstructs the database with redo records reconstructs and create the indexes simultaneously~\cite{malviya14, hekaton2013}.
+For each redo record, the recovery system finds the table that it modified, then retrieves all of its indexes from the database catalog ((\cref{fig:recovery_index_update}).
+Finally, for each index, the recovery system either updates it with new values, or remove entries from the index.
+
+\begin{figure}[t!]
 \centering
-\includegraphics[width=0.9\textwidth]{images/Logging.png}
-\caption{Write Ahead Logging.}
-\label{fig:logging}
+\centering
+\includegraphics[width=0.9\textwidth]{images/RecoverIndex.png}
+\caption{Index Update During Recovery.}
+ \label{fig:recovery_index_update}
 \end{figure}
-Database recovery systems are responsible for preserving the ACID principles of a DBMS in case of failure.
-Prior to 1980s, early recovery systems rely heavily on non-volatile memory storage.
-Most of the DBMSs were disk-based and the recovery protocols assume the existence of dirty pages.
-System R\cite{system_r}, for instance, creates shadow copy of updated pages and maintains log record updates.
-Nowadays, both the capacity and reliability of main memory have increased significantly.
-This greatly improved the capabilities of in-memory DBMSs, but their performance is still be threatened by the need to synchronize non-volatile memory with persistent storage.
-Modern recovery systems aim to reduce the synchronization overhead.
 
-Recovery systems consist of two components: logging and recovery.
-Logging is the action of storing transactional information on disk.
-These information are stored in a special data structure called a log record.
-A log record can either contain physical information (e.g.
-changes made to a specific physical address) or higher-level information (e.g.
-user-input query).
-In Write-ahead logging \cite{aries}, the DBMS records database changes in a log file before the changes can be persisted on disk.
-Each log record in the log file usually consists of redo records (install the effects of a committed transaction) and undo records (remove the effects on an incomplete or aborted transaction).
-Since in-memory DBMSs do not persist uncommitted changes, undo records are no longer required for the recovery system.
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.9\textwidth]{images/RecoveryCheckpoint.png}
+\caption{Recovery with Log Records and Checkpoints.}
+\label{fig:recovery_checkpoint}
+\end{figure}
 
-Recovery is the action of restoring the database system into a consistent state.
-If available, the DBMS first attempts to restore its previously recorded checkpoint.
-The database then initiates the log replay process to go through every log record stored since that checkpoint.
-SiloR\cite{silo_r} uses physical logging, checkpoints and an epoch-based recovery system.
-Transactions are assigned with different epochs and can only be persisted when their epochs are smaller then the current persistent epoch.
-VoltDB\cite{malviya14} uses a form of physiological logging that reduces transaction execution overhead.
-SAP HANA\cite{lee18}, however, uses a log-less approach through parallel table replication.
+Aside from logs, recovery systems can store complete snapshots (i.e., checkpoints) of the database.
+The DBMS can decide to take checkpoints based on a certain rule (e.g., timed-interval, number of transactions).
+Checkpoints reduce the time required for the database to recover.
+The DBMS can directly revert to a previous snapshot without replaying any log records before that snapshot.
+DBMSs often store both log records and checkpoints.
+During recovery, the DBMS finds its most recent snapshot, then replays any log records that are stored after that snapshot (\cref{fig:recovery_checkpoint}).
 
 \section{Query Compilation / Code Generation}
-When a query arrives, the DBMS parses the query into an expression tree.
-Eventually, the optimizer of the DBMS converts the expression tree into a physical plan tree.
-The physical plan represents how the query will be executed.
-It usually consists of operators that specify physical operations on the DBMS (e.g.
-inserting values into a certain memory location).
-For instance, \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100} can be represented by an update plan tree with a sequential plan as a child.
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.5\textwidth]{images/PhysicalPlanExample.png}
+\caption{Physical Plan that corresponds to \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100}.}
+\label{fig:physical_plan_example}
+\end{figure}
+When a query arrives, the DBMS parses the query into an abstract syntax tree (AST).
+The optimizer of the DBMS then converts the AST into a physical plan tree.
+The physical plan represents how the DBMS will execute the query.
+It consists of operators that specify physical operations on the DBMS (e.g., inserting values into a certain memory location).
+For instance, \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100} can be represented by an update plan tree with a sequential plan as a child (\cref{fig:physical_plan_example}).
 The execution engine then uses the plan tree for execution.
 
-The Iterator model\cite{volcano} is widely used in many disk-based DBMS (e.g.
-ORACLE RDBMS, MySQL, Microsft SQL Server) to interpret plan trees.
-The model "iterates" over each operator by implementing a \dbSQL{next()} function in each operator that returns the next tuples to be operated.
-The expressions inside the operators are evaluated using the provided tuples and generates output tuples.
-This approach suffers from frequent virtual function calls for in-memory DBMS.
-The Materialization model reduces this overhead by processing all the input tuples and emitting all the outputs at once ("materializing" the result as a single row or column).
-The Materialization model is suitable for OLTP DBMS since each OLTP query only accesses a small number of tuples at a time.
-However, since the Materialization results are stored in an in-memory buffer, the execution throughput is then bound by the size of the available memory space.
-While disk-based DBMSs are less concerned about the in-memory access since most of their operations are disk-based, this is especially critical for in-memory DBMSs.
+In a DBMS that uses query interpretation, its execution engine (i.e., a component of a DBMS that is responsible for execution) processes the physical plan by traversing the plan tree.
+Therefore, for every query, DBMS needs to follow pointers and resolve branching conditions (e.g., if, switch statements).
+This incurs overhead caused from virtual function calls, branch mispredictions and instruction cache misses.
 
-Vectorization and Query Compilation are two state-of-the-art techniques employed for in-memory DBMSs to reduce memory overhead\cite{kersten18}.
-The Vectorization model is a variant of the Iterator model.
-Instead of emitting a single tuple at a time, the Vectorization model emits a batch of tuples at a time to amortize the cost of memory access.
-The functions that operate on these tuples are also specialized to be able to take in a batch of tuples at a time and more efficient compared to iterating over a single tuple at a time.
-
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{images/QueryCompilation.png}
-\caption{Query Compilation architecture.}
+\caption{Query Compilation Architecture.}
 \label{fig:query_compilation}
@@
 
-Query Compilation eliminates the cost of interpretation by compiling the physical plan to machine code.
-The machine code of the corresponding physical plan can then be executed repeatedly by the execution engine, removing any need for virtual function calls.
-Before being compiled to machine code, the physical plan is often first converted into an intermediate representation (e.g.
-C/C++ code, LLVM IR, or custom Domain Specific Language) that follows the grammar of an imperative language.
-This simplifies the process of converting the physical plan into machine code, as the intermediate representation form is designed to resembles closely to SQL statements.
+Query compilation eliminates the cost of query interpretation by compiling the physical plan to machine code that is specific for that query.
+The machine code of the corresponding physical plan can then be executed repeatedly by the execution engine, removing any need for virtual function calls and branching resolutions.
+Generally, there are two ways of compiling queries.
+In the first approach, the DBMS creates source code that is compiled into native code with an external compiler (e.g., gcc) ~\cite{krikellas10, klonatos14}.
+This approach is used in Amazon Redshift and pre-2016 MemSQL~\cite{paroski16}.
+In the second approach, the DBMS first converts physical plan into an intermediate representation (e.g., LLVM) that follows the grammar of an imperative language.
+This approach simplifies the process of converting the physical plan into machine code, as the intermediate representation form is designed to resemble SQL statements.
+Moreover, it removes the need for the DBMS to use the compiler as an external process.
+This approach is adopted by NoisePage~\cite{menon17}, Hyper~\cite{neumann11}, Hekaton~\cite{craig14}, and SingleStore~\cite{paroski16}.
 
-Hyper\cite{neumann11} is one of the pioneers of query compilation.
-It uses a data-centric compilation approach that splits the query plan into pipelines and compiles each pipeline separately.
-Vectorwise\cite{raducanu13} chooses to pre-compile huge amounts of micro function primitives that perform simple operations on typed data.
-During runtime, the DBMS can directly invoke these compiled primitives for query execution.
+\section{Motivation}
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.9\textwidth]{images/Motivation.png}
+\caption{Possible ways to integrate a physical logging system with code generation.}
+\label{fig:motivation}
+\end{figure}
+While code generation is a powerful optimization technique, it poses design challenges for the recovery system.
+Both the execution engine and physical log records perform identical physical operations on a database table.
+The native code generated through query compilation operates directly on physical tables.
+For physical log records, they are already designed as a data structure that represents physical changes made to the database.
+During recovery, the DBMS simply replays those changes on the table.
+
+However, the recovery system has no means of reusing functions within the execution engine to install physical changes from log records.
+The execution engine does not accept physical log records because it follows along the query compilation pipeline.
+This pipeline transforms one logical representation of a query to another, until the query becomes native code.
+There is no place for physical log records to exist within this transformation pipeline.
+
+There are two solutions to this problem.
+One solution is to re-implement functions to install physical changes on data tables within the recovery system.
+However, this approach completely isolates the recovery system from the execution engine.
+This means that if the recovery system wishes to use a functionality within the execution engine, it has to rewrite that same functionality within the recovery system (\cref{fig:problem}).
+
+The other solution is to integrate physical logging with the execution engine.
+\cref{fig:motivation} shows four possible paths that allow the execution engine to accept physical log records.
+Converting the log record directly into native code requires the implementation of a new compiling framework, which makes this option infeasible.
+Therefore, the closest point where the log records can reach before entering the execution engine is physical plans.
+We then reach the conclusion to convert log records into physical plans.
+Firstly, compared to raw queries or abstract syntax trees, physical plans are closer to the end of the query compilation pipeline.
+If we convert a log record into an abstract syntax tree (AST), the DBMS still needs to do additional work to convert the AST into a physical plan.
+Secondly, we will show in later sections that it is enough for physical plans to represent contents of a log record.
+
+If the DBMS can convert log records into physical plans, the recovery system can rely on code generation for recovery operation.
+This code generation recovery approach brings around several benefits.
+Firstly, it significantly reduces engineering overhead.
+Compiling contents of the log records into machine code allows the recovery system to seamlessly integrate with the built-in execution engine.
+We have explained in Section 2.1 that the recovery system needs to perform extra operations (e.g.
+index updates, constraint checking in \cref{fig:problem}) besides updates to the data tables.
+These functionalities exist within the execution engine.
+If we allow the recovery system to reuse existing functionalities by communicating with the execution engine through native code, then the recovery system no longer needs to implement its own functionalities.
+Secondly, utilizing the code generation execution engine provides more efficient functionalities.
+A function implemented by the recovery system will not be as efficient as one that exists in the execution engine.
+
+These potential benefits motivate us to design a code generation recovery system based on conversion from log records to physical plans.
+In the next chapter, we will present the implementation of our approachhbmmn .
 
 \chapter{Method}
-\begin{figure}[H]
+\afterpage{%
+\begin{figure}[t!]
 \centering
-\includegraphics[width=0.9\textwidth]{images/Proposal.png}
+\includegraphics[width=0.9\textwidth]{images/Method.png}
 \caption{Code Generation Recovery Mechanism.}
 \label{fig:codegen_recovery}
-\end{figure}
+\end{figure}}
 
-Our proposal is to extend the recovery layer and query compilation layer to support the interpretation of log records. This approach can be realized in three steps:
+We now describe how to extend the recovery system and query compilation pipeline to support log replay with code generation.
+Our implementation follows along the third option shown in \cref{fig:motivation}.
+We first enable the recovery system to convert log records into compiled native code.
+To achieve this, the DBMS needs to interpret the log records into physical plans and prepare them for compilation.
+We then need to make minor modifications to the execution engine, so it will be able to correctly execute the native code generated by the recovery system.
+We further summarize our approach (\cref{fig:codegen_recovery}) in three steps:
 
 \begin{enumerate}
-\item Extract information from log records.
-\item Convert the information into expressions that can be accepted and compiled by the code generation engine (Code Generation Step).
+\item Interpret and extract contents from log records.
+\item Convert the extracted contents into physical plans that can be compiled into native code (Code Generation).
 \item Run the compiled machine code using the execution engine.
 \end{enumerate}
-
-This code generation approach brings around several benefits.
-Firstly, it significantly reduces engineering overhead.
-Compiling contents of the log records into machine code allows the recovery manager to seamlessly integrate with the built-in execution engine.
-This removes the need to specifically construct code to handle tasks that are not compiled to machine code.
-
-\begin{figure}[H]
-\centering
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RecoverIndex.png}
- \caption{Index Update During Recovery.}
-  \label{fig:pipeline_graph}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
- \centering
- \includegraphics[width=0.9\textwidth]{images/RecoverConstraint.png}
- \caption{Constraint Checking During Recovery.}
-  \label{fig:pipeline_code}
-\end{subfigure}
-\caption{Problems faced by conventional recovery in code generation DBMS.}
-\label{fig:log_record}
-\end{figure}
-
-Secondly, it also resolves tricky problems that will appear when writing specialized code for log records.
-An example of this is index updates.
-Without code generation, the recovery manager needs to specifically reconstruct the index given an insert or update record.
-With code generation, the index updates are taken care directly by the execution engine that already contains high-performance specialized code for index maintenance.
-Similar problems, such as constraint checking
-
-Our proposed architecture is implemented and tested in \textit{Noisepage} \cite{noisepage}, a self-driving in-memory DBMS developed at Carnegie Mellon University.
-Noisepage is built in C++ and uses the PostgreSQL wire protocol for user communication.
-The DBMS supports ACID transactions and \texttt{SNAPSHOT-ISOLATION}.
-Noisepage uses the conventional WAL recovery scheme with physical logging and depends on query compilation for query execution.
-We will dive into the details of Noisepage's recovery structure in the following sections.
+We implement our approach in NoisePage~\cite{noisepage}.
+Noisepage depends on a recovery system that uses physical logging schemes.
+It also adopts query compilation optimization for its execution engine.
 
 \section{Recovery System Architecture}
-\subsection{Transaction}
-Transactions are the atomic operations of any DBMS.
-Noisepage uses a transaction to perform physical updates on the database and physical logging to disk.
-Noisepage uses a multi-versioned delta store transaction engine that ensures \texttt{SNAPSHOT-ISOLATION}.
-The transaction engine allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis.
+ To show how we will extend the recovery system, we first present some background information on the recovery system architecture of NoisePage.
+NoisePage's recovery system consists of three major components: Transaction Manager, Logging Manager, and Recovery Manager.
+Both the Logging Manager and the Recovery Manager interact with the Transaction Manager to achieve logging and recovery respectively.
+We will first give an overview of NoisePage's transaction manager,
+
+\subsection{Transaction Manager}
+The Transaction Manager is responsible for creating and maintaining database transactions.
+NoisePage uses a transaction to perform physical updates on the database and physical logging to disk.
+NoisePage uses a multi-versioned delta store Transaction Manager that ensures \texttt{SNAPSHOT-ISOLATION}.
+The Transaction Manager allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis.
 A transaction's lifecycle begins when the transaction begins, and ends after the log manager has serialized its changes to disk.
 
-Each data tuple in noisepage is uniquely identified through a tuple slot that stores the offset of a tuple.
+Each data tuple in NoisePage is uniquely identified through a tuple slot that stores the offset of a tuple.
 When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table.
@@
 These redo records are essential for executing log replay correctly.
-The structure of a redo record is shown later in \ref{fig:log_record}.
+The structure of a redo record is shown later in \cref{fig:log_record}.
 
 A transaction uses a buffer (redo buffer) that stores all the delta records that will appear in its lifetime.
-Each redo buffer has a fixed size of 4096 bytes and is allocated from a centralized buffer pool.
+Each redo buffer has a fixed size and is allocated from a centralized memory pool.
 Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to dump a corresponding delta record of the change.
-If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the buffer pool.
+If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the memory pool.
 The redo buffer allows the logging components to process the changes before the transaction is completed.
 
-When a transaction commits, the transaction engine creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager.
+When a transaction commits, the Transaction Manager creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager.
 The commit record is appended to the redo buffer.
@@
 
-\subsection{Write-Ahead Logging}
-\begin{figure}[H]
+\subsection{Log Manager}
+\begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{images/WAL.png}
-\caption{Logging Architecture. Log manager uses the log serializer and log consumers to convert redo buffers into log records.}
-\label{fig:wal}
+\caption{\textbf{Log Manager}.}
+\label{fig:log_manager}
 \end{figure}
 
-The log manager is responsible for performing write-ahead logging in Noisepage.
+The Log Manager (\cref{fig:log_manager}) is responsible for performing write-ahead logging in NoisePage.
 All the changes are serialized to disk as logs before they are performed on the DBMS storage.
 Each log record is self-contained and does not require any additional metadata to be replayed.
-The log manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.
+The Log Manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.
 
 \subsubsection{Log Serializer}
-The log serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit.
-The log serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers of 4096 bytes.
-These buffers will be consumed by the Disk Log Consumer for persistence.
+The Log Serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit.
+The Log Serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers.
+These buffers will be consumed by the Log Consumer for persistence.
 
@@
 However, the log records performed by an individual transaction is guaranteed to appear in the same order as they were created.
-This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of Noisepage during log replay.
+This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of NoisePage during log replay.
 \subsubsection{Log Consumer}
-Buffers supplied by the log serializer can be consumed by different log consumer tasks.
-Each log consumer is given a copy of the original buffer to achieve parallelism.
-Noisepage uses two type of log consumers: disk log consumer and replication log consumer.
+Buffers supplied by the log serializer can be consumed by different Log Consumer tasks.
+Each Log Consumer is given a copy of the original buffer to achieve parallelism.
+NoisePage uses two type of log consumers: Disk Log Consumer and Replication Log Consumer.
 
-The disk log consumer repeatedly polls for new buffers from the log serializer and writes the changes to disk.
+The Disk Log Consumer repeatedly polls for new buffers from the log serializer and writes the changes to disk.
 To improve performance, disk log consumer uses group commit that allows a batch of changes to be committed over a time period specified by the user.
@@
 
-The replication log consumer also polls for new buffers, but instead of persisting them to disk, sends the new buffers over across the network to database replicas.
+The Replication Log Consumer also polls for new buffers. Instead of persisting them to disk, it sends the new buffers over across the network to database replicas.
 
-\subsection{Log Replay}
-\begin{figure}[H]
+\subsection{Recovery Manager}
+\begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{images/RecoveryManager.png}
-\caption{How recovery operates in Noisepage.}
-\label{fig:recoverymanager}
+\caption{Recovery Manager.}
+\label{fig:recovery_manager}
 \end{figure}
 
-Noisepage uses four different record types: \textit{Redo, Delete, Commit, Abort}.
-Redo and Delete records are not processed immediately.
+\cref{fig:recovery_manager} shows the architecture of the Recovery Manager that is responsible for log replay.
+NoisePage uses four different record types: redo, delete, commit, abort.
+Redo and delete records are not processed immediately.
 Instead, they are placed inside a deferred changes buffer, ordered by their transaction timestamp.
-An Abort record removes the corresponding Redo/Delete record from the buffered changes.
-A Commit record initiates a transaction to iterate over the deferred changes buffer.
+An abort record removes the corresponding redo/delete record from the buffered changes.
+A commit record initiates a transaction to iterate over the deferred changes buffer.
 Once all the deferred changes are processed, the buffer is cleared and the transaction is committed.
-The transaction aborts if any log record is malformed or any recovery operation fails.
+The transaction aborts if any log record is malformed or any replay fails.
 
-\subsubsection{Data Structure}
-\begin{figure}[H]
+\subsubsection{Log Record}
+\begin{figure}[t!]
 \centering
@@
 \end{subfigure}
-\caption{The internal log structures.}
+\caption{Log Record.}
 \label{fig:log_record}
@@
 
-A physical log record consists of a header followed by a record body.
+A physical log record consists of a header followed by a record body (\cref{fig:log_record}).
 The header stores metadata about the log record itself, including its record type, size and begin timestamp of the transaction that generated the log record as mentioned in Section 3.1.1.
@@
 Depending on the value type, the values may bee stored in-line or not in-inline.
-Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may stores the pointer in the values column that points to another location in the record.
+Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may store a pointer in the values column that points to another physical location in the record.
 
 Tuple slots in the log records will no longer be valid memory locations during recovery.
-Instead, the recovery manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted.
-The mapping allows the recovery manager to apply changes to the correct memory locations in the new memory environment.
+Instead, the Recovery Manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted.
+The mapping allows the Recovery Manager to apply changes to the correct memory locations in the new memory environment.
 
-Since only Redo and Delete records will effectively modify physical storage, we only need to analyze the record body of those two record types.
+Since only redo and delete records will effectively modify physical storage, we only need to analyze the record body of those two record types.
 Modifications of a query to the DBMS can be resolved down to the modification of tuple slots on the physical level.
 
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{images/Delta.png}
-\caption{Redo Record Delta.}
-\label{fig:logging}
+\caption{Redo Record Projected Row.}
+\label{fig:projected_row}
 \end{figure}
 
- While storing tuple slots is sufficient for delete records, Redo Records require additional data to perform Recovery Insert and Update operations.
-These data are kept in the Record Delta, which stores a mapping from column ids to values.
+ While storing tuple slots is sufficient for delete records, redo records require additional data to perform log replay.
+These data are kept in the record projected row (\cref{fig:projected_row}), which stores a mapping from column ids to values.
+The DBMS retrieves individual values by computing offsets.
 
-\subsubsection{Recovery Operation}
-A Noisepage recovery step has three different operations:
+\subsubsection{Replay Step}
+A replay step refers to the recovery manager replaying a single redo record or delete record. A NoisePage replay step has three different scenarios:
 \begin{itemize}
-    \item \textbf{Insert}: Represented by a Redo Record. The recovery manager inserts the values stored within the redo record into the given tuple slot and updates the indexes.
-    \item \textbf{Delete}: Represented by a Delete Record. The recovery manager deletes the tuple slot specified by the delete record and updates the indexes.
-    \item \textbf{Update}: Represented by a Redo Record. The recovery manager updates the tuple slot using the values within the redo record and updates the indexes.
+    \item \textbf{Insert Replay}: The Recovery Manager inserts the values stored within a redo record into the given tuple slot and updates the indexes.
+    \item \textbf{Delete Replay}: The Recovery Manager deletes the tuple slot specified by a delete record and updates the indexes.
+    \item \textbf{Update Replay}: The Recovery Manager updates the tuple slot using the values within a redo record and updates the indexes.
 \end{itemize}
 
-While both Insert and Update are represented using the same Redo Record structure, they can be differentiated by keeping track of all the tuple slots found in log records.
-If a tuple slot inside a Redo Record is unseen before, then the Redo Record represents an Insert operation.
+While both insert and update redo records are represented using the same redo record structure, they can be differentiated by keeping track of all the tuple slots found in log records.
+If a tuple slot inside a redo record is unseen before, then the redo record represents an Insert operation.
 Otherwise, the Redo Record represents an Update operation.
@@
 \section{Recovery Code Generation}
-The code generation structure in Noisepage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \ref{fig:query_compilation}.
+The code generation structure in NoisePage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \cref{fig:query_compilation}.
 A query is interpreted, optimized and converted into a physical plan tree.
@@
 
-A physical plan tree/node is then converted into an operator pipeline through a plan translator.
+A plan translator then converts a physical plan tree/node into an operator pipeline.
 Each plan translator correspond to a plan node in the original plan tree.
-A pipeline contains an ordered collection of relational operators that are not yet materialized, ended with a pipeline breaker (e.g.
-sort tuples, creating a hash table).
+A pipeline contains an ordered collection of relational operators that are not yet materialized, ended with a pipeline breaker (e.g., sort tuples, creating a hash table).
 The pipeline is only materialized whenever a pipeline breaker has been encountered.
 
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
@@
 \end{figure}
-Noisepage then compiles each operator pipeline into a domain-specific language referred to as TPL.
-Each function of a query inside the TPL code is then compiled to bytecode called TBC.
-TBC can be interpreted by a built-in virtual machine, or further compiled into an LLVM module.
-The execution engine can either interpret TBC, or run the compiled LLVM module.
+NoisePage then compiles each operator pipeline into a customized imperative language referred to as (TPL).
+The DBMS then compiles each function of a query inside the TPL code into some bytecode (TBC).
+The bytecode can be interpreted by a built-in virtual machine, or further compiled into an LLVM module.
+The execution engine can either interpret TBC, or run the compiled LLVM module (\cref{fig:llvm}).
 
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
@@
 
-We further demonstrate the process of query compilation through the example in \ref{fig:codegen_example}.
-To run the query \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}, the DBMS needs to first do a sequential scan to find out the columns, and then do a delete operation on the found columns.
+We further demonstrate the process of query compilation through the example in \cref{fig:codegen_example}.
+To run the query \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}, the DBMS needs to first do a sequential scan to find out the columns, then do a delete operation on the found columns.
 These two steps are manifested in the delete plan node that includes a sequential scan plan node as a child.
-The plan node is translated into an operator pipeline that specifies how each operation will be exactly executed in the DBMS.
-The operator pipeline is then converted to TPL code.
+The plan node then translates into an operator pipeline that specifies how each operation will be exactly executed in the DBMS.
+The operator pipeline then converts to TPL code.
 Each function inside the TPL code correspond to a piece of function inside the execution engine that will be evetually invoked once the compiled native code is run.
@@
 
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
@@
 \caption{Process for converting log records into native code.}
-\label{fig:logging}
+\label{fig:expression_generation}
 \end{figure}
-We imagine each recovery operation as a form of SQL statement that can be converted into native code through the code generation process discussed above.
+We imagine each replay as a form of SQL statement that can be converted into native code through the code generation process discussed above (\cref{fig:expression_generation}).
 The code generation process for a log record can be divided into two steps: plan generation (creating the physical plan node) and plan translation (translating the plan into pipelines).
-Each recovery operation (\textbf{Insert}, \textbf{Update}, \textbf{Delete}) will be matched to a corresponding plan node.
+Each replay (Insert Replay, Update Replay, Delete Replay) will be matched to a corresponding plan node.
 
-\subsection{Insert Operation Conversion}
+\subsection{Insert Replay Conversion}
 
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
@@
 \end{figure}
-Insert plan nodes correspond to \texttt{INSERT} statements.
-For an insert redo record, we can represent it as the SQL statement \dbSQL{INSERT INTO x VALUES y}.
-The insert plan node for this statement does not require a child plan node to function.
+Insert plan nodes correspond to \dbSQL{INSERT} statements.
+We can represent an insert redo record as the SQL statement \dbSQL{INSERT INTO x VALUES y}.
+The insert plan node for this statement does not require any child plan nodes to function (\cref{fig:insert_conversion}).
 This makes it straightforward to convert from an insert redo record to a plan node.
 Index information and column ids can be retrieved from the catalog.
-Values stored in the redo records are in their most basic form (e.g.
-integer, float, double) as constants and can be converted into expressions to be accepted by the insert plan node.
+Values stored in the redo records are in their most basic form (e.g., integer, float, double) as constants and can be converted into expressions to be accepted by the insert plan node.
 The operation of transferring values, however, require one-to-one copy of each value from the log record into the plan node.
 
-\subsection{Delete Operation Conversion}
-\begin{figure}[H]
+\subsection{Delete Replay Conversion}
+\begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{images/DeleteConversion.png}
-\caption{Conversion from a Delete Redo Record into a Delete Plan Node. Tuple slot that will be used in the pipeline is "smuggled" from the redo record through execution context.}
+\caption{Conversion from a Delete Redo Record into a Delete Plan Node. Tuple slot that will be used in the pipeline is "smuggled" from the redo record through Execution Context.}
 \label{fig:delete_conversion}
@@
 
-Delete plan nodes correspond to \texttt{DELETE} statements and require child nodes to function.
+Delete plan nodes correspond to \dbSQL{DELETE} statements and require child nodes to function.
 A delete plan node uses the child plan node to figure out which tuple slot to schedule for deletion.
 This is different for delete records.
-The mapping from recovery manager combined with the original tuple slot already points to the tuple slot to delete.
+The mapping from Recovery Manager combined with the original tuple slot already points to the tuple slot to delete.
 In this case, there will be no need for the plan node to go through the child node and resolve the delete tuple slot.
-This is because, eventually, the delete plan node will attempt to create a Delete Record, which is structurally identical to the Delete Record from the recovery manager.
-Other operations, such as index deletions, are still necessary.
+This is because, eventually, the delete plan node will attempt to create a Delete Record, which is structurally identical to the Delete Record from the Recovery Manager.
+We will therefore change the behavior of the operator pipeline to retrieve the delete slot supplied by the recovery manager during recovery (\cref{fig:delete_conversion}), and remove the step to resolve which tuple slot to delete.
+Other operations, such as index deletions, are still necessary and are kept in place.
 
-\subsection{Update Operation Conversion}
-Update plan nodes correspond to \texttt{Update} statements and require child nodes to function.
+\subsection{Update Replay Conversion}
+Update plan nodes correspond to \dbSQL{Update} statements and require child nodes to function.
 Unfortunately, update plan nodes are not very compatible with update redo records.
-After an update plan is translated into an update operator pipeline, it expects set clauses returned from the child node (e.g.
-sequential scan).
-These set clauses, however, describe higher-level logical operations and cannot be constructed easily as creating primitive expressions for insert plan nodes.
-\begin{figure}[H]
+After an update plan is translated into an update operator pipeline, it expects set clauses returned from the child node (e.g., sequential scan).
+These set clauses, however, describe higher-level logical operations and cannot be constructed as easily and efficiently as creating primitive expressions for insert plan nodes.
+\begin{figure}[t!]
 \centering
@@
 \end{figure}
-The update plan node therefore needs to be extended to accept primitive values just like an insert plan node.
-We then modify the operator pipeline to process the primitive values instead of set clauses during a recovery step.
-Functions inside the operator pipeline will now apply to the primitive value expression.
-Similar to delete record transformations, there is no need for the update pipeline to resolve the update tuple slot or creating a new update record, as these information already exist in an update redo record and the tuple slot mapping in the recovery manager.
+The update plan node therefore needs to be extended to accept primitive values like an insert plan node.
+We also modify the operator pipeline to process the primitive values instead of set clauses during a replay step (\cref{fig:update_conversion}).
+Functions inside the operator pipeline will now apply to the primitive value expressions.
+Similar to delete record transformations, there is no need for the update pipeline to resolve the update tuple slot or to create a new update record, as these information already exist in an update redo record and the tuple slot mapping in the Recovery Manager.
 
-\section{Execution}
-The compiled query is finally run by the execution engine.
-The compiled query invokes different operation codes (e.g.
-\texttt{OpStorageInterfaceTableDelete} in \ref{fig:codegen_example}) to complete the corresponding operator in the TPL code.
+\section{Recovery Execution}
+The execution engine is reponsible for running the compiled query.
+After we have finished the setup for compiling different replays into native code, we will explain how they will affect the execution engine.
 
-We explained earlier that the for physical plan nodes, they are expected to retrieve relevant tuple slots from the recovery manager, instead of supplying the tuple slots for an operation.
-We can circumvent this rule by modifying operation code functions.
+\subsection{Storage Interface}
+The compiled query invokes different operation codes (e.g., \texttt{OpStorageInterfaceTableDelete} in \cref{fig:codegen_example}) to complete the corresponding operator in the TPL code.
+Executing the operation codes requires access to the storage layer of the DBMS.
+In NoisePage, the storage layer is separate from the execution engine.
+Therefore, it uses a Storage Interface to interact with contents in the storage layer.
+The Storage Interface stores objects that allow manipulation over a certain DBMS table.
+These objects include an Execution Context, a redo record that will be used to persist changes done to the table, projected rows associated with the redo record and index objects associated with table.
+A Storage Interface is reconstructed every time an execution of a query begins, and is destructed after the query has been executed.
+We will need to modify functions that rely on the Storage Interface and access members that reside within the Storage Interface.
 
-\begin{figure}[H]
+\subsection{Execution Changes}
+\begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{images/RecoveryExecution.png}
-\caption{Information from log record can be retrieved from the execution context.}
+\caption{Information from log record can be retrieved from the Execution Context.}
 \label{fig:recovery_execution}
@@
 
-An execution context encapsulates information that is supplied by upper layers of the DBMS.
+We explained in Section 3.2.2 and 3.2.3 that for our purposes, we expect deletes/updates to retrieve relevant tuple slots from the Recovery Manager.
+However, the built-in pipeline assumes that the tuple slots are retrieved later during execution.
+We can circumvent this rule by modifying operation code functions and the execution context.
+The operation code functions provide access to the Storage Interface that is required to access the execution context.
+
+An Execution Context encapsulates information that is supplied by upper layers of the DBMS.
 These information include access to the database catalog, the transaction running the query, etc.
-For our purposes, we can add tuple slot information into the execution context.
-Since the execution context can be accessed throughout the execution of a compiled query, we can modify the behavior of operation codes to override with the recovery tuple slots if necessary.
+We can also add tuple slot information into the Execution Context.
+Since the Execution Context can be accessed throughout a Storage Interface, we can modify the behavior of operation codes to override with the recovery tuple slots, as shown in \cref{fig:recovery_execution}.
 
-For Delete and Update operations, we need to supply the execution context with the log record itself.
-Since we modified the delete translator to skip the process of looking through child plan node for objects to delete, we will deliver the Delete Record from the recovery manager directly to the operation codes that will be expecting a Delete Record generated by a regular delete plan node.
+For Delete and Update Replays, we need to supply the Execution Context with the log record itself.
+Since we modified the delete translator to skip the process of looking through child plan node for objects to delete, we will deliver the Delete Record from the Recovery Manager directly to the operation codes that will be expecting a Delete Record generated by a regular delete plan node.
+The same process applies to Update Replays.
 
@@
 
-The naive implementation for code generation searches for table metadata and re-compiles the native code every time the recovery manager reads in a new log record.
-This is very inefficient and requires huge amounts of memory access and copy.
-We can use caching techniques to reduce performance overhead.
+The naive implementation for code generation searches for table metadata and re-compiles the native code every time the Recovery Manager reads in a new log record.
+This is very inefficient and requires egregious amounts of memory access and copy.
+We can use caching techniques to reduce this performance overhead.
 
@@
 \subsubsection{Physical to Logical Column Mapping}
-The recovery manager needs to maintain a mapping from physical column ids to logical column ids for each table.
+The Recovery Manager needs to maintain a mapping from physical column ids to logical column ids for each table.
 This is required because the column ids recorded in the log records are physical ids, while all the query compilation components expect logical column ids.
@@
 \subsubsection{Column Value Type Mapping}
-For redo records, the recovery manager also needs to maintain a mapping from each column to its corresponding SQL value type (e.g.
-\dbSQL{INTEGER, FLOAT, VARLEN}).
+For redo records, the Recovery Manager also needs to maintain a mapping from each column to its corresponding SQL value type (e.g., \dbSQL{INTEGER, FLOAT, VARLEN}).
 This mapping is used to correctly copy the values out of each redo record into a physical plan node.
+\subsubsection{Catalog Access}
+Retrieving access to a database catalog is costly.
+Profiling results show that a single operation to retrieve catalog access can account for 10\% of total instructions of a replay step.
+Therefore, a mapping is used to keep track of catalog accessors.
 
 \subsection{Compiled Query}
-A major benefit of Query Compilation is that the compiled query can be reused by the execution engine.
-If the contents of a physical plan node remains largely unchanged, then there is no need to reconstruct the physical plan every time the recovery manager processes a log record.
+A major benefit of query compilation is that the compiled query can be reused by the execution engine.
+If the contents of a physical plan node remains largely unchanged, then there is no need to reconstruct the physical plan every time the Recovery Manager processes a log record.
 
-We can cache each type of recovery operation separately and use a combination of database and table id as unique identifiers.
-The cached query can be re-usesd as long as the table schema remains unchanged.
+We can cache each type of replay separately and use a combination of database and table id as unique identifiers.
+The Recovery Manager can re-use cached queries as long as the table schema associated with them remains unchanged.
 
 \subsubsection{Parameterization}
-\begin{figure}[H]
+\begin{figure}[t!]
 \centering
@@
 
-Caching is trickier for recovery operations that need to transfer values into plan nodes.
-A primitive expression will change every time its underlying value changes.
-A solution to this is to parameterize the primitive expression given to a insert/update plan node.
-During run-time, the execution context pass in actual values into the parameterized expressions.
+Caching is trickier for replays that need to transfer values into plan nodes.
+The recovery manager will need to reconstruct a primitive expression every time its underlying value changes.
+A solution to this is to parameterize the primitive expression given to an insert/update plan node.
+During run-time, the Execution Context pass in actual values into the parameterized expressions.
 This approach allows a compiled query to keep value type information for each table column and remain flexible enough to swap in actual values without recompilation.
@@
 \chapter{Experimental Evaluation}
-We run our experiments on Ubuntu 20.04.
-Dual-socket 10-core Intel Xeon E5-2630v4 CPU, 128 GB of DRAM, and a 500 GB Samsung 970 EVO Plus SSD.
+We now evaluate the performance of our code generation recovery.
+The experiments focus on comparing the performance difference between code generation recovery and baseline recovery.
+Each experiment has two phases: the preparation phase and the recovery phase.
+During the preparation phase, we use three operations to generate log records:
+\begin{itemize}
+    \item \dbSQL{INSERT} $\rightarrow$ \textbf{Insert Redo Record}: Create a random row with random values and insert into a random table.
+    \item \dbSQL{UPDATE} $\rightarrow$ \textbf{Update Redo Record}: For a random row from a random table, update a random tuple slot with a random value.
+    \item \dbSQL{DELETE} $\rightarrow$ \textbf{Delete Record}: For a random row from a random table, delete a random tuple slot.
+\end{itemize}
+We run these operations randomly for some number of times during the preparation phase.
+During the recovery phase, we load the log records into memory from disk and replay them with the Recovery Manager.
+We then measure the performance of only the replay step in our throughput measurement experiments.
+
+We conduct all the experiments on a machine with the following specifications:
+\begin{itemize}
+    \item \textbf{OS}: Ubuntu 20.04
+    \item \textbf{CPU}: Intel(R) Xeon(R) Gold 5218R CPU @ 2.10GHz
+    \item \textbf{RAM}: 188 GB 2 sockets, 20 cores/socket, 40 threads/socket
+\end{itemize}
 
 \section{Throughput Measurement}
+Log replay generates transactions through commit records (See Section 3.1.2). Therefore, we can use transactions throughput to measure the performance of the recovery phase.
+
+We use Google Benchmark for our measurement setup.
+The table starts with an initial table size of $10^6$ tuples.
+During the preparation phase, we run five preparation operations for each transaction, with a total of $10^5$ transactions.
+Each table uses a single column and \dbSQL{INTEGER} data type.
+
 \subsection{Insert}
-We first measure the throughput (number of transactions executed per time period) of insert operations against the baseline recovery performance on a single table.
-The recovery manager uses only our insert implementation for recovery.
 \begin{table}[H]
 \begin{center}
-\begin{tabular}{ |c|c|c| } 
- \hline
-Statements Ratio & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\ 
- \hline
- 100\% Insert & $1.12 \times 10^4$ txns/s & $5.60 \times 10^3$ txns/s\\
- \hline
- 50\% Insert, 50\% Select & $1.32 \times 10$ txns/s & $6.60 \times 10^3$ txns/s \\ 
- \hline
+\begin{tabular}{ c c c } 
+ \toprule
+\textbf{Statements Ratio} & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\ 
+ \toprule
+ 100\% Insert & $1.12 \times 10^4$ txns/s & $6.00 \times 10^3$ txns/s\\
+ \bottomrule
 \end{tabular}
-\caption{\textbf{Insert throughput measurement setup on a single table}. The table uses an initial table size of $10^6$. Each transaction runs 5 statements. A total of $10^5$ transactions are run.}
+\caption{\textbf{Insert Replay Throughput}.}
 \label{tab:throughput_exp_insert}
@@
 \end{table}
+We first measure the throughput (number of transactions executed per time period) of Insert Replays against the baseline recovery performance on a single table.
+The Recovery Manager uses only our insert replay implementation for recovery.
 
-2x difference. Memory copies.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ l c c } 
+ \toprule
+\textbf{Functionality} & \textbf{IR\%} & \textbf{Normalized IR\%} \\ 
+ \toprule
+Query Execution (Op Codes) & 79.35 & 81.7\% \\
+ \midrule
+ Execution Context Initialization & 9.35 & 9.62\%  \\ 
+ \midrule
+ Retrieving Redo Values & 6.19 & 6.37\%  \\
+ \midrule
+ Metadata Cache Map Access & 2.26 & 2.32 \%  \\
+ \midrule
+ Compile & 0.03 & 0.03 \% \\
+ \bottomrule
+\end{tabular}
+
+\caption{\textbf{Callgrind breakdown on an Insert Replay step}. IR (instructions read) refers to the number of instructions read in total by the selected function during a certain function execution.}
+\label{tab:throughput_exp_insert_overhead_all}
+\end{center}
+\end{table}
+
+\begin{table}[H]
+\begin{center}
+\resizebox{\columnwidth}{!}{\begin{tabular}{ l c c l } 
+ \toprule
+\textbf{Operation Code} & \textbf{IR\%} & \textbf{Normalized IR\%} & \textbf{Description}\\ 
+ \toprule
+\texttt{StorageInterfaceInit} & 29.88 & 74.4\% & Initializes a Storage Interface. \\
+ \midrule
+\texttt{StorageInterfaceTableInsert} & 6.00 & 14.9\% & Inserts a tuple into a table.\\ 
+ \midrule
+\texttt{StorageInterfaceFree} & 2.76 & 6.87\% & Deallocates a Storage Interface.\\
+ \midrule
+\texttt{StorageInterfaceGetTablePR} & 1.40 & 3.48\% & Retrieves a table's projected row.\\
+ \midrule
+Other & 0.12 & 0.35\% & \\
+ \bottomrule
+\end{tabular}}
+\caption{\textbf{Callgrind breakdown on the Query Execution step of an Insert Replay step}.}
+\label{tab:throughput_exp_insert_overhead_all}
+\end{center}
+\end{table}
+Results show that the code generation throughput is 47\% slower than the baseline implementation.
+Profiling result \cref{tab:throughput_exp_insert_overhead_all} explains that 81.7\% of the slow down comes from the execution of the compiled query.
+This is unexpected, as we imangined that most of the overhead would come from memory copies for extracting values, but those memory copy overhead only accounts for 6\%.
+
+A further dive into the composition of query execution shows that 74.4\% of the overhead comes from \texttt{StorageInterfaceInit}, which is a function that initializes a Storage Interface, as explained in Section 3.3.
 
 \subsection{Delete}
-We then measure throughput of delete operations against the baseline. The recovery manager uses only our delete implementation for recovery.
+We then measure throughput of Delete Replays against the baseline. The Recovery Manager uses only our delete replay implementation for recovery.
 \begin{table}[H]
 \begin{center}
-\begin{tabular}{ |c|c|c| } 
- \hline
-Statements Ratio & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\
- \hline
- 90\% Insert, 10\% Delete & $1.12 \times 10^4$ txns/s & $1.06 \times 10^4$ txns/s \\ 
- \hline
- 80\% Insert, 20\% Delete & $1.11 \times 10^4$ txns/s & $1.07 \times 10^4$ txns/s\\
-  \hline
- 75\% Insert, 25\% Delete & $1.13 \times 10^4$ txns/s & $1.03 \times 10^4$ txns/s\\
- \hline
+\begin{tabular}{ c c c } 
+ \toprule
+\textbf{Statements Ratio} & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\
+ \toprule
+ 90\% Insert, 10\% Delete & $1.12 \times 10^4$ txns/s & $1.08\times 10^4$ txns/s \\ 
+ \midrule
+ 75\% Insert, 25\% Delete & $1.11 \times 10^4$ txns/s & $1.07 \times 10^4$ txns/s\\
+ \midrule
+ 50\% Insert, 50\% Delete & $1.09 \times 10^4$ txns/s & $9.31 \times 10^3$ txns/s\\
+ \bottomrule
 \end{tabular}
-\caption{\textbf{Delete throughput measurement setup on a single table}. The table uses the same setup in \ref{tab:throughput_exp_insert}.}
+\caption{\textbf{Delete Replay Throughput}.}
 \label{tab:throughput_exp_delete}
@@
 
-\subsection{Update}
-Finally, we measure throughput of delete operations against the baseline. The recovery manager uses only our delete implementation for recovery.
+Code generation throughput results of Delete Replays are much closer to the baseline (90\% to 95\%), compared to those of Insert Replays.
+This is mainly because a delete operation does not require any memory copies from a redo record.
+Moreover, executing a delete operation is also a lightweight operation, as we have prevented the pipeline from performing extra work as explained in Section 3.2.2.
+
 \begin{table}[H]
 \begin{center}
-\begin{tabular}{ |c|c|c| } 
- \hline
-Statements Ratio & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\ 
- \hline
- 90\% Insert, 10\% Update & $???$ txns/s & $???$ txns/s\\
-  \hline
- 75\% Insert, 25\% Update & $???$ txns/s & $5.6 \times 10^3$ txns/s\\
- \hline
- 50\% Insert, 50\% Update & $???$ txns/s & $10^4$ txns/s \\ 
- \hline
+\begin{tabular}{ l c c l } 
+ \toprule
+\textbf{Functionality} & \textbf{IR\%} & \textbf{Normalized IR\%}\\ 
+ \toprule
+Query Execution (Op Codes) & 90.40\% & 91.46\%\\
+ \midrule
+ Execution Context Initialization & 3.70\% & 3.74\% \\ 
+ \midrule
+ Delete Record Initialization &4.22\% & 4.27\% \\
+ \midrule
+ Metadata Cache Map Access & 0.36\% & 0.36\% \\
+ \midrule
+ Compile & 0.17\% & 0.17\% \\
+ \bottomrule
 \end{tabular}
-\caption{\textbf{Update throughput measurement setup on a single table}. The table uses the same setup in \ref{tab:throughput_exp_insert}.}
+
+\caption{\textbf{Callgrind breakdown on a Delete Replay step}.}
+\label{tab:throughput_exp_delete_overhead_all}
+\end{center}
+\end{table}
+
+\begin{table}[H]
+\begin{center}
+\resizebox{\columnwidth}{!}{\begin{tabular}{ l c c l } 
+ \toprule
+\textbf{Operation Code} & \textbf{IR\%} & \textbf{Normalized IR\%} & \textbf{Description}\\ 
+ \toprule
+\texttt{StorageInterfaceInit} & 24.35 & 76.27\% & Initializes a Storage Interface.\\
+ \midrule
+ \texttt{StorageInterfaceTableDelete} & 5.80 & 18.16\% & Deletes a tuple from a table.\\ 
+ \midrule
+ \texttt{StorageInterfaceFree} & 1.55 & 4.85\% & Deallocates a Storage Interface.\\
+ \midrule
+ Other & 0.13 & 0.41\% & \\
+ \midrule
+ \texttt{StorageInterfaceGetTablePR} & 0.10 & 0.31\% & Retrieves a table's projected row.\\
+ \bottomrule
+\end{tabular}}
+\caption{\textbf{Callgrind breakdown on the Query Execution step of a Delete Replay step}.}
+\label{tab:throughput_exp_delete_overhead_all}
+\end{center}
+\end{table}
+
+Callgrind analysis shows that, similar to an insert operation, a delete operation is heavily bottlenecked by query execution and initializing the Storage Interface.
+
+\subsection{Update}
+Finally, we measure throughput of Update Replays against the baseline. The Recovery Manager uses only our update replay implementation for recovery.
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ c c c } 
+ \toprule
+\textbf{Statements Ratio} & \textbf{Baseline Throughput} & \textbf{Codegen Throughput}\\ 
+ \toprule
+ 90\% Insert, 10\% Update & $1.15 \times 10^4$ txns/s & $7.46 \times 10^3$ txns/s\\
+ \midrule
+ 75\% Insert, 25\% Update & $1.15 \times 10^4$ txns/s & $7.31 \times 10^3$ txns/s\\
+ \midrule
+ 50\% Insert, 50\% Update & $1.15 \times 10^4$ txns/s & $6.68 \times 10^3$ txns/s \\ 
+ \bottomrule
+\end{tabular}
+\caption{\textbf{Update Replay Throughput}.}
 \label{tab:throughput_exp_update}
@@
 \end{table}
+The performance of an Update Replay is similar to an Insert Replay.
+Code generated updates are more heavyweight compared to the baseline insert operations and update operations due to initialization overheads of the Storage Interface.
+Therefore, as the proportion of update statements increases, we notice a consistent drop on the code generation throughput.
 
-\section{Throughput Measurement With Indexes}
-In the baseline recovery architecture, the recovery manager implements a custom index update function that is independent from the query execution engine.
-This index update function is largely hard-coded and contains large amounts of memory copy operations.
-The code generation approach, however, use the built-in execution engine for index updates.
-We believe as the number of indexes increases, the performance gap between baseline and the code generation approach will shorten.
 \begin{table}[H]
 \begin{center}
-\begin{tabular}{ |c|c|c|c|c| } 
- \hline
-Statements Ratio & \#Indexes & \textbf{Baseline Throughput} & \textbf{Codegen Throughput} & \textbf{Baseline:Codegen} \\ 
- \hline
- 100\% Insert & 5 & $3.7 \times 10^4$ txns/s & $2.4 \times 10^4$ txns/s & 1.54\\
- \hline
- 100\% Insert & 10 & $2.4 \times 10^4$ txns/s & $1.7 \times 10^4$ txns/s & 1.41\\
- \hline
- 100\% Insert & 100 & $3.2 \times 10^3$ txns/s & $2.4 \times 10^3$ txns/s & 1.3\\
- \hline
+\begin{tabular}{ l c c } 
+ \toprule
+ \textbf{Functionality} & \textbf{IR\%} & \textbf{IR\%/sum(IR\%)} \\ 
+ \toprule
+Query Execution (Op Codes) & 80.08 & 83.50\% \\
+ \midrule
+ Execution Context Initialization & 9.41 & 9.81\% \\ 
+ \midrule
+ Retrieving Record Values & 4.98 & 5.19\% \\
+ \midrule
+ Metadata Cache Map Access & 1.27 & 1.32 \% \\
+ \midrule
+ Compile & 0.18 & 0.18 \% \\
+ \bottomrule
 \end{tabular}
-\caption{\textbf{Insert throughput measurement setup on a single table with indexes}. The table uses the same setup in \ref{tab:throughput_exp_insert}.}
-\label{tab:throughput_exp_index_insert}
+
+\caption{\textbf{Callgrind breakdown on an Update Replay step}.}
+\label{tab:throughput_exp_update_overhead_all}
 \end{center}
@@
 
-\section{Throughput Measurement With Constraints}
+\begin{table}[H]
+\begin{center}
+\resizebox{\columnwidth}{!}{\begin{tabular}{ l c c l } 
+ \toprule
+\textbf{Operation Code} & \textbf{IR\%} & \textbf{IR\%/sum(IR\%)} & \textbf{Description}\\ 
+ \toprule
+\texttt{StorageInterfaceInit} & 29.84 & 72.50\% & Initializes a Storage Interface.\\
+ \midrule
+ \texttt{StorageInterfaceTableUpdate} & 8.42 & 20.46\% & Updates a tuple of a table.\\ 
+ \midrule
+ \texttt{StorageInterfaceFree} & 2.78 & 6.75\% & Deallocates a Storage Interface.\\
+ \midrule
+ Other & 0.12 & 0.29\% & \\
+ \bottomrule
+\end{tabular}}
+\caption{\textbf{Callgrind breakdown on the Query Execution step of an Update Replay step}.}
+\label{tab:throughput_exp_update_overhead_all}
+\end{center}
+\end{table}
+
+Callgrind analysis shows that an update operation is also spending most of its time on query execution and creating the Storage Interface.
+Combined with results from Section 4.1.1 and 4.1.2, we can confirm that the allocation/deallocation Storage Interface becomes the main bottleneck of all the code generation replays.
+
+\section{Throughput Measurement With Indexes}
+For in-memory DBMS, indexes are rarely stored on disk.
+Instead, they are reconstructed during log replay.
+However, this introduces design problems for an execution engine that has been adapted to work with compiled code.
+The recovery system needs to implement functionally redundant code to replay the indexes.
+
+In the baseline recovery architecture, the Recovery Manager implements a custom index update function that is independent from the query execution engine.
+The code generation approach, however, uses the built-in execution engine for index updates.
+We believe as the number of indexes increases, the performance gap between baseline and the code generation approach will diminish.
+
+\begin{figure}[t!]
+\centering
+\begin{subfigure}{.45\textwidth}
+ \centering
+ \includegraphics[width=1.0\textwidth]{images/IndexRecoveryThroughput.png}
+ \caption{Recovery throughput over the number of indexes. Red is baseline, blue is code generation.}
+  \label{fig:index_recovery_throughput}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+ \centering
+ \includegraphics[width=1.0\textwidth]{images/IndexRecoveryRatio.png}
+ \caption{Throughput ratio between code generation and baseline implementation.}
+  \label{fig:index_recovery_ratio}
+\end{subfigure}
+\caption{\textbf{Recovery with Indexes}.}
+\label{fig:index_recovery_throughput}
+\end{figure}
+
+For this experiment, we want to focus on the comparison for index recovery and avoid as much overhead from other operations as possible.
+Therefore, we go for a much lighter insert setup.
+However, it should be noted that some of the heavy overhead is still unavoidable (e.g., value copies, Storage Interface construction/destruction).
+
+We create an empty table with a single column that accepts \dbSQL{INTEGER} data type.
+We execute $10^6$ transactions to insert random integers into the table.
+For each data point, we create an increasing number of indexes for the table.
+Results show that as the number of indexes increases, the gap between the baseline and code generation implementation steadily decreases.
+Note that without any indexes, the gap between baseline and code generation is around two times, which matches the results from \cref{tab:throughput_exp_insert}.
+
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{ l c c } 
+ \toprule
+\textbf{Functionality} & \textbf{IR\%} & \textbf{Normalized IR\%}\\ 
+ \toprule
+Index Update & 96.23 & 96.3\% \\
+ \midrule
+Insert & 3.11 & 3.11\% \\ 
+ \midrule
+Other (e.g., tuple map access) & 0.57 & 0.59\% \\
+ \bottomrule
+\end{tabular}
+\caption{\textbf{Callgrind breakdown on Baseline Index Recovery}.  Number of indexes is 10.}
+\label{tab:throughput_exp_insert_overhead_all}
+\end{center}
+\end{table}
+
+\begin{table}[H]
+\begin{center}
+\resizebox{\columnwidth}{!}{\begin{tabular}{ l c c l }
+ \toprule
+\textbf{Operation Code} & \textbf{IR\%} & \textbf{Normalized IR\%} & \textbf{Description}\\ 
+ \toprule
+\texttt{StorageInterfaceIndexInsertUnique} & 33.77 & 37.9\% & Insert into an index.\\
+ \midrule
+ \texttt{StorageInterfaceInit} & 32.22 & 36.2\% & Initializes a Storage Interface.\\ 
+ \midrule
+ \texttt{StorageInertfaceGetIndexPR} & 20.95 & 23.5\% & Retrieves an index's projected row.\\
+ \midrule
+ \texttt{StorageInterfaceTableInsert} & 1.46 & 1.63\% & Insert a tuple into a table.\\
+ \midrule
+ \texttt{StorageInterfaceTablFree} & 0.39 & 0.43\% & Deallocates a Storage Interface.\\
+ \midrule
+ \texttt{StorageInterfaceGetTablePR} & 0.25 & 0.3\% & Retrieves a table's projected row.\\
+ \midrule
+ Other & 0.02 & 0.02\% &  \\
+ \bottomrule
+\end{tabular}}
+\caption{\textbf{Callgrind breakdown on a Query Execution step of Code Generation Index Recovery}. Number of indexes is 10. The query execution step accounts for 96\% IR for a single replay step.}
+\label{tab:throughput_exp_insert_overhead_all}
+\end{center}
+\end{table}
+
+A deeper dive into the implementations shows a percentage difference in index updates between the two approaches.
+Index updates account for $96.3\%$ of total instructions in the baseline implementation.
+On the other hand, for the code generation approach, index update accounts for $37.9\% + 23.5\% = 61.4\%$ of total instructions.
+This implies that the code generation requires less overhead to prepare for an insert index operation.
+
+\section{Throughput Measurement With Scaling Columns}
+\begin{figure}[ht!]
+\centering
+\begin{subfigure}{.44\textwidth}
+ \centering
+ \includegraphics[width=1.0\textwidth]{images/ColumnRecovery.png}
+ \caption{Throughput over the number of columns. Red is baseline, blue is code generation.}
+  \label{fig:col_recovery_throughput}
+\end{subfigure}%
+\begin{subfigure}{.45\textwidth}
+ \centering
+ \includegraphics[width=1.0\textwidth]{images/ColumnRecoveryRatio.png}
+ \caption{Throughput ratio between code generation and baseline implementation.}
+  \label{fig:col_recovery_ratio}
+\end{subfigure}
+\caption{\textbf{Recovery with Scaling Number of Columns}.}
+\label{fig:col_recovery_throughput}
+\end{figure}
+
+\begin{figure}[ht!]
+\centering
+\begin{subfigure}{.44\textwidth}
+ \centering
+ \includegraphics[width=1.0\textwidth]{images/ColumnRecoveryIndex.png}
+ \caption{Throughput over the number of columns. Red is baseline, blue is code generation.}
+  \label{fig:col_recovery_throughput_index}
+\end{subfigure}%
+\begin{subfigure}{.45\textwidth}
+ \centering
+ \includegraphics[width=1.0\textwidth]{images/ColumnRecoveryIndexRatio.png}
+ \caption{Throughput ratio between code generation and baseline implementation.}
+  \label{fig:col_recovery_ratio_index}
+\end{subfigure}
+\caption{\textbf{Recovery with Scaling Number of Columns}. Number of Indexes is 10. }
+\label{fig:col_recovery_throughput_index}
+\end{figure}
 
 \chapter{Related Work}
-\chapter{Conclusion}
+Our code generation approach applies for DBMSs that use physical logging for recovery.
+It does not apply to DBMSs that use other logging schemes, such as logical logging and physiological logging.
+For instance, VoltDB\cite{malviya14} uses a form of physiological logging known as command logging.
+The DBMS records the transactions instead of physical changes in the log records.
+
+Our method also does not extend to systems that do not have logging architectures. SAP HANA\cite{lee18} uses a log-less approach through parallel table replication.
+
+\chapter{Conclusion and Future Work}
+We presented a code generation recovery approach for DBMS.
+Our method integrates the code generation Execution Engine with the recovery system by converting log records into compiled native code.
+We provided an overview of the logging and query compilation architecture of NoisePage.
+We then showed how we implemented our approach within NoisePage.
+Finally, we evaluated our approach with throughtput experiments to show that while code generation recovery has worse recovery throughput compared to NoisePage baseline recovery, it offers less engineering overhead and is more efficient in processing index updates.
+
+Our results show that the initialization process for executing compiled queries (e.g., creating Storage Interface) is the main bottleneck for code generation recovery in NoisePage.
+One possible optimization is to cache reusable objects that are repeatedly initialization during query execution inside the recovery system or the execution engine.
 

### 2021-08-04 paper.tex

@@
 
-\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={red}}
+\hypersetup{colorlinks,linkcolor={black},citecolor={blue},urlcolor={black}}
 
@@
 We ask the recovery system to convert log records into compiled code that will be easily accepted by the execution engine.
-Our results show that while our approach incurs a higher performance overhead compared to using a separate execution engine, it requires much less engineering effort and is potentially superior in aspects such as index updates and constraint checking.
+Our results show that while our approach incurs a higher performance overhead compared to using a separate execution engine, it requires much less engineering effort and is superior in index updates.
 \end{abstract}
@@
 As a result, the DBMS contains two implementations that provide identical functionalities.
-Similar reasoning applies to other functionalities (e.g., constraint checking, trigger events) that may be used during log replay.
 The number of those functionalities in the recovery system will eventually amount to the scale of a separate execution engine that is built specifically for log replay.
@@
 We then explain how the two components interact within a DBMS and what changes need to be made for them to cooperate.
-
-\section{Transactions}
-Transactions are the atomic operations of any DBMSs~\cite{theo83}.
-They appear throughout a DBMS, including the recovery system and the execution engine responsible for code generation.
-A transaction is a sequence of interactions with the database.
-These interactions can be retrieving data, inserting data, removing data, etc.
 
@@
 However, this approach completely isolates the recovery system from the execution engine.
-This means that if the recovery system wishes to use a functionality within the execution engine, it has to rewrite that same functionality within the recovery system (\cref{fig:problem}).
+This means that if the recovery system wishes to use a functionality within the execution engine, it has to rewrite that same functionality within the recovery system (\cref{fig:recovery_index_update}).
 
@@
 We have explained in Section 2.1 that the recovery system needs to perform extra operations (e.g.
-index updates, constraint checking in \cref{fig:problem}) besides updates to the data tables.
+\cref{fig:recovery_index_update}) besides updates to the data tables.
 These functionalities exist within the execution engine.
@@
 These potential benefits motivate us to design a code generation recovery system based on conversion from log records to physical plans.
-In the next chapter, we will present the implementation of our approachhbmmn .
+In the next chapter, we will present the implementation of our approach .
 
@@
 We then need to make minor modifications to the execution engine, so it will be able to correctly execute the native code generated by the recovery system.
-We further summarize our approach (\cref{fig:codegen_recovery}) in three steps:
+We further summarize our approach (\cref{fig:codegen_recovery}) in two steps:
 
 \begin{enumerate}
-\item Interpret and extract contents from log records.
-\item Convert the extracted contents into physical plans that can be compiled into native code (Code Generation).
-\item Run the compiled machine code using the execution engine.
+\item Convert the log records into physical plans that can be compiled into native code (Recovery Code Generation).
+\item Run the compiled machine code using the execution engine (Recovery Execution) for log replay.
 \end{enumerate}
@@
 NoisePage's recovery system consists of three major components: Transaction Manager, Logging Manager, and Recovery Manager.
-Both the Logging Manager and the Recovery Manager interact with the Transaction Manager to achieve logging and recovery respectively.
-We will first give an overview of NoisePage's transaction manager,
+The Logging Manager interacts with the Transaction Manager to achieve physical logging, while the Recovery Manager interacts with the Transaction Manager to provide database recovery.
 
@@
 NoisePage uses a transaction to perform physical updates on the database and physical logging to disk.
-NoisePage uses a multi-versioned delta store Transaction Manager that ensures \texttt{SNAPSHOT-ISOLATION}.
+NoisePage uses a multi-versioned delta store~\cite{reed78} Transaction Manager that ensures \texttt{SNAPSHOT-ISOLATION}~\cite{berenson95}.
 The Transaction Manager allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis.
-A transaction's lifecycle begins when the transaction begins, and ends after the log manager has serialized its changes to disk.
-
-Each data tuple in NoisePage is uniquely identified through a tuple slot that stores the offset of a tuple.
-When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table.
+A transaction's lifecycle starts when the transaction begins, and ends after the log manager has serialized its changes to disk.
+
+Each data tuple in NoisePage is uniquely identified through a tuple slot that stores the offset of a tuple in a data block.
+When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table's data block.
 Subsequent updates to the tuple does not create direct copies of the tuple, but stores delta information (redo records) about the updates instead.
@@
 
-A transaction uses a buffer (redo buffer) that stores all the delta records that will appear in its lifetime.
+A transaction uses a buffer (redo buffer) that stores all the delta records that will be generated in its lifetime.
 Each redo buffer has a fixed size and is allocated from a centralized memory pool.
-Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to dump a corresponding delta record of the change.
+Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to store a corresponding delta record of the change.
 If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the memory pool.
-The redo buffer allows the logging components to process the changes before the transaction is completed.
+The redo buffer allows the logging components to process the changes before the transaction ends.
 
@@
 The redo buffer is then carried over to the log manager, where the changes will be serialized to disk.
-When a transaction aborts, the transaction creates an abort record that will prevent corresponding records to be replayed during recovery.
-Abort records are essential; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk.
-Aborted transactions that are persisted on disk are rare, however, since it would be difficult for a typical OLTP transaction to fill up the entire redo buffer.
+When a transaction aborts, the transaction creates an abort record that prevents corresponding records to be replayed during recovery.
+Abort records are essential to recovery; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk.
+During recovery, the DBMS needs to remove aborted transactions.
+Aborted transactions that are persisted on disk are rare~\cite{tpc15}, however, since it would be difficult for an OLTP transaction to fill up the entire redo buffer.
 
@@
 
-The Log Manager (\cref{fig:log_manager}) is responsible for performing write-ahead logging in NoisePage.
-All the changes are serialized to disk as logs before they are performed on the DBMS storage.
+The Log Manager (\cref{fig:log_manager}) is responsible for performing write-ahead logging~\cite{aries} in NoisePage.
 Each log record is self-contained and does not require any additional metadata to be replayed.
@@
 The Log Serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers.
-These buffers will be consumed by the Log Consumer for persistence.
-
-The log records must be serialized in the correct order for recovery to perform correctly.
+The Log Consumer consumes these buffers for persistence.
+
+The DBMS must serialize log records in the correct order for recovery to perform correctly.
 Different transactions can share a same redo buffer that is filled with each of their own log records.
@@
 \subsubsection{Log Consumer}
-Buffers supplied by the log serializer can be consumed by different Log Consumer tasks.
-Each Log Consumer is given a copy of the original buffer to achieve parallelism.
+Different Log Consumers are responsible for consuming buffers supplied by the Log Serializer.
+The Log Serializer provides each Log Consumer a copy of the original buffer to achieve parallelism.
 NoisePage uses two type of log consumers: Disk Log Consumer and Replication Log Consumer.
@@
 \cref{fig:recovery_manager} shows the architecture of the Recovery Manager that is responsible for log replay.
+The Recovery Manager loads the log records into memory and use them to performs log replay.
 NoisePage uses four different record types: redo, delete, commit, abort.
-Redo and delete records are not processed immediately.
-Instead, they are placed inside a deferred changes buffer, ordered by their transaction timestamp.
-An abort record removes the corresponding redo/delete record from the buffered changes.
+The Recovery Manager do not process redo and delete records are immediately.
+Instead, the Recovery Manager place redo and delete records inside a deferred changes buffer, ordered by their transaction timestamp.
+An abort record removes a corresponding redo/delete record from the buffer using the begin transaction timestamp (\cref{fig:log_record}).
 A commit record initiates a transaction to iterate over the deferred changes buffer.
-Once all the deferred changes are processed, the buffer is cleared and the transaction is committed.
+Once all the Recovery Manager processed all the deferred changes, it clears the buffer and commits the transaction.
 The transaction aborts if any log record is malformed or any replay fails.
@@
 
-A physical log record consists of a header followed by a record body (\cref{fig:log_record}).
-The header stores metadata about the log record itself, including its record type, size and begin timestamp of the transaction that generated the log record as mentioned in Section 3.1.1.
+A physical log record consists of a header, followed by a record body (\cref{fig:log_record}).
+The header stores metadata about the log record itself.
+This includes a log record's record type, size, and begin timestamp of the transaction (See Section 3.1.1).
 The record body contains physical information that is necessary for recovery and varies depending on the record type.
@@
 Tuple slots in the log records will no longer be valid memory locations during recovery.
-Instead, the Recovery Manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted.
-The mapping allows the Recovery Manager to apply changes to the correct memory locations in the new memory environment.
-
-Since only redo and delete records will effectively modify physical storage, we only need to analyze the record body of those two record types.
-Modifications of a query to the DBMS can be resolved down to the modification of tuple slots on the physical level.
+Instead, the Recovery Manager creates an internal mapping from original tuple slots to new tuple slots that will be retrieved after new values are inserted.
+The tuple slot mapping allows the Recovery Manager to apply changes to the correct memory locations in the new memory environment.
 
@@
 \end{figure}
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.9\textwidth]{images/PRExample.png}
+\caption{Projected Row that represents [15, 721, \dbSQL{NULL}].}
+\label{fig:projected_row_example}
+\end{figure}
 
  While storing tuple slots is sufficient for delete records, redo records require additional data to perform log replay.
-These data are kept in the record projected row (\cref{fig:projected_row}), which stores a mapping from column ids to values.
-The DBMS retrieves individual values by computing offsets.
+For instance, to perform an insert, the redo record must contain the values to insert.
+The Log Manager stores these data in a log record's projected row (\cref{fig:projected_row}).
+A projected row represents a row in a database table.
+The DBMS can access an individual value within the projected row by identifying a column id and reading the value using that column's value offset (\cref{fig:projected_row_example}).
 
 \subsubsection{Replay Step}
-A replay step refers to the recovery manager replaying a single redo record or delete record. A NoisePage replay step has three different scenarios:
+The Recovery Manager processes each log record differently depending on its record type.
+Commit and abort records do not change the physical storage of the the database.
+Therefore, to simplify, we only need to consider the effects of redo and delete records.
+
+We refer to a replay step as the Recovery manager replaying a single redo record or delete record. A NoisePage replay step has three different scenarios:
 \begin{itemize}
@@
 
-While both insert and update redo records are represented using the same redo record structure, they can be differentiated by keeping track of all the tuple slots found in log records.
-If a tuple slot inside a redo record is unseen before, then the redo record represents an Insert operation.
+While both insert and update redo records follow the same redo record structure, the Recovery Manager differentiates between them by keeping track of all the tuple slots found in log records.
+If the Recovery Manager has never seen a tuple slot inside a redo record before, then the redo record represents an Insert operation.
 Otherwise, the Redo Record represents an Update operation.
@@
 \section{Recovery Code Generation}
-The code generation structure in NoisePage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \cref{fig:query_compilation}.
-A query is interpreted, optimized and converted into a physical plan tree.
+In this section, we introduce how we integrate the NoisePage's recovery system with its code generation pipeline.
+We first provide an overview over NoisePage's code generation pipeline.
+We then show how we convert redo and delete records into physical plans that can be accepted by the code generation pipeline for compilation.
+
+\subsection{Code Generation Pipeline}
+The DBMS interprets a query, parses it into an AST and converts it into a physical plan tree.
 A plan tree consists of plan nodes and specifies how the query should be executed on the physical level of the database.
 
-A plan translator then converts a physical plan tree/node into an operator pipeline.
+Code generation is responsible for compiling physical plans into machine code.
+The code generation pipeline in NoisePage follows from the data-centric compilation approach\cite{neumann11}, as shown in \cref{fig:query_compilation}.
+The DBMS feeds each plan node into a matching plan translator.
 Each plan translator correspond to a plan node in the original plan tree.
-A pipeline contains an ordered collection of relational operators that are not yet materialized, ended with a pipeline breaker (e.g., sort tuples, creating a hash table).
-The pipeline is only materialized whenever a pipeline breaker has been encountered.
+For instance, an insert plan node has a matching insert plan translator; a delete plan node has a matching delete plan translator.
+A plan translator converts a physical plan tree/node into an operator pipeline.
+The operator pipeline contains an ordered collection of relational operators (e.g.
+insert, get table row) that are not yet materialized, ended with a pipeline breaker (e.g., sort tuples, creating a hash table).
+The execution engine only materializes the pipeline results when a pipeline breaker has been encountered.
 
@@
 \end{figure}
-NoisePage then compiles each operator pipeline into a customized imperative language referred to as (TPL).
-The DBMS then compiles each function of a query inside the TPL code into some bytecode (TBC).
+NoisePage then compiles each operator pipeline into a customized imperative language (TPL).
+The DBMS then compiles each function of a query inside TPL into some bytecode (TBC).
 The bytecode can be interpreted by a built-in virtual machine, or further compiled into an LLVM module.
@@
 To run the query \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}, the DBMS needs to first do a sequential scan to find out the columns, then do a delete operation on the found columns.
-These two steps are manifested in the delete plan node that includes a sequential scan plan node as a child.
-The plan node then translates into an operator pipeline that specifies how each operation will be exactly executed in the DBMS.
-The operator pipeline then converts to TPL code.
-Each function inside the TPL code correspond to a piece of function inside the execution engine that will be evetually invoked once the compiled native code is run.
+The DBMS represents these two operations in a delete plan node that includes a sequential scan plan node as a child.
+The DBMS then translates the plan node into an operator pipeline that specifies how each operation will be exactly executed in the DBMS.
+After this, the DBMS compiles the operator pipeline into machine code (TPL, TBC, LLVM).
+Each function inside the machine code matches to an operation code function inside the execution engine.
+When the execution engine runs the compiled code, it invokes its corresponding operation code functions.
 For instance, \texttt{@tableDelete} corresponds to a C++ function \texttt{OpStorageInterfaceTableDelete} (a delete operation on a tuple slot of a table) in the execution engine.
 
+\subsection{Recovery Integration}
 \begin{figure}[t!]
@@
 \end{figure}
-We imagine each replay as a form of SQL statement that can be converted into native code through the code generation process discussed above (\cref{fig:expression_generation}).
-The code generation process for a log record can be divided into two steps: plan generation (creating the physical plan node) and plan translation (translating the plan into pipelines).
-Each replay (Insert Replay, Update Replay, Delete Replay) will be matched to a corresponding plan node.
-
-\subsection{Insert Replay Conversion}
+We can consider each type of replay as a form of SQL statement that the DBMS can convert into native code through the code generation process discussed above (\cref{fig:expression_generation}).
+We further divide the code generation process for a log record into three steps: (1) plan generation (creating the physical plan node), (2) plan translation (translating the plan into operator pipelines), and (3) compilation.
+
+In plan generation, the Recovery Manager converts each replay process (insert/update/delete replay) into a corresponding plan node.
+In plan translation, the Recovery Manager uses matching plan translators to convert physical plans into operator pipelines.
+Once the Recovery Manager has the operator pipeline, it can then use the compiler to generate machine code for execution.
+
+\subsubsection{Insert Replay Conversion}
+\label{insert_replay_conversion}
 
@@
 \end{figure}
-Insert plan nodes correspond to \dbSQL{INSERT} statements.
-We can represent an insert redo record as the SQL statement \dbSQL{INSERT INTO x VALUES y}.
-The insert plan node for this statement does not require any child plan nodes to function (\cref{fig:insert_conversion}).
-This makes it straightforward to convert from an insert redo record to a plan node.
-Index information and column ids can be retrieved from the catalog.
-Values stored in the redo records are in their most basic form (e.g., integer, float, double) as constants and can be converted into expressions to be accepted by the insert plan node.
-The operation of transferring values, however, require one-to-one copy of each value from the log record into the plan node.
-
-\subsection{Delete Replay Conversion}
+Insert plan nodes correspond to \dbSQL{INSERT} statements. We can represent an insert redo record as the SQL statement \dbSQL{INSERT INTO x VALUES y}.
+
+\cref{fig:insert_conversion} shows the conversion from an insert redo record into an insert plan node.
+The Recovery Manager can retrieve metadata (e.g.
+table id, column ids) from the redo record and the database catalog.
+The redo record's projected row contains values for insert (\cref{fig:projected_row}).
+Those values are in primitive forms (e.g., integer, float, double) as constants.
+The Recovery Manager then feeds those constants into the insert plan node.
+This conversion requires no changes to the plan nodes, nor to the operator pipelines.
+
+\subsubsection{Delete Replay Conversion}
+\label{delete_replay_conversion}
+
 \begin{figure}[t!]
@@
 \includegraphics[width=0.9\textwidth]{images/DeleteConversion.png}
-\caption{Conversion from a Delete Redo Record into a Delete Plan Node. Tuple slot that will be used in the pipeline is "smuggled" from the redo record through Execution Context.}
+\caption{Conversion from a Delete Redo Record into a Delete Plan Node. The operator pipeline uses the tuple slot from the delete record.}
 \label{fig:delete_conversion}
@@
 
-Delete plan nodes correspond to \dbSQL{DELETE} statements and require child nodes to function.
-A delete plan node uses the child plan node to figure out which tuple slot to schedule for deletion.
-This is different for delete records.
-The mapping from Recovery Manager combined with the original tuple slot already points to the tuple slot to delete.
-In this case, there will be no need for the plan node to go through the child node and resolve the delete tuple slot.
-This is because, eventually, the delete plan node will attempt to create a Delete Record, which is structurally identical to the Delete Record from the Recovery Manager.
-We will therefore change the behavior of the operator pipeline to retrieve the delete slot supplied by the recovery manager during recovery (\cref{fig:delete_conversion}), and remove the step to resolve which tuple slot to delete.
-Other operations, such as index deletions, are still necessary and are kept in place.
-
-\subsection{Update Replay Conversion}
-Update plan nodes correspond to \dbSQL{Update} statements and require child nodes to function.
-Unfortunately, update plan nodes are not very compatible with update redo records.
-After an update plan is translated into an update operator pipeline, it expects set clauses returned from the child node (e.g., sequential scan).
-These set clauses, however, describe higher-level logical operations and cannot be constructed as easily and efficiently as creating primitive expressions for insert plan nodes.
+Delete plan nodes correspond to \dbSQL{DELETE} statements.
+\dbSQL{DELETE} statements require specifications on where the DBMS should perform delete operations (e.g., \dbSQL{WHERE} clause in \cref{fig:codegen_example}).
+Therefore, a delete plan node requires a child plan node to function.
+It uses its child plan node to figure out which tuple slot to schedule for deletion.
+
+However, a delete record's tuple slot, combined with the tuple mapping from Recovery Manager, already points to the tuple slot to delete.
+In this case, there will be no need for the plan node to go through the child node and find out the delete tuple slot.
+Moreover, delete records do not contain value information that is required for index update (\cref{fig:log_record}).
+Therefore, we need to provide the operator pipeline with a projected row that contains values that correspond to the row of the tuple slot from the Recovery Manager.
+
+We change the behavior of the operator pipeline as follows (\cref{fig:delete_conversion}): (1) remove the step to look for a new delete tuple slot, (2) retrieve the delete slot supplied by the Recovery Manager during execution, and (3) use the projected row provided by the Recovery Manager for index updates.
+
+\subsubsection{Update Replay Conversion}
+\label{update_replay_conversion}
+
+Update plan nodes correspond to \dbSQL{UPDATE} statements.
+Similar to \dbSQL{DELETE} statements, update plan nodes require child plan nodes.
+An update operator pipeline expects \dbSQL{SET} clauses returned from the child node (e.g., sequential scan).
+These \dbSQL{SET} clauses, however, describe higher-level logical operations (i.e., update a column with a certain value).
+On the other hand, redo records contain low-level physical information (i.e., update a tuple slot with a certain value).
+Therefore, the Recovery Manager cannot create \dbSQL{SET} clauses efficiently from redo records.
+
 \begin{figure}[t!]
@@
 \includegraphics[width=0.9\textwidth]{images/UpdateConversion.png}
-\caption{Conversion from an Update Redo Record into an Update Plan Node. Set clauses and child plan nodes inside the update plan node are discarded.}
+\caption{Conversion from an Update Redo Record into an Update Plan Node. \dbSQL{SET} clauses and child plan nodes inside the update plan node are discarded.}
 \label{fig:update_conversion}
 \end{figure}
-The update plan node therefore needs to be extended to accept primitive values like an insert plan node.
-We also modify the operator pipeline to process the primitive values instead of set clauses during a replay step (\cref{fig:update_conversion}).
-Functions inside the operator pipeline will now apply to the primitive value expressions.
-Similar to delete record transformations, there is no need for the update pipeline to resolve the update tuple slot or to create a new update record, as these information already exist in an update redo record and the tuple slot mapping in the Recovery Manager.
+
+We therefore perform the following changes (\cref{fig:update_conversion}): (1) extend the update plan node to accept primitive values like an insert plan node does, (2) modify the operator pipeline to process the primitive values instead of \dbSQL{SET} clauses during a replay step (\cref{fig:update_conversion}), (3) remove the step for update operator pipeline to find out the update tuple slot, and (4) retrieve the update tuple slot from the recovery manager.
 
 \section{Recovery Execution}
-The execution engine is reponsible for running the compiled query.
-After we have finished the setup for compiling different replays into native code, we will explain how they will affect the execution engine.
-
-\subsection{Storage Interface}
-The compiled query invokes different operation codes (e.g., \texttt{OpStorageInterfaceTableDelete} in \cref{fig:codegen_example}) to complete the corresponding operator in the TPL code.
-Executing the operation codes requires access to the storage layer of the DBMS.
-In NoisePage, the storage layer is separate from the execution engine.
-Therefore, it uses a Storage Interface to interact with contents in the storage layer.
-The Storage Interface stores objects that allow manipulation over a certain DBMS table.
-These objects include an Execution Context, a redo record that will be used to persist changes done to the table, projected rows associated with the redo record and index objects associated with table.
-A Storage Interface is reconstructed every time an execution of a query begins, and is destructed after the query has been executed.
-We will need to modify functions that rely on the Storage Interface and access members that reside within the Storage Interface.
+We explained in \cref{delete_replay_conversion} that we modified deletes/update operator pipelines to retrieve certain objects (e.g., tuple slots, projected row) from the Recovery Manager.
+To achieve this, we need to modify the execution engine.
+The execution engine is responsible for running the compiled query.
+Inside the execution engine, the compiled query invokes different operation codes (e.g., \texttt{OpStorageInterfaceTableDelete} in \cref{fig:codegen_example}) that correspond to operators in the operator pipeline.
+Since we modified the operator pipelines earlier, we now need to change operation codes to read in objects from the Recovery Manager.
 
@@
 \includegraphics[width=0.9\textwidth]{images/RecoveryExecution.png}
-\caption{Information from log record can be retrieved from the Execution Context.}
+\caption{Operation code functions can access log records through an Execution Context.}
 \label{fig:recovery_execution}
@@
 
-We explained in Section 3.2.2 and 3.2.3 that for our purposes, we expect deletes/updates to retrieve relevant tuple slots from the Recovery Manager.
-However, the built-in pipeline assumes that the tuple slots are retrieved later during execution.
-We can circumvent this rule by modifying operation code functions and the execution context.
-The operation code functions provide access to the Storage Interface that is required to access the execution context.
-
-An Execution Context encapsulates information that is supplied by upper layers of the DBMS.
-These information include access to the database catalog, the transaction running the query, etc.
-We can also add tuple slot information into the Execution Context.
-Since the Execution Context can be accessed throughout a Storage Interface, we can modify the behavior of operation codes to override with the recovery tuple slots, as shown in \cref{fig:recovery_execution}.
-
-For Delete and Update Replays, we need to supply the Execution Context with the log record itself.
-Since we modified the delete translator to skip the process of looking through child plan node for objects to delete, we will deliver the Delete Record from the Recovery Manager directly to the operation codes that will be expecting a Delete Record generated by a regular delete plan node.
-The same process applies to Update Replays.
-
+We first use the Recovery Manager to set the required tuple slots and projected row into a log record.
+Before the execution engine runs the query, the Recovery Manager needs to setup a preparation context for execution.
+In NoisePage, an execution context encapsulates information that is supplied by upper layers of the DBMS (e.g.
+access to the database catalog, the transaction running the query to execute, database memory pool).
+We pass the log record from the Recovery Manager into the execution context.
+
+During recovery, for each operation code function that needs to use predetermined tuple slots/projected row for delete/update replays, we enforce it to access those values from the execution context (\cref{fig:recovery_execution}).
+During the execution of operation codes, NoisePage uses a storage interface object to manipulate over a certain DBMS table.
+The execution engine initalizes the storage interface object with the execution context.
+Therefore, the storage interface provides access for the operation codes to access information in the log record from the Recovery Manager.
 \section{Caching Optimization}
@@
 \subsection{Metadata}
-Most of the metadata overhead is generated during the conversion step from a redo record into an insert plan node.
+Most of the metadata overhead is generated during the conversion step from a log record into a plan node.
 For each table, the DBMS needs to figure out its relevant metadata.
 These operations require frequent lookup into the database catalog.
-As long as those metadata are bound to some table id, the DBMS can cache those metadata and re-use them for the same table.
-Some significant metadata are listed below:
+However, all of those metadata are bound to a fixed database and table id.
+Therefore, the DBMS can cache those metadata with a hash table and re-use them for the same table.
+Below is a list of important metadata that we have cached successfully:
 
@@
 The Recovery Manager needs to maintain a mapping from physical column ids to logical column ids for each table.
-This is required because the column ids recorded in the log records are physical ids, while all the query compilation components expect logical column ids.
-The mappings can be generated by looking into the DBMS catalog.
+This is required because the column ids recorded in the log records are physical ids, while query compilation expects logical column ids.
+The Recovery Manager generates this mapping by accessing the database catalog.
 \subsubsection{Column Value Type Mapping}
 For redo records, the Recovery Manager also needs to maintain a mapping from each column to its corresponding SQL value type (e.g., \dbSQL{INTEGER, FLOAT, VARLEN}).
-This mapping is used to correctly copy the values out of each redo record into a physical plan node.
+The Recovery Manager uses this mapping to correctly copy the values out of each redo record into an insert/update plan node.
 \subsubsection{Catalog Access}
@@
 Profiling results show that a single operation to retrieve catalog access can account for 10\% of total instructions of a replay step.
-Therefore, a mapping is used to keep track of catalog accessors.
+Therefore, the Recovery Manager uses a hash table to keep track of catalog accessors.
 
 \subsection{Compiled Query}
-A major benefit of query compilation is that the compiled query can be reused by the execution engine.
+A major benefit of query compilation is that the execution engine can reuse the compiled query.
 If the contents of a physical plan node remains largely unchanged, then there is no need to reconstruct the physical plan every time the Recovery Manager processes a log record.
+For instance, we only need to create a new delete plan node if we receive a delete record with a new database and table id.
 
@@
 
-Caching is trickier for replays that need to transfer values into plan nodes.
-The recovery manager will need to reconstruct a primitive expression every time its underlying value changes.
-A solution to this is to parameterize the primitive expression given to an insert/update plan node.
-During run-time, the Execution Context pass in actual values into the parameterized expressions.
-This approach allows a compiled query to keep value type information for each table column and remain flexible enough to swap in actual values without recompilation.
+Caching is trickier for replay types that need to transfer values into plan nodes.
+We mentioned in \cref{insert_replay_conversion} that insert/update needs to copy primitive values from the redo records into plan nodes.
+In most DBMSs, however, the plan nodes expect primitive values in some wrapper objects~\cite{volcano, kersten18} referred to as expressions.
+In NoisePage, the corresponding expression for primitive values is a constant value expression.
+As a result, the Recovery Manager needs to reconstruct a constant value expression for each new value.
+
+A solution to this is to parameterize the constant value expression given to an insert/update plan node.
+We can represent an insert replay as \dbSQL{INSERT INTO x VALUES @params}.
+The Recovery Manager caches the plan nodes that contains parameter expressions.
+The parameters expressions only contain value type information, but do not contain any actual values.
+During run-time, the execution context passes in primitive values from the Recovery Manager into the plan node.
 
@@
 \subsection{Insert}
+\label{insert_test}
+
 \begin{table}[H]
@@
 \caption{\textbf{Callgrind breakdown on the Query Execution step of an Insert Replay step}.}
-\label{tab:throughput_exp_insert_overhead_all}
+\label{tab:throughput_exp_insert_exec_overhead_all}
 \end{center}
@@
 
-A further dive into the composition of query execution shows that 74.4\% of the overhead comes from \texttt{StorageInterfaceInit}, which is a function that initializes a Storage Interface, as explained in Section 3.3.
+A further dive into the composition of query execution shows that 74.4\% of the overhead comes from \texttt{StorageInterfaceInit} (\cref{tab:throughput_exp_insert_exec_overhead_all}), which is a function that initializes a storage interface, as explained in Section 3.3.
+Therefore, the major of overhead of code generation recovery is coming from execution initialization.
 
@@
 \caption{\textbf{Callgrind breakdown on the Query Execution step of a Delete Replay step}.}
-\label{tab:throughput_exp_delete_overhead_all}
+\label{tab:throughput_exp_delete_exec_overhead_all}
 \end{center}
@@
 
-Callgrind analysis shows that, similar to an insert operation, a delete operation is heavily bottlenecked by query execution and initializing the Storage Interface.
+Callgrind results show that (\cref{tab:throughput_exp_delete_overhead_all}, \cref{tab:throughput_exp_delete_exec_overhead_all}), similar to an insert operation, a delete operation is heavily bottlenecked by query execution and initializing the Storage Interface.
 
@@
 \caption{\textbf{Callgrind breakdown on the Query Execution step of an Update Replay step}.}
-\label{tab:throughput_exp_update_overhead_all}
+\label{tab:throughput_exp_update_exec_overhead_all}
 \end{center}
@@
 
-Callgrind analysis shows that an update operation is also spending most of its time on query execution and creating the Storage Interface.
-Combined with results from Section 4.1.1 and 4.1.2, we can confirm that the allocation/deallocation Storage Interface becomes the main bottleneck of all the code generation replays.
-
-\section{Throughput Measurement With Indexes}
-For in-memory DBMS, indexes are rarely stored on disk.
-Instead, they are reconstructed during log replay.
+Callgrind results (\cref{tab:throughput_exp_update_overhead_all}, \cref{tab:throughput_exp_update_exec_overhead_all}) show that an update operation is also spending most of its time on query execution and creating the Storage Interface.
+Combined with results from Section 4.1.1 and 4.1.2, we can confirm that the execution initialization is the main bottleneck of all the code generation recovery.
+
+\section{Index Recovery Throughput Measurement}
+\label{insert_index_test}
+
+Most in-memory DBMS do not log indexes to disk~\cite{hekaton2013, noisepage, malviya14, silo_r}.
+Instead, DBMSs reconstruct indexes during log replay.
 However, this introduces design problems for an execution engine that has been adapted to work with compiled code.
@@
 Therefore, we go for a much lighter insert setup.
-However, it should be noted that some of the heavy overhead is still unavoidable (e.g., value copies, Storage Interface construction/destruction).
+However, it should be noted that some of the heavy overhead is still unavoidable (e.g., insert value copies, Storage Interface construction/destruction).
 
@@
 \caption{\textbf{Callgrind breakdown on Baseline Index Recovery}.  Number of indexes is 10.}
-\label{tab:throughput_exp_insert_overhead_all}
+\label{tab:throughput_exp_index_overhead_all}
 \end{center}
@@
 \caption{\textbf{Callgrind breakdown on a Query Execution step of Code Generation Index Recovery}. Number of indexes is 10. The query execution step accounts for 96\% IR for a single replay step.}
-\label{tab:throughput_exp_insert_overhead_all}
+\label{tab:throughput_exp_index_exec_overhead_all}
 \end{center}
@@
 A deeper dive into the implementations shows a percentage difference in index updates between the two approaches.
-Index updates account for $96.3\%$ of total instructions in the baseline implementation.
-On the other hand, for the code generation approach, index update accounts for $37.9\% + 23.5\% = 61.4\%$ of total instructions.
+Index updates account for $96.3\%$ of total instructions in the baseline implementation (\cref{tab:throughput_exp_index_overhead_all}).
+On the other hand, for the code generation approach, index update accounts for $37.9\% + 23.5\% = 61.4\%$ of total instructions (\cref{tab:throughput_exp_index_exec_overhead_all}).
 This implies that the code generation requires less overhead to prepare for an insert index operation.
-
-\section{Throughput Measurement With Scaling Columns}
+Therefore, if the DBMS can find a way to mitigate execution initialization overhead, code generation recovery can potentially outperform baseline recovery when the table has high number of indexes.
+
+\section{Throughput Measurement With Scaling Number of Columns}
+Our test results with indexes have shown the potential strengths of code generation recovery.
+Besides indexes, we believe code generation recovery will also be more efficient compared to baseline recovery on a table with a higher number columns.
+The execution engine has access to more efficient function implementations compared to the Recovery Manager.
+If we scale the table with higher number of columns, this can potentially reduce the performance gap between the two implementations.
+To verify our assumption, we performed two tests: (1) An insert recovery test with the same setup as in \cref{insert_test}, and (2) an index recovery test with the same setup in \cref{insert_index_test}.
+
 \begin{figure}[ht!]
@@
 \caption{\textbf{Recovery with Scaling Number of Columns}.}
-\label{fig:col_recovery_throughput}
-\end{figure}
+\label{fig:col_insert_recovery_throughput}
+\end{figure}
+
+Results from the insert test (\cref{fig:col_insert_recovery_throughput}) show that code generation recovery failed to shorten the throughput performance gap.
+Callgrind analysis shows that as the number of columns increase, the execution initialization overhead of code generation also increases.
+Moreover, a higher number of columns also results in more memory copies for the Recovery Manager to convert an insert redo record int an insert plan node.
 
@@
 
+The index recovery test showed that for index recovery, a higher number of columns have a positive effect in decreasing the gap between code generation recovery and baseline recovery (\cref{fig:col_recovery_throughput_index}).
+Index recovery has lower execution initialization overhead compared to insert recovery (See \cref{tab:throughput_exp_index_exec_overhead_all}, \cref{tab:throughput_exp_insert_exec_overhead_all}).
+
 \chapter{Related Work}
-Our code generation approach applies for DBMSs that use physical logging for recovery.
-It does not apply to DBMSs that use other logging schemes, such as logical logging and physiological logging.
+We have not yet located papers that discuss database recovery with code generation.
+Therefore, in this section, we discuss systems that use implementations that differ from our method assumptions.
+We then show whether and how our code generation approach applies to them.
+
+To begin with, our method does not extend to systems that do not have logging architectures.
+Some in-memory databases follows the K+1 replication protocol to ensure database durability.
+SAP HANA\cite{lee18} uses a log-less approach through parallel table replication.
+
+Our code generation approach focuses on DBMSs that use physical logging for recovery.
+However, there are many DBMSs that use other logging schemes, such as logical logging and physiological logging.
 For instance, VoltDB\cite{malviya14} uses a form of physiological logging known as command logging.
 The DBMS records the transactions instead of physical changes in the log records.
-
-Our method also does not extend to systems that do not have logging architectures. SAP HANA\cite{lee18} uses a log-less approach through parallel table replication.
 
@@
 
-Our results show that the initialization process for executing compiled queries (e.g., creating Storage Interface) is the main bottleneck for code generation recovery in NoisePage.
-One possible optimization is to cache reusable objects that are repeatedly initialization during query execution inside the recovery system or the execution engine.
+Our results show that the initialization process for executing compiled queries (i.e., creating storage interface) is the main bottleneck for code generation recovery in NoisePage.
+One possible optimization is to cache reusable objects that are repeatedly initialized during query execution.
 

### 2021-08-05 paper.tex
Updates from Overleaf

@@
 % Provides a draft mark at the top of the document. 
-\draftstamp{\today}{DRAFT}
+% \draftstamp{\today}{DRAFT}
 
@@
 \Year{2021}
-\trnumber{}
+\trnumber{CMU-CS-21-131}
 
@@
 \begin{acknowledgments}
-yes
+I would like to thank my professor Andy Pavlo for guiding me through my journey at the CMU Database Group.
+His passion, energy and commitment motivated me to pursue research in database systems in the first place.
+I will carry on with my life with the valuable lessons that I learned from him.
+I would also like to thank my peers in the CMU Database Group for working with me and carrying me through my research work.
+Finally, I want to thank my family for their unwavering support.
+\newline
+
+I wish you all the best.
 \end{acknowledgments}
@@
 \section{Caching Optimization}
+\label{caching_optimization}
 
@@
 \chapter{Related Work}
-We have not yet located papers that discuss database recovery with code generation.
-Therefore, in this section, we discuss systems that use implementations that differ from our method assumptions.
-We then show whether and how our code generation approach applies to them.
-
-To begin with, our method does not extend to systems that do not have logging architectures.
-Some in-memory databases follows the K+1 replication protocol to ensure database durability.
-SAP HANA\cite{lee18} uses a log-less approach through parallel table replication.
+In this section, we discuss DBMSs with recovery systems that differ from our method assumptions. We then show whether and how our code generation approach applies to them.
+\newline
+
+To begin with, our method does not extend to DBMSs that do not have logging architectures.
+We cannot apply our log record conversion logic to those systems.
+For instance, Scuba~\cite{scuba14} proposed a recovery system that uses shared memory to persist data between processes.
+On database shutdown, Scuba copies data directly from memory heap to shared memory.
+On restart, Scuba retrieves stored DBMS memory from shared memory.
+This recovery system does not depend on log records, so we cannot apply our approach to such systems.
+\newline
 
 Our code generation approach focuses on DBMSs that use physical logging for recovery.
-However, there are many DBMSs that use other logging schemes, such as logical logging and physiological logging.
-For instance, VoltDB\cite{malviya14} uses a form of physiological logging known as command logging.
-The DBMS records the transactions instead of physical changes in the log records.
+However, there are DBMSs that use other logging schemes, such as logical logging and physiological logging~\cite{gray92}.
+For instance, VoltDB~\cite{malviya14} uses a form of physiological logging known as command logging.
+The DBMS records transactions instead of physical table changes in the log records.
+Nevertheless, those log records still correspond to insert, delete, or update operations on the DBMS table.
+Therefore, we can still apply our log record conversion process to the VoltDB's recovery framework.
+Hyper~\cite{kemper11} uses logical redo records that keeps the parameters of the stored procedures (i.e.
+a custom high-level scripting language) that represent the transactions.
+Hyper uses the stored procedures to provide functionalities such as index, delete, and update.
+Therefore, we can apply our approach to Hyper by restoring stored procedures using parameters from the log records.
+We then convert the procedures into respective plan nodes.
+In fact, converting from logical log records into physical plans might be easier than that for physical log records.
+According to \cref{insert_replay_conversion}, physical plans expect logical values and data structures (e.g.
+column ids, \dbSQL{SET} clauses) that can potentially exist directly in logical log records.
+\newline
+
+Many in-memory DBMSs also implement parallel recovery to reduce recovery time~\cite{faeber17}.
+Hekaton~\cite{hekaton2013} uses a thread per core architecture to handle parallel insertion of the data from log records.
+SiloR~\cite{silo_r} parallelizes recovery by using an epoch system and value logging that allow the recovery system to process log records in arbitrary order.
+During replay, the recovery system uses multiple log processors to replay log records from newest to oldest.
+Our code generation recovery approach might introduce extra overhead for parallelization.
+The recovery system needs to access multiple hash tables (\cref{caching_optimization}) that are necessary for executing compiled queries.
+This means that with multiple recovery threads, they need to read and modify those hash tables concurrently.
+All of the hash tables use keys based on database and table ids.
+Unless the recovery threads manage to operate on separate tables, the recovery system needs to install extra latches on the hash tables.
+As as result, this incurs performance bottlenecks for parallel recovery.
 
 \chapter{Conclusion and Future Work}
-We presented a code generation recovery approach for DBMS.
-Our method integrates the code generation Execution Engine with the recovery system by converting log records into compiled native code.
+We presented a code generation recovery approach for in-memory DBMSs.
+Our method integrates the code generation execution engine with the recovery system by converting log records into physical plans and compiling them into native code.
 We provided an overview of the logging and query compilation architecture of NoisePage.
 We then showed how we implemented our approach within NoisePage.
-Finally, we evaluated our approach with throughtput experiments to show that while code generation recovery has worse recovery throughput compared to NoisePage baseline recovery, it offers less engineering overhead and is more efficient in processing index updates.
-
-Our results show that the initialization process for executing compiled queries (i.e., creating storage interface) is the main bottleneck for code generation recovery in NoisePage.
-One possible optimization is to cache reusable objects that are repeatedly initialized during query execution.
+Finally, we evaluated our approach with throughput experiments.
+The results show that while code generation recovery has worse recovery throughput compared to NoisePage's baseline recovery, it offers less engineering overhead and is more efficient in processing index updates.
+
+We confirmed that the execution initialization for compiled queries is the main bottleneck for code generation recovery in NoisePage.
+A possible solution to this is to cache reusable objects that are repeatedly initialized during query execution.
+We will explore on more solutions to this problem and continue to improve our code generation recovery based on our current implementation in NoisePage.
 
